### 2024-06-25 annalisa_tex/main.tex
start on paper

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath, geometry, url, amsfonts}
%\geometry{margin=2.5cm}
%\title{High Dimensional Data Visualisation: Ellipsoid projections}
\title{Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution.}
\author{}

\renewcommand{\v}[1]{\boldsymbol{#1}}

\begin{document}

\maketitle

\section{Introduction}
\begin{itemize}
    \item Projection pursuit
    \item Elliptical distributions
    \item Explanations of application areas
\end{itemize}

%Representing data collected with multiple parameters, with applications to physics.

\section{Projecting a pD ellipse into 2D}

Notation:

\begin{itemize}
\item $A_{p\times p}$ is the variance-covariance representing the normal population. 
\item $\mu_{p\times 1}$ vector of means of the normal population
\item $P_{p\times 2}$ is the orthonormal basis on which the data is projected.
\item $X_{n\times p}$ data matrix, let $x$ represent a row, or abstractly a random $p$=dimensional vector. 
\item Equation of $p$D ellipse is $(\v{x}-\mu)^\top A^{-1} (\v{x}-\mu) = c^2$.
\item Equation of projected ellipse is $(\v{x}-\mu)^\top (P^\top AP)^{-1}(\v{x}-\mu) = c^2$. 
\item $BB^\top = (P^\top AP)^{-1}$, so $B=(P^\top AP)^{-1/2}$
\item The projected data is $\v{y}=P\v{x}$
\item $B^{-1}\v{y}$ are points that lie of the 2D ellipse
\item $c^2$ is a quantile from a $\chi^2_p$
\end{itemize}

% Potential reference: https://doi.org/10.1214/18-AOS1699

\section{Projection matrix onto a plane spanned by two vectors} \label{sec:orthonormal}

For orthonormal $p$-dimensional vectors $\v{v_1}$ and $\v{v_2}$, and a $p$-dimensional vector $\v{x}$, the projection of $\v{x}$ onto the plane spanned by $\v{v_1}$ and $\v{v_2}$ is given by:
\begin{equation}
    (\v{x}\cdot\v{v_1})\v{v_1}+(\v{x}\cdot\v{v_2})\v{v_2}
\end{equation}

Now, if we consider $\v{v_1}$ and $\v{v_2}$ to be our new basis vectors, we can write the projection with two coordinates as
\begin{align}
\begin{bmatrix}
    \text{---} & \v{v_1} & \text{---}  \\
    \text{---} & \v{v_2} & \text{---}
\end{bmatrix}
\v{x} = P \v{x}
\end{align}
where $P$ is the $2 \times p$ projection matrix onto the plane.

If we are given two vectors $\v{u}$ and $\v{v}$ that are not orthonormal, we can still construct an orthonormal projection matrix onto the plane they span. Let $\v{m} = (\v{\hat{u}} + \v{\hat{v}})/|\v{\hat{u}} + \v{\hat{v}}|$ be the unit vector bisecting the angle between $\v{u}$ and $\v{v}$. Then $\v{p} = (\v{\hat{u}} - \v{\hat{v}})/|\v{\hat{u}} - \v{\hat{v}}|$ is perpendicular to $\v{m}$. Setting

\begin{equation}
P := \frac{1}{\sqrt{2}}
\begin{bmatrix}
    \text{---} & \v{m} + \v{p} & \text{---}  \\
    \text{---} & \v{m} - \v{p} & \text{---}
\end{bmatrix},
\end{equation}
we have an orthonormalised projection matrix spanning the same plane as $\v{u}$ and $\v{v}$.

\section{Finding the projection of a hyperellipsoid onto a plane}

Consider an ellipsoid of $p$-dimensional points $\v{x}$ satisfying $\v{x}^T A \v{x} = c^2$, where $A$ is a positive definite $p \times p$ matrix and $c^2$ is a constant dependent on the confidence interval. If the ellipsoid is constructed from a covariance matrix, $A$ is the \emph{inverse} of the given covariance matrix. Let $L$ be the plane spanned by orthonormal vectors $\v{u}$ and $\v{v}$. We aim to project this ellipsoid onto $L$. Let
\begin{align}
P:=
\begin{bmatrix}
    \text{---} & \v{u} & \text{---}  \\
    \text{---} & \v{v} & \text{---}
\end{bmatrix}
\end{align}
be the $2 \times p$ projection matrix onto $L$.

Let $f(\v{x}) = \v{x}^T A \v{x}$. We want to find $\v{x}$ for which $\nabla f (\v{x})$ is parallel to $L$. These points, when projected onto $L$, will form the boundary of the projected ellipse.

Since $\nabla f(\v{x}) = 2A\v{x}$, we must have $A\v{x}$ parallel to $L$. That is, there exists some $2 \times 1$ vector $\v{s} = (s_1, s_2)$ such that $A \v{x} = s_1\v{u} + s_2\v{v} = P^T \v{s}$. We can set $\v{x} = A^{-1}P^T\v{s}$. For $\v{x}$ to be a point on the ellipsoid, we require that $c^2 = \v{x}^T A \v{x}$; substituting out $\v{x}$ gives
\begin{align}
    c^2 =& \v{x}^T A \v{x}\\
    =& \v{s}^T P (A^{-1})^T A A^{-1} P^T \v{s}\\
    =& \v{s}^T P A^{-1} P^T \v{s}
\end{align}
where $(A^{-1})^T = A^{-1}$ by the symmetry of $A$. We see from this equation that $\v{s}$ lies on an ellipse with $2 \times 2$ matrix $P A^{-1} P^T$.

Now, let $\bar{\v{x}} := P \v{x}$ be the projection of $\v{x}$ onto $L$. For $\bar{\v{x}}$ on the boundary of the projected ellipse, we have $\bar{\v{x}} = P \v{x} = P A^{-1} P^T \v{s}$. Then, we can isolate $\v{s} = (P A^{-1} P^T)^{-1} \bar{\v{x}}$, and, substituting out $\v{s}$, find
\begin{align}
    c^2 =& \v{s}^T P A^{-1} P^T \v{s}\\
    =& \bar{\v{x}}^T ((P A^{-1} P^T)^{-1})^T P A^{-1} P^T (P A^{-1} P^T)^{-1} \bar{\v{x}}\\
    =& \bar{\v{x}}^T (P A^{-1} P^T)^{-1} \bar{\v{x}}.
\end{align}
Again, $((P A^{-1} P^T)^{-1})^T = (P A^{-1} P^T)^{-1}$ by symmetry of $A$.

Therefore, the projection of our $p$-dimensional ellipsoid onto $L$ is the ellipse with equation $\bar{\v{x}}^T (P A^{-1} P^T)^{-1} \bar{\v{x}} = c^2$.

In order to plot this ellipse, it can be useful to construct a sequence of points on its boundary. As $P A^{-1}P^T$ is positive definite, it can be deconstructed as $P A^{-1}P^T = B B^T$. Then our ellipse equation becomes
\begin{align*}
    c^2 &= \bar{\v{x}}^T (BB^T)^{-1} \bar{\v{x}}\\
    &= (B^{-1}\bar{\v{x}})^T (B^{-1}\bar{\v{x}});
\end{align*}
that is, the points $B^{-1}\bar{\v{x}}$ lie on the circle with radius $c$ centred at the origin. As that circle is parameterised by $(c\cos(t), c\sin(t)),\; t \in [0, 2\pi)$, we can parameterise $\bar{\v{x}}$ as
\begin{equation}
    \bar{\v{x}} = cB
    \begin{bmatrix}
        \cos(t)\\
        \sin(t)
    \end{bmatrix} + P\v{\mu}_E, \; t \in [0, 2\pi)
\end{equation}
where $\v{\mu}_E$ is the $p$-dimensional center of the hyperellipsoid. Plotting 50--100 such points gives a good resolution for the shape without taking too long to render.

The constant $c$ depends on the confidence interval that the ellipse represents. In the application, users can choose to show projections of the hyperellipsoids corresponding to 1 s.d., 2 s.d., ..., 6 s.d. Let $\chi(t, p)$ denote the percentage point function of the chi-square distribution, with probability $t$ and $p$ degrees of freedom. Let $\Phi(z)$ be the cumulative distribution function of the standard, univariate normal distribution. For $p$-dimensional data, the ellipsoid corresponding to $z$ standard deviations has
\begin{equation} \label{ztoc}
c^2 = \chi(2\Phi(z) -1, p).
\end{equation}
In $p$-dimensional space, we should then expect to have 68\% of the data within 1 s.d., 95\% of the data within 2 s.d., etc., matching a univariate normal distribution. After projection, however, more points may end up inside the ellipse. The only guarantee of the projected ellipse is if a point is \emph{outside} an ellipse in the 2-dimensional projection, then the point is outside that ellipsoid in $p$-dimensional space. The converse is generally not true, but this is still an improvement on a slice ellipse, which cannot make either guarantee.

\section{Derivation using tour notation}
\subsection{Notation}

The main difference is that we consider row-vectors such $x$ or $mu$ are considered rows of an $(n\times p)$ data matrix, i.e. they are $(1\times p)$ which is needed to work directly with projections as defined in the tour.

\begin{itemize}
\item $A_{p\times d}$ is the orthonormal basis defining the projection plane, for us $d=2$
\item $\mu_{1\times p}$ vector of means of the normal population
\item $\Sigma_{p\times p}$ is the variance-covariance matrix for the normal population
\item $X_{n\times p}$ data matrix, let $x$ represent a row, or abstractly a random $p$=dimensional vector, but note that in our notation $x$ is $(1\times p)$ 
\end{itemize}

Our starting point is the $p$-dimensional ellipsoid defined by the variance-covariance matrix. For a selected size of the ellipsoid this is defined via the precision matrix, for points $x$ on the surface we have

\begin{equation}
    (x-\mu)\Sigma^{-1}(x-\mu)^T = c^2
\end{equation}

where $c$ controls the size of the ellipsoid, in the special case of a sphere $c$ would be the radius of the sphere. In our case of an ellipsoid defined via the variance-covariance matrix $c$ measures the size with scale determined by the Malahanobis distance.

\subsection{Projection of the ellipse}

The projection of the ellipsoid onto the $d=2$-dimensional plane defined via $A$ is defining an ellipse. To find the outline of this ellipse consider that the projection should capture where the tangent hyperplane on the ellipsoid surface is normal on the projection plane. This is the point along the surface curvature where the direction is changing. This is equivalent to requiring that the gradient of the ellipsoid equation should be parallel to the projection plane. For simplicity here we assume $\mu=0$.

We therefore are interested in the gradient of the function
\begin{equation}
    f(x) = x \Sigma^{-1} x^T
\end{equation}

which is

\begin{equation}
    \nabla f(x) = 2 x \Sigma^{-1}
\end{equation}

and since we require that this is parallel to the plane defined via $A$ it means that we can express the gradient in this basis:

\begin{equation}
    x \Sigma^{-1} = s A^T
\end{equation}

therefore we have

\begin{equation}
    x = s A^T \Sigma
\end{equation}

This is defining the second condition for $x$ to define a point on the ellipse. The first condition was the requirement to be on the ellipsoid surface. Putting both conditions together by replacing $x$ we get

\begin{equation}
    s A^T \Sigma \Sigma^{-1} \Sigma^T A s^T = c^2
\end{equation}

and using $\Sigma^T = \Sigma$ this gives

\begin{equation}
    s A^T \Sigma A s^T = c^2
        \label{eq:ells}

\end{equation}

Since $s$ should be defined via a projection $y$ of our $p$-dimensional $x$ we rewrite it in those terms. From

\begin{equation}
    y = x A = s A^T \Sigma A
\end{equation}

we obtain

\begin{equation}
    s = y (A^T \Sigma A)^{-1}
\end{equation}

and substitute this into Eq.~\ref{eq:ells} to find

\begin{equation}
    y (A^T \Sigma A)^{-1} A^T \Sigma A ((A^T \Sigma A)^{-1})^T y^T = c^2
\end{equation}

again we can use the symmetry of $A$ when simplifying the equation, and we arrive at our final equation for the ellipse outline

\begin{equation}
    y (A^T \Sigma A)^{-1} y^T = c^2
    \label{eq:elly}
\end{equation}

\subsection{Drawing the ellipse}

XXX not done, need to check this still

To draw the projected ellipse we use the decomposition of a real-valued positive definite matrix as

\begin{equation}
    M = (A^T \Sigma A)^{-1} = B B^T
\end{equation}

where we can define $B$ as the square-root of $M$. This can be computed via the eigendecomposition of $M$, and taking the square-root of the eigenvalues. Using the decomposition we can rewrite equation (\ref{eq:elly}) as

\begin{equation}
    y B B^T y^T = c^2
\end{equation}

therefore $y B$ defines a circle of radius $c$. Thus to find points on the ellipse we first sample points on a circle and then use $B$ to find the corresponding points $y$.

\section{A projection pursuit index using the projected ellipse}
A shortfall of this orthogonal projection is that some interesting points, that are many standard deviations from $\v{\mu}_E$ in $p$-dimensional space, can be projected inside the ellipse and not stand out. The proposed projection pursuit index aims to remedy this. Let $W$ be a matrix where each column is an 'interesting' $p$-dimensional point $\v{w}$. Let $F$ be a function on a projection matrix $P$, with parameter $W$, such that:
\begin{equation} \label{ppi}
    F(P; W) = \sum_{\v{w}\in W} (\v{w} - \v{\mu}_E)^T P^T (P A^{-1} P^T)^{-1} P(\v{w} - \v{\mu}_E)
\end{equation}
Notice that $P(\v{w} - \v{\mu}_E)$ is the 2D projection of the vector pointing from the centre of the ellipsoid to $\v{w}$. Furthermore, $(P A^{-1} P^T)^{-1}$ is the 2D projection of the hyperellipsoid. That is, the summand of (\ref{ppi}) gives the squared Mahalanobis distance of $P\v{w}$ from $P\v{\mu}_E$, with respect to the shape of the \emph{projected} ellipse. The 'optimal' projection, for a given $W$, is the matrix $P$ which maximises $F(P; W)$. A way of numerically finding $P$ to do this is detailed in the appendix.

One way of selecting an interesting $W$ is to focus on outliers. Specifically, $W$ can be chosen as the set of points that are more than $z$ standard deviations from $\v{\mu}_E$ in $p$-dimensional space. In the python application, the value of $z$ can be chosen by the user.

To make these changes more efficient, the data is sorted beforehand in order of increasing elliptical distance from $\v{\mu}_E$; we also keep a list of these elliptical distances. When the user selects some $z$, we calculate a cutoff elliptical distance $c$ from $z$ by equation (\ref{ztoc}). We can binary search our list of elliptical distances for the index of this $c$. Then we get $W$ by truncating our sorted data matrix at this index.

Alternatively, the user can select $W$ using the lasso tool, specifying $W$ to be the set of points whose projects land in the selected region.

\section{Manual tour functionality}
We have in our projection matrix $p$ columns, where the $i$-th column is the 2D projection of the $i$-th basis vector of our high-dimensional space. When we drag the $i$-th basis vector to a position $(A, B)$ in 2D space, we want to construct a new projection matrix with the $i$-th column equal to $(A, B)$, or as close to it as possible.

Firstly, as $(A, B)$ is the orthogonal projection of a unit vector, we must have $|(A, B)| \le 1$. We normalise this vector if the user drags further out.

Let $\v{u}$ and $\v{v}$ be the first and second rows of our new projection matrix $P$, so that the $i$-th component of $\v{u}$ is $A$ and the $i$-th component of $\v{v}$ is $B$. Let $\v{a}$ and $\v{b}$ be $p-1$-dimensional vectors, obtained by removing the $i$-th component from $\v{u}$ and $\v{v}$ respectively. To have $\v{u}$ and $\v{v}$ perpendicular we require:
\[
0 = \v{u}\cdot \v{v} = \v{a} \cdot \v{b} + AB
\]
so that $\v{a} \cdot \v{b} = -AB$. Then, to have $\v{u}$ and $\v{v}$ normal, we require $|\v{a}| = \sqrt{1 - A^2}$, and $|\v{b}| = \sqrt{1 - B^2}$. Let $\theta$ be the angle between $\v{a}$ and $\v{b}$. Combining our previous requirements, we get 
\begin{equation} \label{cos}
\cos(\theta) = \frac{\v{a}\cdot \v{b}}{|\v{a}||\v{b}|} = \frac{-AB}{\sqrt{(1 - A^2)(1 - B^2)}}
\end{equation}
We can then construct $\v{a}$ and $\v{b}$ to be as close to the previous projection as possible, using their previous values $\v{a^*}$ and $\v{b^*}$. We set $\v{p}$ to be the normal vector in the direction of $\v{a^*}+\v{b^*}$, and $\v{m}$ to be the normal vector in the direction of $\v{a^*} - \v{b^*}$. We can find $\cos(\theta/2)$ and $\sin(\theta/2)$ using:
\begin{align*}
    \cos(\theta/2) = \sqrt{\frac{1 + \cos(\theta)}{2}}; & & \sin(\theta/2) = \sqrt{\frac{1 - \cos(\theta)}{2}}
\end{align*}

Then, set
\begin{align*}
    \v{a} = \sqrt{1 - A^2}(\cos(\theta/2) \v{p} + \sin(\theta/2) \v{m})\\
    \v{b} = \sqrt{1 - B^2}(\cos(\theta/2) \v{p} - \sin(\theta/2) \v{m})
\end{align*}
It is then trivial to check, using (\ref{cos}), the orthonormality of $\v{p}$ and $\v{m}$, and the identity $\cos(\theta) = \cos^2(\theta/2) - \sin^2(\theta/2)$, that $\v{a} \cdot \v{b} = -AB$, $|\v{a}| = \sqrt{1 - A^2}$, and $|\v{b}| = \sqrt{1 - B^2}$. The corresponding vectors $\v{u}$ and $\v{v}$ can then be stacked back into a new projection matrix.

\textbf{Error handling.} This method struggles when $\v{a^*} + \v{b^*}$ or $\v{a^*} - \v{b^*}$ turns out to be the zero vector. This is a frequent issue, occurring whenever the axes not being dragged are projected onto one line. In this case, the code sets the $i$-th components of $\v{u}$ and $\v{v}$ to $A$ and $B$ respectively, then reverts to the orthonormalisation method detailed in section \ref{sec:orthonormal}.

\section{Appendix: Numerically optimising the projection pursuit index}
Let $C = A^{-1}$ be the covariance matrix of the data. For brevity we will suppose our dataset has been centred at the origin. As the matrix of the projected ellipse is $(P C P^T)^{-1}$, we need to find $P$ to maximise
\[
\sum_{\v{w}\in W} \v{w}^T P^T (P C P^T)^{-1} P\v{w}.
\]
Let the rows of $P$ be $\v{u}$ and $\v{v}$. Suppose $W$ has $p$ rows with $m$ columns, corresponding to $m$ data points in $p$ dimensions. We need a multivariate function $f$ to maximise. Define $f: \mathbb{R}^{2n} \to \mathbb{R}$ as:
\begin{align*}
    f(\v{u}, \v{v}) :&= \sum_{\v{w}\in W} \v{w}^T
    \begin{bmatrix}
        \vert & \vert \\
        \v{u} & \v{v}  \\
        \vert & \vert
    \end{bmatrix}
    \left(
    \begin{bmatrix}
        \text{---} & \v{u} & \text{---}  \\
        \text{---} & \v{v} & \text{---}
    \end{bmatrix}
    C
    \begin{bmatrix}
        \vert & \vert \\
        \v{u} & \v{v}  \\
        \vert & \vert
    \end{bmatrix}
    \right)^{-1} 
    \begin{bmatrix}
        \text{---} & \v{u} & \text{---}  \\
        \text{---} & \v{v} & \text{---}
    \end{bmatrix} 
    \v{w} \\
    &= 
    \sum_{\v{w}\in W} \v{w}^T
    \begin{bmatrix}
        \v{u}\cdot\v{w} & \v{v}\cdot\v{w}
    \end{bmatrix}
    \begin{bmatrix}
        \v{u}^T C \v{u} & \v{u}^T C \v{v}\\
        \v{v}^T C \v{u} & \v{v}^T C \v{v}
    \end{bmatrix}
    ^{-1} 
    \begin{bmatrix}
        \v{u}\cdot\v{w}  \\
        \v{v}\cdot\v{w}
    \end{bmatrix} 
    \v{w}\\
    &=
    \sum_{\v{w}\in W} \frac{
    \begin{bmatrix}
        \v{u}\cdot\v{w} & \v{v}\cdot\v{w}
    \end{bmatrix}
    \begin{bmatrix}
        \v{v}^T C \v{v} & -\v{u}^T C \v{v}\\
        -\v{v}^T C \v{u} & \v{u}^T C \v{u}
    \end{bmatrix}
    \begin{bmatrix}
        \v{u}\cdot\v{w}  \\
        \v{v}\cdot\v{w}
    \end{bmatrix} 
    }
    {(\v{u}^T C \v{u})(\v{v}^T C \v{v}) - (\v{u}^T C \v{v})^2}\\
    &=
    \sum_{\v{w}\in W} \frac{
    (\v{u}^TC\v{u})(\v{v}\cdot\v{w})^2 
    - 2 (\v{u}^TC\v{v})(\v{u}\cdot\v{w})(\v{v}\cdot\v{w})
    + (\v{v}^TC\v{v})(\v{u}\cdot\v{w})^2 
    }
    {(\v{u}^T C \v{u})(\v{v}^T C \v{v}) - (\v{u}^T C \v{v})^2}
\end{align*}
With $\odot$ denoting element-wise multiplication, let $\v{g}: \mathbb{R}^{2n} \to \mathbb{R}^m$ be the function
\[
\v{g}(\v{u}, \v{v}) := (\v{u}^TC\v{u})(\v{v}^TW\odot\v{v}^TW)
    - 2 (\v{u}^TC\v{v})(\v{u}^TW\odot\v{v}^TW)
    + (\v{v}^TC\v{v})(\v{u}^TW\odot\v{u}^TW)
\]
and let $h: \mathbb{R}^{2n} \to \mathbb{R}$ be
\[
h(\v{u}, \v{v}) := (\v{u}^T C \v{u})(\v{v}^T C \v{v}) - (\v{u}^T C \v{v})^2.
\]
With $\v{s}$ as the $1 \times m$ summation vector (vector of ones), we have $f = \v{g}\v{s}/h$. Then the vector $(d\v{u}, d\v{v})$ giving the direction of the steepest increase of $f$ is
\begin{align*}
\nabla f(\v{u}, \v{v}) &= \left(f_{\v{u}}(\v{u}, \v{v}), f_{\v{v}}(\v{u}, \v{v})\right)\\
    &= \left(f_{\v{u}}(\v{u}, \v{v}), f_{\v{u}}(\v{v}, \v{u})\right) \text{\;\; by symmetry of $f$,}\\
    &= \left(
        \frac{h(\v{u}, \v{v})\v{g}_{\v{u}}(\v{u}, \v{v})\v{s} - \v{g}(\v{u}, \v{v}) \v{s} h_{\v{u}}(\v{u}, \v{v})}{(h(\v{u}, \v{v}))^2}, 
        \frac{h(\v{u}, \v{v})\v{g}_{\v{u}}(\v{v}, \v{u})\v{s} - \v{g}(\v{u}, \v{v}) \v{s} h_{\v{u}}(\v{v}, \v{u})}{(h(\v{u}, \v{v}))^2}
    \right).
\end{align*}
Differentiating, we calculate $\v{g}_{\v{u}}: \mathbb{R}^{2n} \to \mathbb{R}^{n \times m}$ to be
%2 * (v @ C @ v.T) * (u @ W) * W - 2*np.outer(C @ v, (u @ W)*(v @ W)) - 2*(u @ C @ v.T) * (v @ W) * W + 2 * np.outer(C @ u, (v @ W)**2)
\[
\v{g}_{\v{u}} (\v{u}, \v{v}) = 2(\v{v}^T C \v{v}) (\v{u}^TW)\odot W - 2 C\v{v} \otimes (\v{u}^TW \odot \v{v}^TW)
 - 2(\v{u}^T C \v{v}) (\v{v}^T W) \odot W + 2 C\v{u} \otimes (\v{v}^TW \odot \v{v}^TW)
\]
and have $h_{\v{u}}: \mathbb{R}^{2n} \to \mathbb{R}^{n}$ given by:
%2*(v @ C @ v.T)*(C @ u.T) - 2*(u @ C @ v.T)*(C @ v.T)
\[
h_{\v{u}}(\v{u}, \v{v}) = 2(\v{v}^T C \v{v}) C \v{u} - 2(\v{u}^T C \v{v}) C \v{v}.
\]
This gives us the means to calculate $\nabla f(\v{u}, \v{v})$. From starting values of $\v{u}$ and $\v{v}$, the optimiser enters a loop; it calculates $\nabla f$, scales it down to a size \texttt{step} (default 0.01), increments $\v{u}$ and $\v{v}$ by this vector, re-orthonormalises them, and recalculates $f$, keeping track of the increase $\Delta f$. When $\Delta f$ falls below a given \texttt{tol} (default 0.00001), we exit the loop and return $\v{u}$ and $\v{v}$.

\textbf{Finding the starting values.}
I wanted this algorithm to be deterministic, and independent of the order of the dimensions in the input file. The code attempts to start with a significant $\v{u}$ and $\v{v}$. It is possible to algebraically find a projection matrix $P$ to maximise:

\[
\sum_{\v{w}\in W} \v{w}^T P^T P A P^T P\v{w}.
\]
As $A$ is a positive definite matrix, it has a Cholesky decomposition $A = BB^T$. Substituting this in, we are maximising
\begin{align*}
&\sum_{\v{w}\in W} \v{w}^T P^T P B B^T P^T P\v{w}\\
=& \sum_{\v{w}\in W} |B^T P^T P\v{w}|^2
\end{align*}
Let $T: = B^{-1}W$, with columns $\v{t} = B^{-1}\v{w}$. 
Let $K := PB$ be a $2 \times n$ projection matrix, so that $B^T P^T P \v{w} = B^T P^T P B \v{t} = K^T K \v{t}$. Now we aim to find $K$ to maximise 
\begin{align*}
\sum_{\v{t}\in T} |K^T K\v{t}|^2.
\end{align*}
That is, $K$ is the best fitted plane to $T$, or equivalently the best \emph{rank 2 approximation}. The method for finding this is well-documented.
First, we modify $T$, setting $\hat{T}= (T - TS/m)^T$, where $S$ is the $m \times m$ matrix of ones. That is, $\hat{T}$ is the translation of $T$ to be centred at the origin.

Now we can calculate the singular value decomposition $\bar{T} = U \Sigma V^T$, where $U$ is $m \times m$ orthonormal, $\Sigma$ is $m\times n$ diagonal with entries decreasing along the diagonal, and $V$ is $n \times n$ orthonormal. Then the first two rows of $V^T$ give the best rank-2 approximation to $\bar{T}$ (see \url{https://www.cs.cmu.edu/~avrim/598/chap3only.pdf}); that is, they give the plane of best fit. These rows are already orthonormal, so we can set $K = V^T[:2]$.

Having found $K$, we can set $P = K B^{-1}$. We orthonormalise the rows of $P$ to get our initial $\v{u}$ and $\v{v}$ and start our numerical optimising process.

\end{document}

### 2024-06-25 paper.qmd
start on paper

---
title: "Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
---

## Introduction

Linear projections are useful in many aspects of statistical analysis of multivariate data, and especially useful for visualising the data. A linear projection provides a dimension reduction while maintaining interpretability. For example, a biplot (REF) shows the structure creating the maximum variance in the data, and also visualizing the projection matrix to learn which variables contribute to it. We might find clusters of outliers that were hiding in high dimensions.

More generally, projection pursuit (REF) defines a quantitative criterion for the interestingness of a projection (a projection pursuit index), and searches the space of possible projections for the most interesting one to display. We can also define sequences of interpolated linear projections to better understand a multivariate distribution. Animating a randomly selected interpolated sequence of linear projections is called a grand tour (REF). The combination of these two approaches would then use a projection pursuit index to select interesting projections, but display them via an interpolated path to provide context. This is called a guided tour (REF).

The question is whether we can use these techniques to assess new data samples in the context of an established normal, such as a specific multivariate normal distribution. In physics, the normal distribution may describe experimental results, or a global fit for a selected model, and we might want to compare to a set of other models. In medical applications, the normal distribution might summarize historic data of a healthy population and we compare it to samples from new patients. In outlier detection we might use robust measures to define the normal distribution and look for anomalies.

This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution. It is organised as follows. @sec-background provides more context for the methods and visualisation. @sec-anomaly-index provides the details of the new index, and example use is illustrated in @sec-examples.

## Background {#sec-background}

### Projection pursuit

### Tours for visualisation

## Anomaly index {#sec-anomaly-index}

## Examples {#sec-examples}

## Conclusion

### 2024-06-25 paper.tex
start on paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
  pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{\bf Are you normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi, Ursula Laa, Dianne Cook\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}

\end{abstract}

\newpage
\spacingset{1.9} % DON'T change the spacing!

\section{Introduction}\label{introduction}

Linear projections are useful in many aspects of statistical analysis of
multivariate data, and especially useful for visualising the data. A
linear projection provides a dimension reduction while maintaining
interpretability. For example, a biplot (REF) shows the structure
creating the maximum variance in the data, and also visualizing the
projection matrix to learn which variables contribute to it. We might
find clusters of outliers that were hiding in high dimensions.

More generally, projection pursuit (REF) defines a quantitative
criterion for the interestingness of a projection (a projection pursuit
index), and searches the space of possible projections for the most
interesting one to display. We can also define sequences of interpolated
linear projections to better understand a multivariate distribution.
Animating a randomly selected interpolated sequence of linear
projections is called a grand tour (REF). The combination of these two
approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (REF).

The question is whether we can use these techniques to assess new data
samples in the context of an established normal, such as a specific
multivariate normal distribution. In physics, the normal distribution
may describe experimental results, or a global fit for a selected model,
and we might want to compare to a set of other models. In medical
applications, the normal distribution might summarize historic data of a
healthy population and we compare it to samples from new patients. In
outlier detection we might use robust measures to define the normal
distribution and look for anomalies.

This paper describes a new projection pursuit index which is optimized
by projections where a new sample is most distant from the existing
normal distribution. It is organised as follows.
Section~\ref{sec-background} provides more context for the methods and
visualisation. Section~\ref{sec-anomaly-index} provides the details of
the new index, and example use is illustrated in
Section~\ref{sec-examples}.

\section{Background}\label{sec-background}

\subsection{Projection pursuit}\label{projection-pursuit}

\subsection{Tours for visualisation}\label{tours-for-visualisation}

\section{Anomaly index}\label{sec-anomaly-index}

\section{Examples}\label{sec-examples}

\end{document}

### 2024-06-25 data/liver.R

@@ -57,11 +57,12 @@ vc_f <- vc_f/4
 liver_f_means <- c(-0.3, 0.5, -0.1, 0.1)
 set.seed(523)
 liver_f <- rmvnorm(54, mean=liver_f_means, sigma=as.matrix(vc_f))
+colnames(liver_f) <- colnames(liver_stats)[c(1,2,4,5)]
 
 x <- rbind(liver_f, liver_norm)
 x <- data.frame(x)
 x$type <- factor(c(rep("f", 54), rep("norm", 193)))
-ggscatmat(x, columns = 1:4, color="type")
+ggscatmat(x, columns = 1:4, color="type", alpha=0.5)
 
 norm_mu <- liver_f_means*(-1)
 norm_mu <- matrix(norm_mu, ncol=4)
@@ -77,8 +78,7 @@ animate_xy(liver_f,
            ellipse=as.matrix(norm_vc), 
            ellc = qchisq(0.99, 4),
            ellmu = t(norm_mu), 
-           axes="off", half_range=6)
-colnames(liver_f) <- colnames(liver_stats)[c(1,2,4,5)]
+           axes="bottomleft", half_range=6)
 write_csv(as.data.frame(liver_f), file="data/liver_f.csv")
 
 colnames(vc_f) <- colnames(liver_stats)[c(1,2,4,5)]

### 2024-06-25 paper.qmd
ellipse example

@@
 ---
+
+```{r include=FALSE}
+# Set up chunk for for knitr
+knitr::opts_chunk$set(
+  fig.width = 5,
+  fig.height = 5,
+  fig.align = "center",
+  out.width = "100%",
+  code.line.numbers = FALSE,
+  fig.retina = 4,
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  cache = FALSE
+)
+```
+
+```{r}
+#| label: load-libraries
+#| warning: false
+#| echo: false
+library(tourr)
+library(mulgar)
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+```
+
+```{r}
+#| label: plot-theme
+theme_set(theme_linedraw() +
+   theme(
+     aspect.ratio = 1,
+     plot.background = element_rect(fill = 'transparent', colour = NA),
+     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
+     panel.background = element_rect(fill = 'transparent', 
+                                     colour = NA),
+     panel.grid.major = element_blank(), 
+     panel.grid.minor = element_blank(), 
+     axis.title.x = element_blank(), axis.title.y = element_blank(),
+     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
+     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+     legend.background = element_rect(fill = 'transparent', 
+                                      colour = NA),
+     legend.key = element_rect(fill = 'transparent', 
+                               colour = NA),
+     legend.position = "bottom", 
+     legend.title = element_blank(), 
+     legend.text = element_text(size=4),
+     legend.key.height = unit(0.25, 'cm'),
+     legend.key.width = unit(0.25, 'cm')
+   )
+)
+interior_annotation <- function(label, position = c(0.92, 0.92)) {
+  annotation_custom(grid::textGrob(label = label,
+      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
+      gp = grid::gpar(cex = 1, col="grey70")))
+}
+```
 
@@
 
-### Projection pursuit
+To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points.
+The norm is represented by points on the surface of a $p$-dimensional ellipse, corresponding to a confidence level.
+A sample of points uniformly distributed on a $p$-dimensional sphere is generated by
 
-### Tours for visualisation
+1. Simulating a sample of observations from $N_p(\mu, \Sigma)$. 
+2. Transforming each observation to have unit distance from the mean.
+
+```{r}
+#| echo: false
+#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1.
+A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
+set.seed(717)
+d1 <- matrix(rnorm(716 * 2), ncol = 2) 
+d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
+vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
+evc <- eigen(vc)
+vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
+d3 <- d2 %*% vc2 |>
+  as.data.frame()
+
+d4 <- d3 |>
+  bind_rows(data.frame(V1=0.6, V2=0.6)) |>
+  bind_rows(data.frame(V1=0.3, V2=0.3)) |>
+  mutate(type = c(rep("ci", nrow(d3)), "in", "out"))
+
+d1 <- as.data.frame(d1)
+d2 <- as.data.frame(d2)
+d3 <- as.data.frame(d3)
+
+p1 <- ggplot(d1, aes(x=V1, y=V2)) +
+  geom_point() +
+  xlab(expression(x[1])) +
+  ylab(expression(x[2])) +
+  ggtitle("a. Bivariate standard normal sample") +
+  theme(axis.text = element_blank())
+
+p2 <- ggplot(d2, aes(x=V1, y=V2)) +
+  geom_point() +
+  xlab(expression(x[1])) +
+  ylab(expression(x[2])) +
+  ggtitle("b. Uniform on a sphere") +
+  theme(axis.text = element_blank())
+
+p3 <- ggplot(d3, aes(x=V1, y=V2)) +
+  geom_point() +
+  xlab(expression(x[1])) +
+  ylab(expression(x[2])) +
+  ggtitle("c. Points on surface of confidence ellipse.") +
+  theme(axis.text = element_blank())
+
+p4 <- ggplot(d4, aes(x=V1, y=V2, shape=type)) +
+  geom_point() +
+  xlab(expression(x[1])) +
+  ylab(expression(x[2])) +
+  ggtitle("d. Observations inside and outside confidence ellipse.") +
+  theme(legend.position = "none", 
+        axis.text = element_blank())
+
+p1 + p2 + p3 + p4 + plot_layout(ncol=2)
+```
 

### 2024-06-25 paper.tex
ellipse example

@@
 
-\subsection{Projection pursuit}\label{projection-pursuit}
+To compare a new sample with an existing norm, like a multivariate
+normal distribution, in higher than two dimensions, we have typically
+used two samples of points. The norm is represented by points on the
+surface of a \(p\)-dimensional ellipse, corresponding to a confidence
+level. A sample of points uniformly distributed on a \(p\)-dimensional
+sphere is generated by
 
-\subsection{Tours for visualisation}\label{tours-for-visualisation}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Simulating a sample of observations from \(N_p(\mu, \Sigma)\).
+\item
+  Transforming each observation to have unit distance from the mean.
+\end{enumerate}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}
+
+}
+
+\caption{Simulating a uniform sample on a sphere involves sampling from
+a multivariate normal (a) and transforming each observation to have
+length equal to 1. A confidence ellipse is generated by transforming the
+sphere relative to a specified variance-covariance matrix (c), and new
+observations can be visually assessed to be inside or outside by
+plotting with the ellipse (d).}
+
+\end{figure}%
 

### 2024-06-25 paper.qmd
finished explanaiton of ellipse

@@
     keep-tex: true    
+bibliography: bibliography.bib  
+header-includes: | 
+  \usepackage{amsmath}
+  \usepackage{float}
+  \usepackage{hyperref}
+  \usepackage[utf8]{inputenc}
+  \usepackage{bm}
+  \def\tightlist{}
+  \usepackage{setspace}
+  \newcommand\pD{$p\text{-}D$}
+  \newcommand\kD{$k\text{-}D$}
+  \newcommand\dD{$d\text{-}D$}
+  \newcommand\gD{$2\text{-}D$}
 ---
@@
 The combination of these two approaches would then use a projection pursuit index to select interesting projections, but display them via an interpolated path to provide context.
-This is called a guided tour (REF).
+This is called a guided tour (@cook1995).
 
@@
 
-1. Simulating a sample of observations from $N_p(\mu, \Sigma)$. 
-2. Transforming each observation to have unit distance from the mean.
+1. Simulating a sample of observations ($\mathbfit{x}$)
+from $N_p(\mu, \Sigma)$. 
+2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.
+
+To convert this to points on the surface of a confidence ellipse,
+
+3. transform the shape using a specific variance covariance, and shift to center on the mean vector.
+
+Finally, new observations can be visually compared with this ellipse by
+
+4. plotting them together.
+
+@fig-ci illustrates this process for \gD. This is easiest way to view this normal region relative to a new sample for any \pD{} problem.
 
@@
 #| echo: false
+#| label: fig-ci
 #| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1.
 A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
+#| out-width: 80%
 set.seed(717)

### 2024-06-25 paper.tex
finished explanaiton of ellipse

@@
 \addtolength{\topmargin}{-1in}%
+\usepackage{amsmath}
+\usepackage{float}
+\usepackage{hyperref}
+\usepackage[utf8]{inputenc}
+\usepackage{bm}
+\def\tightlist{}
+\usepackage{setspace}
+\newcommand\pD{$p\text{-}D$}
+\newcommand\kD{$k\text{-}D$}
+\newcommand\dD{$d\text{-}D$}
+\newcommand\gD{$2\text{-}D$}
 \makeatletter
@@
 interesting projections, but display them via an interpolated path to
-provide context. This is called a guided tour (REF).
+provide context. This is called a guided tour (\citet{cook1995}).
 
@@
 \item
-  Simulating a sample of observations from \(N_p(\mu, \Sigma)\).
+  Simulating a sample of observations (\(\mathbfit{x}\)) from
+  \(N_p(\mu, \Sigma)\).
 \item
-  Transforming each observation to have unit distance from the mean.
+  Transforming each observation to have unit distance from the mean,
+  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
 \end{enumerate}
 
-\begin{figure}[H]
+To convert this to points on the surface of a confidence ellipse,
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  transform the shape using a specific variance covariance, and shift to
+  center on the mean vector.
+\end{enumerate}
+
+Finally, new observations can be visually compared with this ellipse by
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  plotting them together.
+\end{enumerate}
+
+Figure~\ref{fig-ci} illustrates this process for \gD. This is easiest
+way to view this normal region relative to a new sample for any \pD{}
+problem.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}
 
@@
 
-\caption{Simulating a uniform sample on a sphere involves sampling from
-a multivariate normal (a) and transforming each observation to have
-length equal to 1. A confidence ellipse is generated by transforming the
-sphere relative to a specified variance-covariance matrix (c), and new
-observations can be visually assessed to be inside or outside by
-plotting with the ellipse (d).}
+\caption{\label{fig-ci}Simulating a uniform sample on a sphere involves
+sampling from a multivariate normal (a) and transforming each
+observation to have length equal to 1. A confidence ellipse is generated
+by transforming the sphere relative to a specified variance-covariance
+matrix (c), and new observations can be visually assessed to be inside
+or outside by plotting with the ellipse (d).}
 
@@
 
+\renewcommand\refname{Conclusion}
+  \bibliography{bibliography.bib}
+
 \end{document}

### 2024-06-25 paper.qmd
index definition

@@
 To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points.
-The norm is represented by points on the surface of a $p$-dimensional ellipse, corresponding to a confidence level.
+The norm is represented by points on the surface of a $p$-dimensional ellipsoid, corresponding to a confidence level.
 A sample of points uniformly distributed on a $p$-dimensional sphere is generated by
 
-1. Simulating a sample of observations ($\mathbfit{x}$)
-from $N_p(\mu, \Sigma)$. 
+1. Simulating a sample of observations ($\mathbfit{x}$, which are \pD{} vectors)
+from $N_p(\mathbfit{\mu}, \Sigma)$. 
 2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.
 
-To convert this to points on the surface of a confidence ellipse,
+To convert this to points on the surface of a confidence ellipsoid,
 
@@
 
-Finally, new observations can be visually compared with this ellipse by
+Finally, new observations can be visually compared with this ellipsoid by
 
@@
 #| label: fig-ci
+#| fig-width: 8
+#| fig-height: 2
 #| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1.
-A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
-#| out-width: 80%
+A confidence ellipsoid is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipsoid (d)."
 set.seed(717)
@@
 p1 <- ggplot(d1, aes(x=V1, y=V2)) +
-  geom_point() +
+  geom_point(alpha=0.5) +
   xlab(expression(x[1])) +
   ylab(expression(x[2])) +
-  ggtitle("a. Bivariate standard normal sample") +
+  xlim(c(-3.2, 3.2)) +
+  ylim(c(-3.2, 3.2)) +
+  interior_annotation("a") +
+  #ggtitle("a. Bivariate standard normal sample") +
   theme(axis.text = element_blank())
@@
 p2 <- ggplot(d2, aes(x=V1, y=V2)) +
-  geom_point() +
+  geom_point(alpha=0.5) +
   xlab(expression(x[1])) +
   ylab(expression(x[2])) +
-  ggtitle("b. Uniform on a sphere") +
+  xlim(c(-1.2, 1.2)) +
+  ylim(c(-1.2, 1.2)) +
+  interior_annotation("b") +
+  #ggtitle("b. Uniform on a sphere") +
   theme(axis.text = element_blank())
@@
 p3 <- ggplot(d3, aes(x=V1, y=V2)) +
-  geom_point() +
+  geom_point(alpha=0.5) +
   xlab(expression(x[1])) +
   ylab(expression(x[2])) +
-  ggtitle("c. Points on surface of confidence ellipse.") +
+  xlim(c(-1.2, 1.2)) +
+  ylim(c(-1.2, 1.2)) +
+  interior_annotation("c") +
+  #ggtitle("c. Points on surface of confidence ellipsoid.") +
   theme(axis.text = element_blank())
@@
 p4 <- ggplot(d4, aes(x=V1, y=V2, shape=type)) +
-  geom_point() +
+  geom_point(alpha=0.5) +
   xlab(expression(x[1])) +
   ylab(expression(x[2])) +
-  ggtitle("d. Observations inside and outside confidence ellipse.") +
+  xlim(c(-1.2, 1.2)) +
+  ylim(c(-1.2, 1.2)) +
+  interior_annotation("d") +
+  #ggtitle("d. Observations inside and outside confidence ellipsoid.") +
   theme(legend.position = "none", 
@@
 
-p1 + p2 + p3 + p4 + plot_layout(ncol=2)
+p1 + p2 + p3 + p4 + plot_layout(ncol=4)
 ```
 
+<PLOTS FROM USCHI's PHYSICS EXAMPLE HERE>
+
+Although this is flexible, this does not make it easy to guide the tour towards the directions (projections) where the samples are most different from the normal.
+What would be desirable is to analytically define the confidence ellipsoid, compute flag observations that are outside, steer the tour to projections that reveal the extent of the difference.
+And also display the projected ellipsoid as a geometric shape rather than a sample of points.
+These are the procedures that are described in the next section.
+
 ## Anomaly index {#sec-anomaly-index}
+
+## Projecting an ellipsoid
+
+A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by
+
+$$
+(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
+$$
+where $c$ a constant that depends on a specific confidence level.
+
+The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
+We call points in the projection that are on the curve $\mathbfit{y}$.
+From this we can compute the analog equation for the projection as
+
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
+where $P$ is a $(p\times 2)$ orthonormal basis defining the projection and $\mathbfit{\mu}_p = \mathbfit{\mu} P$ the projected mean.
+This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection.
+In general $c$ could be any constant, but typically we would select it as a quantile of the $\chi^2$ distribution, so that the size of the ellipse corresponds to a selected probability.
+
+<IMAGES OF PROJECTED ELLIPSE>
+
+### Index specification
+
+To define a measure of an interesting projection is to maximize the **average Malahanobis distance in the projection** for a subset of points, $W$.
+The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}.
+Alternatives could be to select a set of observations with the largest Mahalanobis distance, manually select observations or possibly a group of points identified via clustering of the extremes.
+
+The index is written as
+
+$$
+\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
+$$
+
+where by default $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$, is the set of observations outside the \pD{} ellipsoid.
+
+## Implementation
+
+This was implemented in the `tourr` (REF) package, where the projected ellipsoid can be drawn for each projection.
 

### 2024-06-25 paper.tex
index definition

@@
 used two samples of points. The norm is represented by points on the
-surface of a \(p\)-dimensional ellipse, corresponding to a confidence
+surface of a \(p\)-dimensional ellipsoid, corresponding to a confidence
 level. A sample of points uniformly distributed on a \(p\)-dimensional
@@
 \item
-  Simulating a sample of observations (\(\mathbfit{x}\)) from
-  \(N_p(\mu, \Sigma)\).
+  Simulating a sample of observations (\(\mathbfit{x}\), which are \pD{}
+  vectors) from \(N_p(\mathbfit{\mu}, \Sigma)\).
 \item
@@
 
-To convert this to points on the surface of a confidence ellipse,
+To convert this to points on the surface of a confidence ellipsoid,
 
@@
 
-Finally, new observations can be visually compared with this ellipse by
+Finally, new observations can be visually compared with this ellipsoid
+by
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}
 
@@
 sampling from a multivariate normal (a) and transforming each
-observation to have length equal to 1. A confidence ellipse is generated
-by transforming the sphere relative to a specified variance-covariance
-matrix (c), and new observations can be visually assessed to be inside
-or outside by plotting with the ellipse (d).}
+observation to have length equal to 1. A confidence ellipsoid is
+generated by transforming the sphere relative to a specified
+variance-covariance matrix (c), and new observations can be visually
+assessed to be inside or outside by plotting with the ellipsoid (d).}
 
@@
 
+\textless PLOTS FROM USCHI's PHYSICS EXAMPLE HERE\textgreater{}
+
+Although this is flexible, this does not make it easy to guide the tour
+towards the directions (projections) where the samples are most
+different from the normal. What would be desirable is to analytically
+define the confidence ellipsoid, compute flag observations that are
+outside, steer the tour to projections that reveal the extent of the
+difference. And also display the projected ellipsoid as a geometric
+shape rather than a sample of points. These are the procedures that are
+described in the next section.
+
 \section{Anomaly index}\label{sec-anomaly-index}
+
+\section{Projecting an ellipsoid}\label{projecting-an-ellipsoid}
+
+A \pD{} ellipsoid corresponding to a given variance-covariance
+(\(\Sigma\)) is defined by
+
+\[
+(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
+\] where \(c\) a constant that depends on a specific confidence level.
+
+The projection of an ellipsoid onto \gD{} is an ellipse, where the curve
+of the ellipse is defined through the set of points \(\mathbfit{x}\) for
+which the gradient is parallel to the projection plane. We call points
+in the projection that are on the curve \(\mathbfit{y}\). From this we
+can compute the analog equation for the projection as
+
+\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2\]
+where \(P\) is a \((p\times 2)\) orthonormal basis defining the
+projection and \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) the projected
+mean. This means the matrix \((P^T \Sigma P)^{-1}\) is defining the
+ellipse in the \gD{} projection. In general \(c\) could be any constant,
+but typically we would select it as a quantile of the \(\chi^2\)
+distribution, so that the size of the ellipse corresponds to a selected
+probability.
+
+\subsection{Index specification}\label{index-specification}
+
+To define a measure of an interesting projection is to maximize the
+\textbf{average Malahanobis distance in the projection} for a subset of
+points, \(W\). The set of points could be chosen in different ways, but
+the default is those that are outside the specified ellipsoid in \pD{}.
+Alternatives could be to select a set of observations with the largest
+Mahalanobis distance, manually select observations or possibly a group
+of points identified via clustering of the extremes.
+
+The index is written as
+
+\[
+\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
+\]
+
+where by default
+\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\),
+is the set of observations outside the \pD{} ellipsoid.
+
+\section{Implementation}\label{implementation}
+
+This was implemented in the \texttt{tourr} (REF) package, where the
+projected ellipsoid can be drawn for each projection.
 

### 2024-06-25 paper.qmd
sample code in implementation

@@
 
-This was implemented in the `tourr` (REF) package, where the projected ellipsoid can be drawn for each projection.
+This is implemented in the `tourr` (REF) package, where the projected ellipsoid can be drawn for each projection.
+The guided tour will take arguments specifying the data, and the null variance-covariance matrix.
+
+```{r}
+#| eval: false
+#| echo: true
+library(tourr)
+library(mulgar)
+set.seed(929)
+vc_null <- matrix(rep(0.5, 5*5), ncol=5)
+diag(vc_null) <- 1
+m_null <- rep(0, 5)
+vc_samp <- matrix(rep(0, 5*5), ncol=5)
+diag(vc_samp) <- 1
+vc_samp[4,5] <- -0.47
+vc_samp[5,4] <- -0.56
+vc_samp <- vc_samp*0.1
+m_samp <- c(0, 0, 0, 1.9, 2.3)
+samp <- as.data.frame(rmvn(6, 
+                           mn = m_samp, 
+                           vc = vc_samp))
+animate_xy(samp, guided_anomaly_tour(anomaly_index(),
+  ellipse=vc_null), ellipse=vc_null, 
+  axes = "bottomleft", half_range=5, center=FALSE)
+```
+
+::: {#fig-anomaly layout-ncol=2}
+
+![Random projection.](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}
+
+![Optimal projection.](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}
+
+Two projections of simulated example data corresponding to the sample code: (a) Random projection where sample is inside the 2-D ellipse, (b) Optimal projection from index, showing most of the sample outside.
+A red cross indicates that the point is outside the p-D ellipse.
+The optimal projection uses mostly variables $x_4, x_5$, which is expected because these are the two directions where the sample most differs from the norm.
+:::
 

### 2024-06-25 paper.tex
sample code in implementation

@@
 \fi
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 
-This was implemented in the \texttt{tourr} (REF) package, where the
-projected ellipsoid can be drawn for each projection.
+This is implemented in the \texttt{tourr} (REF) package, where the
+projected ellipsoid can be drawn for each projection. The guided tour
+will take arguments specifying the data, and the null
+variance-covariance matrix.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tourr)}
+\FunctionTok{library}\NormalTok{(mulgar)}
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{929}\NormalTok{)}
+\NormalTok{vc\_null }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
+\FunctionTok{diag}\NormalTok{(vc\_null) }\OtherTok{\textless{}{-}} \DecValTok{1}
+\NormalTok{m\_null }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{)}
+\NormalTok{vc\_samp }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
+\FunctionTok{diag}\NormalTok{(vc\_samp) }\OtherTok{\textless{}{-}} \DecValTok{1}
+\NormalTok{vc\_samp[}\DecValTok{4}\NormalTok{,}\DecValTok{5}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.47}
+\NormalTok{vc\_samp[}\DecValTok{5}\NormalTok{,}\DecValTok{4}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.56}
+\NormalTok{vc\_samp }\OtherTok{\textless{}{-}}\NormalTok{ vc\_samp}\SpecialCharTok{*}\FloatTok{0.1}
+\NormalTok{m\_samp }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{1.9}\NormalTok{, }\FloatTok{2.3}\NormalTok{)}
+\NormalTok{samp }\OtherTok{\textless{}{-}} \FunctionTok{as.data.frame}\NormalTok{(}\FunctionTok{rmvn}\NormalTok{(}\DecValTok{6}\NormalTok{, }
+                           \AttributeTok{mn =}\NormalTok{ m\_samp, }
+                           \AttributeTok{vc =}\NormalTok{ vc\_samp))}
+\FunctionTok{animate\_xy}\NormalTok{(samp, }\FunctionTok{guided\_anomaly\_tour}\NormalTok{(}\FunctionTok{anomaly\_index}\NormalTok{(),}
+  \AttributeTok{ellipse=}\NormalTok{vc\_null), }\AttributeTok{ellipse=}\NormalTok{vc\_null, }
+  \AttributeTok{axes =} \StringTok{"bottomleft"}\NormalTok{, }\AttributeTok{half\_range=}\DecValTok{5}\NormalTok{, }\AttributeTok{center=}\ConstantTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/anomaly1.png}
+
+}
+
+\subcaption{\label{fig-anomaly1}Random projection.}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/anomaly2.png}
+
+}
+
+\subcaption{\label{fig-anomaly2}Optimal projection.}
+
+\end{minipage}%
+
+\caption{\label{fig-anomaly}Two projections of simulated example data
+corresponding to the sample code: (a) Random projection where sample is
+inside the 2-D ellipse, (b) Optimal projection from index, showing most
+of the sample outside. A red cross indicates that the point is outside
+the p-D ellipse. The optimal projection uses mostly variables
+\(x_4, x_5\), which is expected because these are the two directions
+where the sample most differs from the norm.}
+
+\end{figure}%
 

### 2024-06-25 paper.qmd
added part on the need to cluster

@@
 
+### Additional considerations
+
+If the observations in $W$ are primarily departing from the normal range in the same direction, the index will be expected to perform well in finding this average direction.
+However, if the observations have very different departures from the norm, it may be useful to break them into groups, and separately optimize on these subsets.
+One could consider clustering these observations using angular distance to find groups of observations that have similar directions of departure.
+
 ## Implementation
@@
 
-Two projections of simulated example data corresponding to the sample code: (a) Random projection where sample is inside the 2-D ellipse, (b) Optimal projection from index, showing most of the sample outside.
+Two projections of simulated example data corresponding to the sample code: (a) random projection where sample is inside the 2-D ellipse, (b) optimal projection from index, showing most of the sample outside.
 A red cross indicates that the point is outside the p-D ellipse.
@@
 
+### Medical patients
+
+### Robust statistics
+
+### Physics?
+
 ## Conclusion

### 2024-06-25 paper.tex
added part on the need to cluster

@@
 
+\subsection{Additional considerations}\label{additional-considerations}
+
+If the observations in \(W\) are primarily departing from the normal
+range in the same direction, the index will be expected to perform well
+in finding this average direction. However, if the observations have
+very different departures from the norm, it may be useful to break them
+into groups, and separately optimize on these subsets. One could
+consider clustering these observations using angular distance to find
+groups of observations that have similar directions of departure.
+
 \section{Implementation}\label{implementation}
@@
 \caption{\label{fig-anomaly}Two projections of simulated example data
-corresponding to the sample code: (a) Random projection where sample is
-inside the 2-D ellipse, (b) Optimal projection from index, showing most
+corresponding to the sample code: (a) random projection where sample is
+inside the 2-D ellipse, (b) optimal projection from index, showing most
 of the sample outside. A red cross indicates that the point is outside
@@
 
+\subsection{Medical patients}\label{medical-patients}
+
+\subsection{Robust statistics}\label{robust-statistics}
+
+\subsection{Physics?}\label{physics}
+
 \renewcommand\refname{Conclusion}

### 2024-06-27 paper.tex

@@
 of the ellipse is defined through the set of points \(\mathbfit{x}\) for
-which the gradient is parallel to the projection plane. We call points
-in the projection that are on the curve \(\mathbfit{y}\). From this we
+which the gradient is parallel to the projection plane. That is, the curve 
+consists of \(\mathbfit{x}\) satisfying
+\[
+\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+\]
+for some \(\mathbfit{s} \in \mathbb{R}^2\), where \(P\) is a \((p\times 2)\) orthonormal basis defining the
+projection. We can write \(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\).
+Making this substitution in the \pD{} ellipsoid equation yields
+\[
+c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+\]
+
+We call points in the projection that are on the curve \(\mathbfit{y}\), so that \(\mathbfit{y} = \mathbfit{x}P\)
+for \(\mathbfit{x}\) on the \pD{} curve. Then
+\(\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P\), where 
+\(\mathbfit{\mu}_p = \mathbfit{\mu} P\) is the projected mean. We then
+substitute \((\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}\) for \(\mathbfit{s}\) in the
+previous equation. From this we
 can compute the analog equation for the projection as
@@
 \[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2\]
-where \(P\) is a \((p\times 2)\) orthonormal basis defining the
-projection and \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) the projected
-mean. This means the matrix \((P^T \Sigma P)^{-1}\) is defining the
+
+This means the matrix \((P^T \Sigma P)^{-1}\) is defining the
 ellipse in the \gD{} projection. In general \(c\) could be any constant,

### 2024-06-27 paper.qmd
integrated Annalisa change

@@
 The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
-We call points in the projection that are on the curve $\mathbfit{y}$.
-From this we can compute the analog equation for the projection as
+That is, the curve
+consists of $\mathbfit{x}$ satisfying
 
-$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
-where $P$ is a $(p\times 2)$ orthonormal basis defining the projection and $\mathbfit{\mu}_p = \mathbfit{\mu} P$ the projected mean.
+$$(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+$$
+
+for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
+projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+Making this substitution in the \pD{} ellipsoid equation yields
+
+$$c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+$$
+
+We call points in the projection that are on the curve $\mathbfit{y}$. From this we can compute the analog equation for the projection as
+
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
 This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection.

### 2024-06-27 paper.tex
integrated Annalisa change

--- 2024-06-25 paper.tex
@@
 of the ellipse is defined through the set of points \(\mathbfit{x}\) for
-which the gradient is parallel to the projection plane. We call points
-in the projection that are on the curve \(\mathbfit{y}\). From this we
-can compute the analog equation for the projection as
+which the gradient is parallel to the projection plane. That is, the
+curve consists of \(\mathbfit{x}\) satisfying
 
-\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2\]
-where \(P\) is a \((p\times 2)\) orthonormal basis defining the
-projection and \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) the projected
-mean. This means the matrix \((P^T \Sigma P)^{-1}\) is defining the
-ellipse in the \gD{} projection. In general \(c\) could be any constant,
-but typically we would select it as a quantile of the \(\chi^2\)
+\[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+\]
+
+for some \(\mathbfit{s} \in \mathbb{R}^2\), where \(P\) is a
+\((p\times 2)\) orthonormal basis defining the projection. We can write
+\(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\). Making this
+substitution in the \pD{} ellipsoid equation yields
+
+\[c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+\]
+
+We call points in the projection that are on the curve \(\mathbfit{y}\).
+From this we can compute the analog equation for the projection as
+
+\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.\]
+This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
+the \gD{} projection. In general \(c\) could be any constant, but
+typically we would select it as a quantile of the \(\chi^2\)
 distribution, so that the size of the ellipse corresponds to a selected

### 2024-06-27 paper.qmd
fixing derivation

@@
 
-$$(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+$$\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
 $$
@@
 
-We call points in the projection that are on the curve $\mathbfit{y}$. From this we can compute the analog equation for the projection as
+We call points in the projection that are on the curve \(\mathbfit{y}\), so that \(\mathbfit{y} = \mathbfit{x}P\) for \(\mathbfit{x}\) on the \pD{} curve.
+Then \(\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P\), where \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) is the projected mean.
+We then substitute \((\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}\) for \(\mathbfit{s}\) in the previous equation.
+From this we can compute the analog equation for the projection as
 

### 2024-06-27 paper.qmd
fixes to tex for Annalias changes

--- 2024-06-27 paper.qmd
@@
 
-We call points in the projection that are on the curve $\mathbfit{y}$. From this we can compute the analog equation for the projection as
+We call points in the projection that are on the curve $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for $\mathbfit{x}$ on the \pD{} curve.
+Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
+We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the previous equation.
+From this we can compute the analog equation for the projection as
 

### 2024-07-01 paper.tex
fixes to tex for Annalias changes
example titles

@@
 
-\[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+\[\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
 \]
@@
 
-We call points in the projection that are on the curve \(\mathbfit{y}\).
-From this we can compute the analog equation for the projection as
+We call points in the projection that are on the curve \(\mathbfit{y}\),
+so that \(\mathbfit{y} = \mathbfit{x}P\) for \(\mathbfit{x}\) on the
+\pD{} curve. Then
+\(\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P\),
+where \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) is the projected mean. We
+then substitute
+\((\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}\) for
+\(\mathbfit{s}\) in the previous equation. From this we can compute the
+analog equation for the projection as
 
@@
 
-\subsection{Robust statistics}\label{robust-statistics}
-
-\subsection{Physics?}\label{physics}
+\subsection{Robust statistics: weather
+extremes}\label{robust-statistics-weather-extremes}
 

### 2024-07-01 paper.qmd
example titles
add theorem/proof structure

@@
   \usepackage{amsmath}
+  \usepackage{amsthm}
   \usepackage{float}
@@
   \newcommand\gD{$2\text{-}D$}
+  \newtheorem*{theorem}{Theorem}
 ---
@@
 
+\begin{theorem}
+  The projection of this \pD{} ellipsoid in \gD{} has the equation
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
+\end{theorem}
+\begin{proof}
 The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
@@
 Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
-We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the previous equation.
+We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the equation \(c^2 = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T\).
 From this we can compute the analog equation for the projection as
 
-$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
+
+as claimed.
+\end{proof}
+
 This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection.
@@
 
-### Robust statistics
-
-### Physics?
+### Robust statistics: weather extremes
 

### 2024-07-02 code/ShapleyOutlierExamples.R
Working on weather data as example

@@ -4,6 +4,7 @@
 library(ShapleyOutlier)
 library(robustHD)
 library(tidyverse)
+library(tourr)
 
 data(TopGear)
 
@@ -90,3 +91,90 @@ mu <-MCD$center
 Sigma <- MCD$cov
 Sigma_inv <- solve(MCD$cov)
 
+size_weather <- qchisq(2*pnorm(5)-1, ncol(X))
+
+set.seed(129)
+animate_xy(X, 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellc = size_weather), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           axes="off")
+
+## now look for angular clusters
+dist_pd <- mahalanobis(X, mu, Sigma)
+X_outside <- X[dist_pd > size_weather,]
+X_outside_scaled <- t(apply(X_outside,
+                            1,
+                            function(x) (x)/sqrt(sum(x^2))))
+#check that it worked
+rowSums(X_outside_scaled^2)
+set.seed(150)
+X_km <- kmeans(X_outside_scaled, 4)
+X_km$cluster
+
+# cluster 1: high temperature
+# cluster 2: 
+# cluster 3: temperature + sun_h, num_clear
+# cluster 4: avg_t_min small for the outlying points
+X_outside_2 <- X_outside[X_km$cluster == 4,]
+set.seed(129)
+animate_xy(X_outside_2, 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellmu = mu,
+                               ellc = size_weather), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           ellmu = mu, center = FALSE,
+           axes="off")
+
+animate_xy(X_outside_2, 
+           grand_tour(),
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           ellmu = mu, center = FALSE,
+           axes="off", half_range = 7)
+
+#### try again but with a bit smaller p
+
+X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year,
+                                        wind_v, num_precp_01))
+rownames(X) <- weather_summer$year
+#> Warning: Setting row names on a tibble is deprecated.
+X <- robStandardize(X)
+as.data.frame(X) |>
+  rownames_to_column("year") |>
[…]

### 2024-07-02 code/weather.R

@@ -1,0 +1,71 @@
+# script following the vignette of ShapleyOutlier
+# for weather data example
+
+library(ShapleyOutlier)
+library(robustHD)
+library(tidyverse)
+library(tourr)
+
+data("WeatherVienna")
+
+weather_summer <- WeatherVienna %>% dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) %>%
+  drop_na() %>%
+  filter(month %in% c("JUN", "JUL", "AUG")) %>%
+  filter(year >= 1955) %>%
+  group_by(year) %>%
+  dplyr::select(-month) %>%
+  summarise(across(.cols = everything(), function(x) mean(x)))
+
+X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year))
+rownames(X) <- weather_summer$year
+#> Warning: Setting row names on a tibble is deprecated.
+X <- robStandardize(X)
+as.data.frame(X) |>
+  rownames_to_column("year") |>
+  select(-year) |>
+  write_csv("data_weather.csv")
+
+set.seed(1)
+MCD <- covMcd(X, alpha = 0.5, nsamp = "best")
+#> Warning in .fastmcd(x, h, nsamp, nmini, kmini, trace = as.integer(trace)): 'nsamp = "best"' allows maximally 100000 subsets;
+#> computing these subsets of size 17 out of 68
+mu <-MCD$center
+Sigma <- MCD$cov
+Sigma_inv <- solve(MCD$cov)
+
+size_weather <- qchisq(2*pnorm(5)-1, ncol(X))
+
+set.seed(129)
+animate_xy(X, 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellc = size_weather), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           axes="off")
+
+## now look for angular clusters
+dist_pd <- mahalanobis(X, mu, Sigma)
+X_outside <- X[dist_pd > size_weather,]
+X_outside_scaled <- t(apply(X_outside,
+                            1,
+                            function(x) (x)/sqrt(sum(x^2))))
+#check that it worked
+rowSums(X_outside_scaled^2)
+set.seed(150)
+X_km <- kmeans(X_outside_scaled, 4)
+X_km$cluster
+
+# cluster 1: high temperature
+# cluster 2: 
+# cluster 3: temperature + sun_h, num_clear
+# cluster 4: avg_t_min small for the outlying points
[…]

### 2024-07-02 code/weather.R
Selecting k for weather data with dunn2 index

@@ -5,6 +5,7 @@ library(ShapleyOutlier)
 library(robustHD)
 library(tidyverse)
 library(tourr)
+library(fpc)
 
 data("WeatherVienna")
 
@@ -51,6 +52,16 @@ X_outside_scaled <- t(apply(X_outside,
                             function(x) (x)/sqrt(sum(x^2))))
 #check that it worked
 rowSums(X_outside_scaled^2)
+for(k in 2:6){
+  set.seed(150)
+  X_km <- kmeans(X_outside_scaled, k)
+  cs <- cluster.stats(dist(X_outside_scaled), X_km$cluster)
+  print(cs$dunn2)
+}
+
+#preferred solution 4-5 clusters using dunn2/dunn index
+#will work with k=4
+
 set.seed(150)
 X_km <- kmeans(X_outside_scaled, 4)
 X_km$cluster
@@ -69,3 +80,14 @@ animate_xy(X_outside_2,
            ellipse=as.matrix(Sigma), ellc = size_weather,
            ellmu = mu, center = FALSE,
            axes="off")
+
+
+# steps:
+# 1 - find outliers, split them into angular clusters
+# 1a - use some cluster metric to decide on k
+# 2 - with all observations run anomaly index, show the
+# clusters in different colors
+# 2a - define an axis display that only labels "long" axes
+# 3 - for some subset(s) of interest run anomaly index
+# only on the subset
+

### 2024-07-02 paper.qmd
liver example added

@@
 library(patchwork)
+library(knitr)
+library(kableExtra)
 ```
@@
 
-### Medical patients
+### Health: liver function tests
+
+This example is motivated by a problem posed during consulting with a pharmaceutical company, but the data shown here is simulated, simply to illustrate the application.
+Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).
+There are normal ranges on these measurements reported by @lib-med-liver-norms and listed in @tbl-liver-norms.
+
+```{r}
+#| label: tbl-liver-norms
+#| tbl-cap: "Normal ranges provided for liver function tests."
+liver_stats <- tribble(
+  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
+  5,     5,   30,      35,        0,    2, 60,   "m", "min",
+  40,     30,  120,      50,        20,   44, 80,   "m", "max",
+  5,     5,   30,      35,        0,    2, 60,   "f", "min",
+  40,     30,  120,      50,        20,   70, 80,    "f", "max"
+)
+liver_stats |> 
+  filter(sex == "m") |>
+  select(-sex) |>
+  select(stat, albumin, protein, bilirubin, GGT, AST, ALP, ALT) |>
+  kable(
+    row.names = FALSE,
+    col.names = c("", "albumin (g/L)", "protein (g/L)", 
+        "bilirubin (µmol/L)", "GGT (IU/L)", "AST (IU/L)", 
+        "ALP (IU/L)", "ALT (IU/L)"),
+    booktabs = TRUE
+  )
+```
+
+These measurements are also likely correlated, based on guidance like the *ratio of AST to ALT of 2:1 indicates possible alcohol abuse*.
+Although this was provided by the pharmaceutical company, correlation between these measurements for normal patients is not readily available.
+When a correlation matrix for normal patients is provided this would allow construction of the null ellipse upon which to examine new samples.
+
+```{r}
+#| eval: false
+#| echo: false
+liver_stats <- tribble(
+  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
+  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
+  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
+  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
+  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
+)
+set.seed(1158)
+norm_vc <- diag(1, 4, 4)
+norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
+for (i in 1:3)
+  for (j in (i+1):4)
+    norm_vc[j, i] <- norm_vc[i, j]
+colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
+norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
+colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
+set.seed(217)
+
+women_vc <- norm_vc
+women_vc[1,4] <- 0.5
+women_vc[4,1] <- 0.5
+women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
+                 sigma = women_vc)
+colnames(women) <- c("GGT", "AST", "ALP", "ALT")
+women <- as.data.frame(women)
+
+animate_xy(women,            
+           ellipse=norm_vc, 
+           axes = "bottomleft", 
+           half_range=5, 
+           center=FALSE)
+
+animate_xy(women, guided_anomaly_tour(anomaly_index(),
+                                    ellipse=norm_vc), 
+           ellipse=norm_vc, 
+           axes = "bottomleft", 
+           half_range=5, 
+           center=FALSE)
+
+set.seed(1208)
+pt1 <- tibble(GGT = rnorm(5, mean=1.3, sd=0.15),
+              AST = rnorm(5, mean=1, sd=0.2),
+              ALP = c(1, 1.3, 1.6, 2.8, 2.9), 
+              ALT = c(-1, -1.3, -1.6, -2.9, -3.1))
+animate_xy(pt1, guided_anomaly_tour(anomaly_index(),
+                                     ellipse=norm_vc), 
+           ellipse=norm_vc, 
+           axes = "bottomleft", 
+           half_range=5, 
+           center=FALSE,
+           edges = matrix(c(1:4, 2:5), byrow=FALSE, ncol=2),
+           obs_labels = c("45", "50", "55", "65", "70"))
+```
+
+@fig-liver illustrates two examples.
+The first is similar to the consulting project.
+A sample of liver test scores for new patients was provided in order to examine their values relative to the normal range.
+Here only four tests are used, GGT, AST, ALP, ALT.
+Plot (a) shows a projection of this sample relative to the normal range.
+Three of the patients are outside the confidence ellipse but all of the patients are located away from the mean.
+What has been typical in the past is to compute normal values based on tests of healthy young males.
+The samples provided for the project were all recorded on women.
+We see that this sample has slightly lower ALT and ALP, which is consistent with what is reported in @lib-med-liver-norms.
+
+The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70.
+The lines connect the records in tome order.
+The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:
+
+$$
+P = \left[ \begin{array}{rr}
+            0.371 & -0.128 \\
+            -0.388 & 0.732 \\
+            0.140 & 0.599 \\
+            -0.832 & -0.297 
+            \end{array} \right]
+$$
+
+From the axes representation in @fig-liver (b) of this projection, we can see that the direction that profile extends is primarily contrasting ALT (fourth row) and ALP (third row).
+This is consistent with aging, where ALP increases and ALT decreases.
+
+::: {#fig-liver layout-ncol=2}
+
+![Sample of female patients.](images/liver1.png){#fig-liver1 width=300 align="center"}
+
+![Longitudinal profile of an aging patient.](images/liver2.png){#fig-liver2 width=300 align="center"}
+
+Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient.
+Red cross indicates observation is outide the 4-D confidence ellipse.
+The female patients tend to have lower ALP and ALT than the normal range.
+As the patient aged, the level of ALP increases and ALT decreases.
+:::
 

### 2024-07-02 paper.tex
liver example added

--- 2024-06-27 paper.tex
@@
 \addtolength{\topmargin}{-1in}%
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{float}
@@
 \newcommand\gD{$2\text{-}D$}
+\newtheorem*{theorem}{Theorem}
 \makeatletter
@@
 
-The projection of an ellipsoid onto \gD{} is an ellipse, where the curve
-of the ellipse is defined through the set of points \(\mathbfit{x}\) for
-which the gradient is parallel to the projection plane. That is, the
-curve consists of \(\mathbfit{x}\) satisfying
+\begin{theorem}
+  The projection of this \pD{} ellipsoid in \gD{} has the equation
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
+\end{theorem}
+\begin{proof}
+The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
+That is, the curve
+consists of $\mathbfit{x}$ satisfying
 
-\[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
-\]
+$$\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+$$
 
-for some \(\mathbfit{s} \in \mathbb{R}^2\), where \(P\) is a
-\((p\times 2)\) orthonormal basis defining the projection. We can write
-\(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\). Making this
-substitution in the \pD{} ellipsoid equation yields
+for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
+projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+Making this substitution in the \pD{} ellipsoid equation yields
 
-\[c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
-\]
+$$c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+$$
 
-We call points in the projection that are on the curve \(\mathbfit{y}\).
+We call points in the projection that are on the curve $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for $\mathbfit{x}$ on the \pD{} curve.
+Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
+We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the equation \(c^2 = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T\).
 From this we can compute the analog equation for the projection as
 
-\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.\]
+$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
+
+as claimed.
+\end{proof}
+
 This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
@@
 
-\subsection{Medical patients}\label{medical-patients}
+\subsection{Health: liver function
+tests}\label{health-liver-function-tests}
 
-\subsection{Robust statistics}\label{robust-statistics}
+This example is motivated by a problem posed during consulting with a
+pharmaceutical company, but the data shown here is simulated, simply to
+illustrate the application. Liver function tests commonly provide
+measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase
+(GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and
+alanine aminotransferase (ALT). There are normal ranges on these
+measurements reported by \citet{lib-med-liver-norms} and listed in
+Table~\ref{tbl-liver-norms}.
 
-\subsection{Physics?}\label{physics}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0421}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2000}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}@{}}
+
+\caption{\label{tbl-liver-norms}Normal ranges provided for liver
+function tests.}
+
+\tabularnewline
+
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+albumin (g/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+protein (g/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+bilirubin (µmol/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+GGT (IU/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+AST (IU/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ALP (IU/L)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ALT (IU/L)
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+min & 35 & 60 & 0 & 2 & 5 & 30 & 5 \\
+max & 50 & 80 & 20 & 44 & 30 & 120 & 40 \\
+
+\end{longtable}
+
+These measurements are also likely correlated, based on guidance like
+the \emph{ratio of AST to ALT of 2:1 indicates possible alcohol abuse}.
+Although this was provided by the pharmaceutical company, correlation
+between these measurements for normal patients is not readily available.
+When a correlation matrix for normal patients is provided this would
+allow construction of the null ellipse upon which to examine new
+samples.
+
+Figure~\ref{fig-liver} illustrates two examples. The first is similar to
+the consulting project. A sample of liver test scores for new patients
+was provided in order to examine their values relative to the normal
+range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows
+a projection of this sample relative to the normal range. Three of the
+patients are outside the confidence ellipse but all of the patients are
+located away from the mean. What has been typical in the past is to
+compute normal values based on tests of healthy young males. The samples
+provided for the project were all recorded on women. We see that this
+sample has slightly lower ALT and ALP, which is consistent with what is
+reported in \citet{lib-med-liver-norms}.
+
+The second example shows a longitudinal record of a single patient,
+measured at ages 45, 50, 55, 65 and 70. The lines connect the records in
+tome order. The projection corresponds to the maxima from a
+projection-pursuit guided tour using the anomaly index:
+
+\[
+P = \left[ \begin{array}{rr}
+            0.371 & -0.128 \\
+            -0.388 & 0.732 \\
+            0.140 & 0.599 \\
+            -0.832 & -0.297 
+            \end{array} \right]
+\]
+
+From the axes representation in Figure~\ref{fig-liver} (b) of this
+projection, we can see that the direction that profile extends is
+primarily contrasting ALT (fourth row) and ALP (third row). This is
+consistent with aging, where ALP increases and ALT decreases.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}
+
+}
+
+\subcaption{\label{fig-liver1}Sample of female patients.}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}
+
+}
+
+\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient.}
+
+\end{minipage}%
+
+\caption{\label{fig-liver}Two projections of simulated example data
+corresponding to common liver tests: (a) sample of female patients, (b)
+longitudinal test results for a single patient. Red cross indicates
+observation is outide the 4-D confidence ellipse. The female patients
+tend to have lower ALP and ALT than the normal range. As the patient
+aged, the level of ALP increases and ALT decreases.}
+
+\end{figure}%
+
+\subsection{Robust statistics: weather
+extremes}\label{robust-statistics-weather-extremes}
 

### 2024-07-02 data/liver.R

@@ -2,14 +2,14 @@
 library(mvtnorm)
 library(tidyverse)
 library(tourr)
-library(GGally)
+#library(GGally)
 
 liver_stats <- tribble(
-  ~ALT, ~AST, ~ALP, ~Albumin, ~Bilirubin, ~GGT, ~LD, ~PT, ~sex, ~stat,
-  29,     14,   44,      3.5,        0.1,    5, 140,   0,   "m", "mn",
-  33,     20,  147,      5.5,        1.2,   40, 280, 1.1,   "m", "mx",
-  19,     10,   44,      3.5,        0.1,    5, 140,   0,   "f", "mn",
-  25,     36,  147,      5.5,        1.2,   40, 280, 1.1,   "f", "mx"
+  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
+  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
+  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
+  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
+  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
 )
 
 liver_stats |>
@@ -23,6 +23,14 @@ liver_stats |>
 # Albumin g/dL
 # Bilirubin mg/dL
 # GGT, LD U/L
+# Protein
+# https://www.racgp.org.au/getattachment/36d1c5e0-9c1d-43fc-a8a0-b323e3ed8fbe/Liver-function-tests.aspx
+# https://www.mayoclinic.org/tests-procedures/liver-function-tests/about/pac-20394595
+# https://www.thepathologycentre.org/test/liver-panel-lft-albumin-bilirubin-alt-alp-ggt-and-protein-2/
+
+# Aging women have elevated ALP
+# ALT decreases with age, for all demog
+# https://link.springer.com/referenceworkentry/10.1007/978-3-319-90761-1_49-1
 
 # Using standardised measurements
 set.seed(1257)
@@ -112,5 +120,75 @@ d <- d |>
 
 animate_xy(d[,1:5], axes = "off", col=d$set)
 animate_xy(d[,1:5], guided_tour(lda_pp(d$set)), col=d$set)
+
+# Try 2 with actual values
+liver_stats |>
+  filter(sex == "m")
+
+liver_stats |>
+  pivot_longer(ALT:PT, names_to = "test", values_to = "value") |>
[…]

### 2024-07-03 paper.qmd
note to add the other outlyingness llit
adding in citations

@@
 A linear projection provides a dimension reduction while maintaining interpretability.
-For example, a biplot (REF) shows the structure creating the maximum variance in the data, and also visualizing the projection matrix to learn which variables contribute to it.
+For example, a biplot [@Ga71;@GH96] shows the structure creating the maximum variance in the data, and also visualizing the projection matrix to learn which variables contribute to it.
 We might find clusters of outliers that were hiding in high dimensions.
 
-More generally, projection pursuit (REF) defines a quantitative criterion for the interestingness of a projection (a projection pursuit index), and searches the space of possible projections for the most interesting one to display.
+More generally, projection pursuit [@FT74;@JS87;@Hu85] defines a quantitative criterion for the interestingness of a projection (a projection pursuit index), and searches the space of possible projections for the most interesting one to display.
 We can also define sequences of interpolated linear projections to better understand a multivariate distribution.
-Animating a randomly selected interpolated sequence of linear projections is called a grand tour (REF).
+Animating a randomly selected interpolated sequence of linear projections is called a grand tour [@As85;@BA86b;@BCAH05;@CLBW06;@tours2022].
 The combination of these two approaches would then use a projection pursuit index to select interesting projections, but display them via an interpolated path to provide context.
-This is called a guided tour (@cook1995).
+This is called a guided tour [@cook1995].
 
@@
 ## Conclusion
+
+Say something about the relationship with @directional-outlyingness and Stahel-Donoho outlyingness

### 2024-07-03 paper.tex
note to add the other outlyingness llit
adding in citations

@@
 linear projection provides a dimension reduction while maintaining
-interpretability. For example, a biplot (REF) shows the structure
-creating the maximum variance in the data, and also visualizing the
-projection matrix to learn which variables contribute to it. We might
-find clusters of outliers that were hiding in high dimensions.
+interpretability. For example, a biplot \citep{Ga71, GH96} shows the
+structure creating the maximum variance in the data, and also
+visualizing the projection matrix to learn which variables contribute to
+it. We might find clusters of outliers that were hiding in high
+dimensions.
 
-More generally, projection pursuit (REF) defines a quantitative
-criterion for the interestingness of a projection (a projection pursuit
-index), and searches the space of possible projections for the most
-interesting one to display. We can also define sequences of interpolated
-linear projections to better understand a multivariate distribution.
-Animating a randomly selected interpolated sequence of linear
-projections is called a grand tour (REF). The combination of these two
-approaches would then use a projection pursuit index to select
+More generally, projection pursuit \citep{FT74, JS87, Hu85} defines a
+quantitative criterion for the interestingness of a projection (a
+projection pursuit index), and searches the space of possible
+projections for the most interesting one to display. We can also define
+sequences of interpolated linear projections to better understand a
+multivariate distribution. Animating a randomly selected interpolated
+sequence of linear projections is called a grand tour
+\citep{As85, BA86b, BCAH05, CLBW06, tours2022}. The combination of these
+two approaches would then use a projection pursuit index to select
 interesting projections, but display them via an interpolated path to
-provide context. This is called a guided tour (\citet{cook1995}).
+provide context. This is called a guided tour \citep{cook1995}.
 
@@
 
-\renewcommand\refname{Conclusion}
+\section{Conclusion}\label{conclusion}
+
+Say something about the relationship with
+\citet{directional-outlyingness} and Stahel-Donoho outlyingness
+
   \bibliography{bibliography.bib}

### 2024-07-03 paper.qmd
mapping out conclusions

@@
 
-To define a measure of an interesting projection is to maximize the **average Malahanobis distance in the projection** for a subset of points, $W$.
+To define a measure of an interesting projection is to maximize the average Mahahlanobis distance [@mahalanobis] in the projection for a subset of points, $W$.
 The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}.
@@
 
-This is implemented in the `tourr` (REF) package, where the projected ellipsoid can be drawn for each projection.
+This is implemented in the `tourr` [@tourr;@tourr-cran] package, where the projected ellipsoid can be drawn for each projection.
 The guided tour will take arguments specifying the data, and the null variance-covariance matrix.
@@
 
-Say something about the relationship with @directional-outlyingness and Stahel-Donoho outlyingness
+Say something about the relationship with @directional-outlyingness and Stahel-Donoho outlyingness.
+This method could be implemented in the way we used to show ellipses, generate points on the surface of the irregular shape, overlay the data on this and make projections.
+
+Difference from general outlier detection, see use of Mahalanobis distance in @filzmoser2018.
+
+Potential new directions.

### 2024-07-03 paper.tex
mapping out conclusions

@@
 To define a measure of an interesting projection is to maximize the
-\textbf{average Malahanobis distance in the projection} for a subset of
-points, \(W\). The set of points could be chosen in different ways, but
-the default is those that are outside the specified ellipsoid in \pD{}.
-Alternatives could be to select a set of observations with the largest
-Mahalanobis distance, manually select observations or possibly a group
-of points identified via clustering of the extremes.
+average Mahahlanobis distance \citep{mahalanobis} in the projection for
+a subset of points, \(W\). The set of points could be chosen in
+different ways, but the default is those that are outside the specified
+ellipsoid in \pD{}. Alternatives could be to select a set of
+observations with the largest Mahalanobis distance, manually select
+observations or possibly a group of points identified via clustering of
+the extremes.
 
@@
 
-This is implemented in the \texttt{tourr} (REF) package, where the
-projected ellipsoid can be drawn for each projection. The guided tour
-will take arguments specifying the data, and the null
+This is implemented in the \texttt{tourr} \citep{tourr, tourr-cran}
+package, where the projected ellipsoid can be drawn for each projection.
+The guided tour will take arguments specifying the data, and the null
 variance-covariance matrix.
@@
 Say something about the relationship with
-\citet{directional-outlyingness} and Stahel-Donoho outlyingness
+\citet{directional-outlyingness} and Stahel-Donoho outlyingness. This
+method could be implemented in the way we used to show ellipses,
+generate points on the surface of the irregular shape, overlay the data
+on this and make projections.
+
+Difference from general outlier detection, see use of Mahalanobis
+distance in \citet{filzmoser2018}.
+
+Potential new directions.
 

### 2024-07-03 paper.qmd
Starting to add robust statistics application

@@
 
+Another application would compare outlying points to a reference normal distribution that is derived using robust methods on the original data.
+To illustrate this we work on the weather data example from @filzmoser2018.
+The data contains 16 numerical measurements that are averages across the three summer months June, July and August, these are provided for 68 years (1955 - 2022), see @filzmoser2018 for more details.
+
+To estimate the underlying normal distribution the data is first centered and scaled using the median and the median absolute deviation (MAD), before applying the minimum covariance determinant (MCD) estimator @rousseeuw1985multivariate using the implementation in @robustbase.
+The MCD estimates for the mean vector and the variance-covariance matrix are then used to define the reference normal distribution.
+
+Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value, given $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger.
+This will identify $20$ out of the $68$ observations as outlying.
+Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters.
+To focus on the direction when grouping the points we first normalize observations to have length $1$ and then apply k-means clustering.
+Considering different cluster validation statistics computed from @fpc suggests that $k=4$ or $k=5$ is preferred, for simplicity we will work with $k=4$.
+
+We first apply the new anomaly index to the full dataset, such that the final projection will identify a compromise and provide a global picture of where the outlying points differ from the normal distribution.
+
+```{r}
+#| eval: false
+#| echo: false
+library(ShapleyOutlier)
+library(robustHD)
+library(tidyverse)
+library(tourr)
+library(fpc)
+
+data("WeatherVienna")
+# first part of the code is preprocessing
+# this is following the vignette from the package
+# see https://cran.r-project.org/web/packages/ShapleyOutlier/vignettes/ShapleyOutlier_examples.html
+weather_summer <- WeatherVienna %>% dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) %>%
+  drop_na() %>%
+  filter(month %in% c("JUN", "JUL", "AUG")) %>%
+  filter(year >= 1955) %>%
+  group_by(year) %>%
+  dplyr::select(-month) %>%
+  summarise(across(.cols = everything(), function(x) mean(x)))
+
+X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year))
+rownames(X) <- weather_summer$year
+#> Warning: Setting row names on a tibble is deprecated.
+X <- robStandardize(X)
+
+set.seed(1)
+MCD <- covMcd(X, alpha = 0.5, nsamp = "best")
+#> Warning in .fastmcd(x, h, nsamp, nmini, kmini, trace = as.integer(trace)): 'nsamp = "best"' allows maximally 100000 subsets;
+#> computing these subsets of size 17 out of 68
+mu <-MCD$center
+Sigma <- MCD$cov
+
+###### preprocessing done
+# this is now our own code to work with these inputs
+
+# start with finding outlying points and clustering
+size_weather <- qchisq(2*pnorm(5)-1, ncol(X))
+dist_pd <- mahalanobis(X, mu, Sigma)
+X_outside <- X[dist_pd > size_weather,]
+X_outside_scaled <- t(apply(X_outside, 1,
+                    function(x) (x)/sqrt(sum(x^2))))
+for(k in 2:6){
+  set.seed(150)
+  X_km <- kmeans(X_outside_scaled, k)
+  cs <- cluster.stats(dist(X_outside_scaled), X_km$cluster)
+  print(cs$dunn)
+}
+
+#preferred solution 5 clusters using dunn/dunn2 index
+set.seed(150)
+X_km <- kmeans(X_outside_scaled, 5)
+# making a vector that labels all points
+# use 0 to indicate point is not an outlier
+cluster_labels_all <- as_tibble(X,
+                               rownames = "year") %>%
+  left_join(as_tibble(X_km$cluster,
+                              rownames = "year")) %>%
+  mutate(cluster = if_else(is.na(value), 0, value))
+cluster_labels <- factor(cluster_labels_all$cluster,
+                         levels = c(1:5, 0))
+
+set.seed(129)
+animate_xy(X, 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellc = size_weather), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           axes="off", col = as.factor(cluster_labels),
+           palette = "Plasma", half_range = 5,
+           center = FALSE, rescale = FALSE)
+
+```
+
 ## Conclusion

### 2024-07-03 paper.tex
Starting to add robust statistics application

@@
 
+Another application would compare outlying points to a reference normal
+distribution that is derived using robust methods on the original data.
+To illustrate this we work on the weather data example from
+\citet{filzmoser2018}. The data contains 16 numerical measurements that
+are averages across the three summer months June, July and August, these
+are provided for 68 years (1955 - 2022), see \citet{filzmoser2018} for
+more details.
+
+To estimate the underlying normal distribution the data is first
+centered and scaled using the median and the median absolute deviation
+(MAD), before applying the minimum covariance determinant (MCD)
+estimator \citet{rousseeuw1985multivariate} using the implementation in
+\citet{robustbase}. The MCD estimates for the mean vector and the
+variance-covariance matrix are then used to define the reference normal
+distribution.
+
+Here we will consider points to be outlying if they are more than
+\(5 \sigma\) away from that mean value, given \(p=16\) this corresponds
+to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
+out of the \(68\) observations as outlying. Since these are outlying in
+different combinations of variables, as found in \citet{filzmoser2018},
+we will further separate the outlying points into clusters. To focus on
+the direction when grouping the points we first normalize observations
+to have length \(1\) and then apply k-means clustering. Considering
+different cluster validation statistics computed from \citet{fpc}
+suggests that \(k=4\) or \(k=5\) is preferred, for simplicity we will
+work with \(k=4\).
+
+We first apply the new anomaly index to the full dataset, such that the
+final projection will identify a compromise and provide a global picture
+of where the outlying points differ from the normal distribution.
+
 \section{Conclusion}\label{conclusion}

### 2024-07-03 paper.qmd
Adding code for running with labels
Robust stats application fixed with providing mu value

@@
 
-set.seed(129)
+set.seed(22)
 animate_xy(X, 
@@
                                ellipse=as.matrix(Sigma),
-                               ellc = size_weather), 
+                               ellc = size_weather,
+                               ellmu = mu), 
            ellipse=as.matrix(Sigma), ellc = size_weather,
+           ellmu = mu,
            axes="off", col = as.factor(cluster_labels),
-           palette = "Plasma", half_range = 5,
+           palette = "Purple-Yellow",
            center = FALSE, rescale = FALSE)

### 2024-07-03 paper.qmd
Next steps for weather data

@@
 
+# extract final view, draw with new axis display that
+# only labels longer arrows, draw without red X
+
+# cluster 4 only has two points, they are more in
+# the center when optimizing for all outliers
+# next step: check optimization for only cluster 4
+# also cluster 3 might be of interest, it appears
+# somewhat spread out in the final view, but it is
+# also the largest cluster
+
 ```

### 2024-07-03 paper.qmd
language smoothing

--- 2024-07-03 paper.qmd
@@
 
-Another application would compare outlying points to a reference normal distribution that is derived using robust methods on the original data.
-To illustrate this we work on the weather data example from @filzmoser2018.
+This example is motivated by the weather data example from @filzmoser2018.
+We illustrate using the anomaly index to  compare potential outliers with a reference normal distribution that is derived using robust methods on the original data.
 The data contains 16 numerical measurements that are averages across the three summer months June, July and August, these are provided for 68 years (1955 - 2022), see @filzmoser2018 for more details.
@@
 
-We first apply the new anomaly index to the full dataset, such that the final projection will identify a compromise and provide a global picture of where the outlying points differ from the normal distribution.
+The new anomaly index is first applied to the full dataset, such that the final projection will identify a compromise and provide a global picture of where the outlying points differ from the normal distribution.
 
@@
 
+# adding labels and running grand tour to better track
+# why there might be a missmatch
+animate_xy(X,
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           axes="off", col = as.factor(cluster_labels),
+           palette = "Plasma", half_range = 5,
+           center = FALSE, rescale = FALSE, obs_labels = rownames(X))
+
 ```

### 2024-07-03 paper.qmd
language smoothing

@@
 
-To estimate the underlying normal distribution the data is first centered and scaled using the median and the median absolute deviation (MAD), before applying the minimum covariance determinant (MCD) estimator @rousseeuw1985multivariate using the implementation in @robustbase.
+To estimate the underlying normal distribution the data is first centered and scaled using the median and the median absolute deviation (MAD), before applying the minimum covariance determinant (MCD) estimator [@rousseeuw1985multivariate] using the implementation in @robustbase.
 The MCD estimates for the mean vector and the variance-covariance matrix are then used to define the reference normal distribution.

### 2024-07-03 paper.tex
language smoothing

@@
 
-Another application would compare outlying points to a reference normal
-distribution that is derived using robust methods on the original data.
-To illustrate this we work on the weather data example from
-\citet{filzmoser2018}. The data contains 16 numerical measurements that
-are averages across the three summer months June, July and August, these
-are provided for 68 years (1955 - 2022), see \citet{filzmoser2018} for
-more details.
+This example is motivated by the weather data example from
+\citet{filzmoser2018}. We illustrate using the anomaly index to compare
+potential outliers with a reference normal distribution that is derived
+using robust methods on the original data. The data contains 16
+numerical measurements that are averages across the three summer months
+June, July and August, these are provided for 68 years (1955 - 2022),
+see \citet{filzmoser2018} for more details.
 
@@
 (MAD), before applying the minimum covariance determinant (MCD)
-estimator \citet{rousseeuw1985multivariate} using the implementation in
+estimator \citep{rousseeuw1985multivariate} using the implementation in
 \citet{robustbase}. The MCD estimates for the mean vector and the
@@
 
-We first apply the new anomaly index to the full dataset, such that the
-final projection will identify a compromise and provide a global picture
-of where the outlying points differ from the normal distribution.
+The new anomaly index is first applied to the full dataset, such that
+the final projection will identify a compromise and provide a global
+picture of where the outlying points differ from the normal
+distribution.
 

### 2024-07-04 paper.qmd
small change to title
filled in early example

@@
 ---
-title: "Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
+title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
 author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
@@
 library(kableExtra)
+library(mvtnorm)
 ```
@@
 2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. 
-
 To convert this to points on the surface of a confidence ellipsoid,
@@
 
-<PLOTS FROM USCHI's PHYSICS EXAMPLE HERE>
+For example, @fit-compare compares a new sample of patient scores against the normal range represented by an ellipse.
+
+```{r}
+#| eval: false
+#| echo: false
+liver_stats <- tribble(
+  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
+  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
+  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
+  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
+  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
+)
+set.seed(1158)
+norm_vc <- diag(1, 4, 4)
+norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
+for (i in 1:3)
+  for (j in (i+1):4)
+    norm_vc[j, i] <- norm_vc[i, j]
+colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
+norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
+colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
+norm_samp <- as.data.frame(norm_samp)
+norm_ci <- gen_vc_ellipse(norm_vc) * 2
+colnames(norm_ci) <- c("GGT", "AST", "ALP", "ALT")
+norm_ci <- as.data.frame(norm_ci)
+
+set.seed(217)
+women_vc <- norm_vc
+women_vc[1,4] <- 0.5
+women_vc[4,1] <- 0.5
+women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
+                 sigma = women_vc)
+colnames(women) <- c("GGT", "AST", "ALP", "ALT")
+women <- as.data.frame(women)
+
+norm_plus <- bind_rows(norm_samp, women)
+norm_plus <- norm_plus |>
+  mutate(type = factor(c(rep("norm", 500), rep("samp", 16))))
+animate_xy(norm_plus[,1:4], col=norm_plus$type)
+
+norm_ci_plus <- bind_rows(norm_ci, women)
+norm_ci_plus <- norm_ci_plus |>
+  mutate(type = factor(c(rep("ci", 500), rep("samp", 16))))
+animate_xy(norm_ci_plus[,1:4], col=norm_ci_plus$type)
+
+```
+
+::: {#fig-compare layout-ncol=2}
+
+![Sample of patients relative to confidence ellipse.](images/example1.png){#fig-compare1 width=300 align="center"}
+
+![Sample of patients relative to relative to simulated sample of normal patients.](images/example2.png){#fig-compare width=300 align="center"}
+
+Illustration of current procedure: (a) compare new sample with points of the surface of a $p-D$ ellipse, (b) compare new sample with a simulated sample of normal patients.
+Although both approaches are useful, the rough edges of the projected ellipse points makes it difficult to precisely assess the positions of the new sample against the normal bounds.
+:::
 
@@
 #| echo: false
-liver_stats <- tribble(
-  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
-  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
-  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
-  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
-  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
-)
-set.seed(1158)
-norm_vc <- diag(1, 4, 4)
-norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
-for (i in 1:3)
-  for (j in (i+1):4)
-    norm_vc[j, i] <- norm_vc[i, j]
-colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
-norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
-colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
-set.seed(217)
-
-women_vc <- norm_vc
-women_vc[1,4] <- 0.5
-women_vc[4,1] <- 0.5
-women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
-                 sigma = women_vc)
-colnames(women) <- c("GGT", "AST", "ALP", "ALT")
-women <- as.data.frame(women)
-
 animate_xy(women,            

### 2024-07-04 paper.tex
small change to title
filled in early example

@@
 \hypersetup{
-  pdftitle={Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
+  pdftitle={Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
   pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
@@
 
-\title{\bf Are you normal? A new projection pursuit index to assess a
+\title{\bf Is this normal? A new projection pursuit index to assess a
 sample against a multivariate null distribution}
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
@@
   Transforming each observation to have unit distance from the mean,
-  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
-\end{enumerate}
-
-To convert this to points on the surface of a confidence ellipsoid,
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
+  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\). To convert this
+  to points on the surface of a confidence ellipsoid,
 \item
@@
 
-\textless PLOTS FROM USCHI's PHYSICS EXAMPLE HERE\textgreater{}
+For example, \citet{fit-compare} compares a new sample of patient scores
+against the normal range represented by an ellipse.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/example1.png}
+
+}
+
+\subcaption{\label{fig-compare1}Sample of patients relative to
+confidence ellipse.}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/example2.png}
+
+}
+
+\subcaption{\label{fig-compare}Sample of patients relative to relative
+to simulated sample of normal patients.}
+
+\end{minipage}%
+
+\caption{\label{fig-compare}Illustration of current procedure: (a)
+compare new sample with points of the surface of a \(p-D\) ellipse, (b)
+compare new sample with a simulated sample of normal patients. Although
+both approaches are useful, the rough edges of the projected ellipse
+points makes it difficult to precisely assess the positions of the new
+sample against the normal bounds.}
+
+\end{figure}%
 

### 2024-07-04 paper.qmd
finished the early example

@@
 
-For example, @fit-compare compares a new sample of patient scores against the normal range represented by an ellipse.
+For example, @fig-compare compares a new sample of patient scores against the normal range represented by an ellipse (a) and also against a simulated sample of normal patients (b).
+While these are useful approaches, the ragged edges of the projected ellipse make it difficult to compare the new sample precisely agains the normal ranges.
 
@@
 
-![Sample of patients relative to confidence ellipse.](images/example1.png){#fig-compare1 width=300 align="center"}
+![Relative to confidence ellipse.](images/example1.png){#fig-compare1 width=250 align="center"}
 
-![Sample of patients relative to relative to simulated sample of normal patients.](images/example2.png){#fig-compare width=300 align="center"}
+![Relative to normal patients.](images/example2.png){#fig-compare width=250 align="center"}
 
-Illustration of current procedure: (a) compare new sample with points of the surface of a $p-D$ ellipse, (b) compare new sample with a simulated sample of normal patients.
+Illustration of current procedure: (a) compare new sample with points of the surface of a $p\text{-}D$ ellipse, (b) compare new sample with a simulated sample of normal patients.
 Although both approaches are useful, the rough edges of the projected ellipse points makes it difficult to precisely assess the positions of the new sample against the normal bounds.

### 2024-07-04 paper.tex
finished the early example

--- 2024-07-03 paper.tex
@@
 
-\textless PLOTS FROM USCHI's PHYSICS EXAMPLE HERE\textgreater{}
+For example, Figure~\ref{fig-compare} compares a new sample of patient
+scores against the normal range represented by an ellipse (a) and also
+against a simulated sample of normal patients (b). While these are
+useful approaches, the ragged edges of the projected ellipse make it
+difficult to compare the new sample precisely agains the normal ranges.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/example1.png}
+
+}
+
+\subcaption{\label{fig-compare1}Relative to confidence ellipse.}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=2.60417in,height=\textheight]{images/example2.png}
+
+}
+
+\subcaption{\label{fig-compare}Relative to normal patients.}
+
+\end{minipage}%
+
+\caption{\label{fig-compare}Illustration of current procedure: (a)
+compare new sample with points of the surface of a \(p\text{-}D\)
+ellipse, (b) compare new sample with a simulated sample of normal
+patients. Although both approaches are useful, the rough edges of the
+projected ellipse points makes it difficult to precisely assess the
+positions of the new sample against the normal bounds.}
+
+\end{figure}%
 

### 2024-07-04 paper.qmd
Placeholder results for weather data

--- 2024-07-03 paper.qmd
@@
                               rownames = "year")) %>%
-  mutate(cluster = if_else(is.na(value), 0, value))
+  mutate(cluster = if_else(is.na(value), 0, value)) %>%
+  arrange(cluster)
 cluster_labels <- factor(cluster_labels_all$cluster,
@@
 
+X_sorted <- cluster_labels_all %>%
+  select(-year, -cluster, -value)
+
 set.seed(22)
-animate_xy(X, 
+animate_xy(as.matrix(X_sorted), 
            guided_anomaly_tour(anomaly_index(),
@@
            ellmu = mu,
-           axes="off", col = as.factor(cluster_labels),
+           axes="bottomleft",
+           col = as.factor(cluster_labels),
            palette = "Purple-Yellow",
+           center = FALSE, rescale = FALSE)
+
+# color codes
+# "#80146E" "#7262B2" "#31A0C2" "#5ECBBB" "#B9E5B9"
+# "#F5F2D8"
+# wait for getting random starting planes for
+# better results
+set.seed(55)
+animate_xy(X_sorted[cluster_labels == 4,], 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellc = size_weather,
+                               ellmu = mu), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           ellmu = mu,
+           axes="bottomleft",
+           col = "#5ECBBB", half_range = 7,
            center = FALSE, rescale = FALSE)
@@
 
+::: {#fig-weather layout-ncol=2}
+
+![All outlying points.](images/weather1.png){#fig-weather1 width=300 align="center"}
+
+![Only cluster 4.](images/weather2.png){#fig-weather2 width=300 align="center"}
+
+Projection identified by the anomaly index.
+(a) shows the preferred projection when averaging all outlying points.
+(b) shows the preferred projection when optimizing only for cluster 4, which could not be resolved in the view in (a).
+:::
+
 ## Conclusion

### 2024-07-04 paper.tex
Placeholder results for weather data

--- 2024-07-03 paper.tex
@@
 
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/weather1.png}
+
+}
+
+\subcaption{\label{fig-weather1}All outlying points.}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/weather2.png}
+
+}
+
+\subcaption{\label{fig-weather2}Only cluster 4.}
+
+\end{minipage}%
+
+\caption{\label{fig-weather}Projection identified by the anomaly index.
+(a) shows the preferred projection when averaging all outlying points.
+(b) shows the preferred projection when optimizing only for cluster 4,
+which could not be resolved in the view in (a).}
+
+\end{figure}%
+
 \section{Conclusion}\label{conclusion}

### 2024-07-04 paper.qmd
revised figures for weather

--- 2024-07-04 paper.qmd
@@
 # see https://cran.r-project.org/web/packages/ShapleyOutlier/vignettes/ShapleyOutlier_examples.html
-weather_summer <- WeatherVienna %>% dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) %>%
-  drop_na() %>%
-  filter(month %in% c("JUN", "JUL", "AUG")) %>%
-  filter(year >= 1955) %>%
-  group_by(year) %>%
-  dplyr::select(-month) %>%
+weather_summer <- WeatherVienna |> dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) |>
+  drop_na() |>
+  filter(month %in% c("JUN", "JUL", "AUG")) |>
+  filter(year >= 1955) |>
+  group_by(year) |>
+  dplyr::select(-month) |>
   summarise(across(.cols = everything(), function(x) mean(x)))
 
-X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year))
+X <- weather_summer |> dplyr::select(-c(num_frost, num_ice, year))
 rownames(X) <- weather_summer$year
@@
 cluster_labels_all <- as_tibble(X,
-                               rownames = "year") %>%
+                               rownames = "year") |>
   left_join(as_tibble(X_km$cluster,
-                              rownames = "year")) %>%
-  mutate(cluster = if_else(is.na(value), 0, value))
+                              rownames = "year")) |>
+  mutate(cluster = if_else(is.na(value), 0, value)) |>
+  arrange(cluster)
 cluster_labels <- factor(cluster_labels_all$cluster,
@@
 
+X_sorted <- cluster_labels_all |>
+  select(-year, -cluster, -value) |>
+  rename(atmax = avg_t_max,
+         atmin = avg_t_min,
+         nsumm = num_summer,
+         nheat = num_heat,
+         sunh = sun_h,
+         nclear = num_clear,
+         ncloud = num_cloud,
+         rh = rel_hum,
+         rhmax = rel_hum_max,
+         rhmin = rel_hum_min,
+         vwind = wind_v,
+         vw60 = num_wind_v60,
+         vwmax = wind_v_max,
+         prec = precp_sum,
+         nprec = num_precp_01)
+
 set.seed(22)
-animate_xy(X, 
+animate_xy(as.matrix(X_sorted), 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=as.matrix(Sigma),
+                               ellc = size_weather,
+                               ellmu = mu), 
+           ellipse=as.matrix(Sigma), ellc = size_weather,
+           ellmu = mu, cex=2, 
+           axes="bottomleft",
+           col = as.factor(cluster_labels),
+           palette = "Purple-Yellow",
+           center = FALSE, rescale = FALSE,
+           ellmarks = FALSE, axislablong = TRUE)
+
+# color codes
+# "#80146E" "#7262B2" "#31A0C2" "#5ECBBB" "#B9E5B9"
+# "#F5F2D8"
+# wait for getting random starting planes for
+# better results
+set.seed(55)
+animate_xy(X_sorted[cluster_labels == 4,], 
            guided_anomaly_tour(anomaly_index(),
@@
            ellmu = mu,
-           axes="off", col = as.factor(cluster_labels),
-           palette = "Purple-Yellow",
-           center = FALSE, rescale = FALSE)
+           axes="bottomleft",
+           col = "#5ECBBB", half_range = 7, cex = 2,
+           center = FALSE, rescale = FALSE,
+           ellmarks = FALSE, axislablong = TRUE)
 

### 2024-07-04 paper.qmd
tweaking weather example

@@
 We illustrate using the anomaly index to  compare potential outliers with a reference normal distribution that is derived using robust methods on the original data.
-The data contains 16 numerical measurements that are averages across the three summer months June, July and August, these are provided for 68 years (1955 - 2022), see @filzmoser2018 for more details.
+The data contains 16 numerical measurements that are averages across the three summer months June, July and August, reported for 68 years (1955 - 2022) (see @filzmoser2018 for more details).
 
@@
 
-Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value, given $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger.
+Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value.
+With $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger.
 This will identify $20$ out of the $68$ observations as outlying.
-Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters.
-To focus on the direction when grouping the points we first normalize observations to have length $1$ and then apply k-means clustering.
-Considering different cluster validation statistics computed from @fpc suggests that $k=4$ or $k=5$ is preferred, for simplicity we will work with $k=4$.
+Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters based on  similarity in direction.
+The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance.
+The cluster validation statistics computed using @fpc suggest that $k=4$ or $k=5$ is reasonable, so for simplicity we will work with $k=4$.
 
-The new anomaly index is first applied to the full dataset, such that the final projection will identify a compromise and provide a global picture of where the outlying points differ from the normal distribution.
+The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions.
+It provides a global picture showing where the outlying points differ from the normal distribution.
 
@@
 
-![All outlying points.](images/weather1.png){#fig-weather1 width=300 align="center"}
+![All outlying points clustered](images/weather1.png){#fig-weather1 width=300 align="center"}
 
-![Only cluster 4.](images/weather2.png){#fig-weather2 width=300 align="center"}
+![Only cluster 4](images/weather2.png){#fig-weather2 width=300 align="center"}
 
-Projection identified by the anomaly index.
-(a) shows the preferred projection when averaging all outlying points.
-(b) shows the preferred projection when optimizing only for cluster 4, which could not be resolved in the view in (a).
+Examining weather anomalies using projections identified by the anomaly index: (a) computed on all outlying points, (b) computed only for cluster 4.
+In (a) the outlying points are somewhat orthogonal to the normal ellipse, which is primarily using the temperature min (`atmn`) and max (`atmx`) variables.
+In (b) the two observations are primarily extreme in the minimum temperature (`atmn`) variable, and have lower than normal minima.
 :::
+
+XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE
 

### 2024-07-04 paper.tex
tweaking weather example

--- 2024-07-04 paper.tex
@@
 numerical measurements that are averages across the three summer months
-June, July and August, these are provided for 68 years (1955 - 2022),
-see \citet{filzmoser2018} for more details.
+June, July and August, reported for 68 years (1955 - 2022) (see
+\citet{filzmoser2018} for more details).
 
@@
 Here we will consider points to be outlying if they are more than
-\(5 \sigma\) away from that mean value, given \(p=16\) this corresponds
+\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
 to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
@@
 different combinations of variables, as found in \citet{filzmoser2018},
-we will further separate the outlying points into clusters. To focus on
-the direction when grouping the points we first normalize observations
-to have length \(1\) and then apply k-means clustering. Considering
-different cluster validation statistics computed from \citet{fpc}
-suggests that \(k=4\) or \(k=5\) is preferred, for simplicity we will
-work with \(k=4\).
+we will further separate the outlying points into clusters based on
+similarity in direction. The similarity is computed by first normalizing
+observations to have length \(1\) and then apply k-means clustering with
+Euclidean distance. The cluster validation statistics computed using
+\citet{fpc} suggest that \(k=4\) or \(k=5\) is reasonable, so for
+simplicity we will work with \(k=4\).
 
 The new anomaly index is first applied to the full dataset, such that
-the final projection will identify a compromise and provide a global
-picture of where the outlying points differ from the normal
-distribution.
+the final projection will be affected by averaging distance of points in
+many directions. It provides a global picture showing where the outlying
+points differ from the normal distribution.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/weather1.png}
+
+}
+
+\subcaption{\label{fig-weather1}All outlying points clustered}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/weather2.png}
+
+}
+
+\subcaption{\label{fig-weather2}Only cluster 4}
+
+\end{minipage}%
+
+\caption{\label{fig-weather}Examining weather anomalies using
+projections identified by the anomaly index: (a) computed on all
+outlying points, (b) computed only for cluster 4. In (a) the outlying
+points are somewhat orthogonal to the normal ellipse, which is primarily
+using the temperature min (\texttt{atmn}) and max (\texttt{atmx})
+variables. In (b) the two observations are primarily extreme in the
+minimum temperature (\texttt{atmn}) variable, and have lower than normal
+minima.}
+
+\end{figure}%
+
+XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE
 

### 2024-07-05 paper.qmd
abstract and affiliation

@@
 title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
-author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
+author: 
+  - name: Annalisa Calvi
+    affiliations:
+      name: Monash University
+      department: School of Mathematics
+  - name: Ursula Laa
+    affiliations:
+      name: University of Natural Resources and Life Sciences, Vienna
+      department: Institute of Statistics
+    email: ursula.laa@boku.ac.at
+    ORCID: 0000-0002-0249-6439
+  - name: Dianne Cook
+    affiliations:
+      name: Monash University
+      department: Department of Econometrics and Business Statistics
+    email: dicook@monash.edu
+    ORCID: 0000-0002-3813-7155
 format:
@@
     keep-tex: true    
+abstract: |
+  Many data problems contain some reference or normal conditions, upon which to compare newly collected data.
+This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms.
+The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution.
+The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation.
+It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference.
+A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically.
+The methods are implemented in the R package, `tourr`.
+keywords:
+  - exploratory data analysis
+  - statistical graphics
+  - dimension reduction
+  - high-dimensional visualisation
+  - climate change
+  - clinical trials
 bibliography: bibliography.bib  
@@
 
-## Projecting an ellipsoid
+### Projecting an ellipsoid
 
@@
 
-![Random projection.](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}
+![Random projection](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}
 
-![Optimal projection.](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}
+![Optimal projection](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}
 
@@
 
-![Sample of female patients.](images/liver1.png){#fig-liver1 width=300 align="center"}
+![Sample of female patients](images/liver1.png){#fig-liver1 width=300 align="center"}
 
-![Longitudinal profile of an aging patient.](images/liver2.png){#fig-liver2 width=300 align="center"}
+![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=300 align="center"}
 

### 2024-07-05 paper.tex
abstract and affiliation

@@
   pdftitle={Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
-  pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
+  pdfauthor={Annalisa Calvi; Ursula Laa; Dianne Cook},
+  pdfkeywords={exploratory data analysis, statistical
+graphics, dimension reduction, high-dimensional visualisation, climate
+change, clinical trials},
   colorlinks=true,
@@
 \author{
-Annalisa Calvi, Ursula Laa, Dianne Cook\\
+Annalisa Calvi\\
+School of Mathematics, Monash University\\
+and\\Ursula Laa\\
+Institute of Statistics, University of Natural Resources and Life
+Sciences, Vienna\\
+and\\Dianne Cook\\
+Department of Econometrics and Business Statistics, Monash University\\
 }
@@
 \begin{abstract}
+Many data problems contain some reference or normal conditions, upon
+which to compare newly collected data. This scenario occurs in data
+collected as part of clinical trials to detect adverse events, and or
+for measuring climate change against historical norms. The data is
+typically multivariate, and often the normal ranges are specified by a
+multivariate normal distribution. The work presented in this paper
+develops methods to compare the new sample against the reference
+distribution with high-dimensional visualisation. It uses a projection
+pursuit guided tour to produce a sequence of low-dimensional projections
+steered towards those where the new sample is most different from the
+reference. A new projection pursuit index is defined, and the drawing of
+the projected ellipse is computed analytically. The methods are
+implemented in the R package, \texttt{tourr}.
+\end{abstract}
 
-\end{abstract}
+\noindent%
+{\it Keywords:} exploratory data analysis, statistical
+graphics, dimension reduction, high-dimensional visualisation, climate
+change, clinical trials
+\vfill
 
@@
 
-\section{Projecting an ellipsoid}\label{projecting-an-ellipsoid}
+\subsection{Projecting an ellipsoid}\label{projecting-an-ellipsoid}
 
@@
 
-\subcaption{\label{fig-anomaly1}Random projection.}
+\subcaption{\label{fig-anomaly1}Random projection}
 
@@
 
-\subcaption{\label{fig-anomaly2}Optimal projection.}
+\subcaption{\label{fig-anomaly2}Optimal projection}
 
@@
 
-\subcaption{\label{fig-liver1}Sample of female patients.}
+\subcaption{\label{fig-liver1}Sample of female patients}
 
@@
 
-\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient.}
+\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient}
 

### 2024-07-05 paper.qmd
finished conclusion

@@
   \newcommand\gD{$2\text{-}D$}
+  \newcommand\bD{$1\text{-}D$}
   \newtheorem*{theorem}{Theorem}
@@
 
-Say something about the relationship with @directional-outlyingness and Stahel-Donoho outlyingness.
-This method could be implemented in the way we used to show ellipses, generate points on the surface of the irregular shape, overlay the data on this and make projections.
+This paper has provided a new projection pursuit index for comparing a sample of data against a multivariate normal reference distribution.
+It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corrsponding to the normal reference distribution.
+These combined provide new ways to visualise multivariate data, for this particular scenario.
 
-Difference from general outlier detection, see use of Mahalanobis distance in @filzmoser2018.
+The work is related to outlier detection methods [see, @filzmoser2018, for example], particularly those that use robust statistics, as illustrated in the weather example.
+Methods discussed in @directional-outlyingness and @donoho1982 describe detecting outlyingness using projections of the data.
+When these projections are collected and used as a large set, they define a high-dimensional confidence region of an indeterminate shape.
+There is no analytical definition.
+However, we could explore the data relative to the high-dimensional shapes using the same approach described in @sec-background.
+You would generate points on the surface of the irregular shape, overlay the data on this to visualise with a tour.
 
-Potential new directions.
+The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, leng themselves to new potential \bD{} projection pursuit indexes.
+These would integrate nicely into a \bD{} projection pursuit guided tour to interactively visualise the potential outliers relative to the rest of the sample.
+
+Another potential direction of this work is with model-based clustering using Gaussian mixtures [@FR02], where ellipses form the basis of the model.
+Currently, the `mclust` software [@mclust] has the capacity to show \gD{} ellipses for two variables, or axis parallel \gD{} ellipses for \pD{}.
+The equations developed here for drawing the \gD{} projection of a \pD{} ellipse would provide more versatile visualisation for examining the model-based clustering fits.

### 2024-07-05 paper.tex
finished conclusion

@@
 \newcommand\gD{$2\text{-}D$}
+\newcommand\bD{$1\text{-}D$}
 \newtheorem*{theorem}{Theorem}
@@
 
-Say something about the relationship with
-\citet{directional-outlyingness} and Stahel-Donoho outlyingness. This
-method could be implemented in the way we used to show ellipses,
-generate points on the surface of the irregular shape, overlay the data
-on this and make projections.
+This paper has provided a new projection pursuit index for comparing a
+sample of data against a multivariate normal reference distribution. It
+has also provided an analytical result for drawing \gD{} projections of
+the \pD{} ellipse corrsponding to the normal reference distribution.
+These combined provide new ways to visualise multivariate data, for this
+particular scenario.
 
-Difference from general outlier detection, see use of Mahalanobis
-distance in \citet{filzmoser2018}.
+The work is related to outlier detection methods \citep[see,][for
+example]{filzmoser2018}, particularly those that use robust statistics,
+as illustrated in the weather example. Methods discussed in
+\citet{directional-outlyingness} and \citet{donoho1982} describe
+detecting outlyingness using projections of the data. When these
+projections are collected and used as a large set, they define a
+high-dimensional confidence region of an indeterminate shape. There is
+no analytical definition. However, we could explore the data relative to
+the high-dimensional shapes using the same approach described in
+Section~\ref{sec-background}. You would generate points on the surface
+of the irregular shape, overlay the data on this to visualise with a
+tour.
 
-Potential new directions.
+The methods on directional outlyingness or Stahel-Donoho outlyingness,
+described above, leng themselves to new potential \bD{} projection
+pursuit indexes. These would integrate nicely into a \bD{} projection
+pursuit guided tour to interactively visualise the potential outliers
+relative to the rest of the sample.
+
+Another potential direction of this work is with model-based clustering
+using Gaussian mixtures \citep{FR02}, where ellipses form the basis of
+the model. Currently, the \texttt{mclust} software \citep{mclust} has
+the capacity to show \gD{} ellipses for two variables, or axis parallel
+\gD{} ellipses for \pD{}. The equations developed here for drawing the
+\gD{} projection of a \pD{} ellipse would provide more versatile
+visualisation for examining the model-based clustering fits.
 

### 2024-07-05 paper.qmd
Comparison with Shapley-Outlier result added.

@@
 #| eval: false
-#| echo: true
-library(tourr)
-library(mulgar)
+#| echo: false
 set.seed(929)
@@
                            vc = vc_samp))
+```
+
+```{r}
+#| eval: false
+#| echo: true
+library(tourr)
 animate_xy(samp, guided_anomaly_tour(anomaly_index(),
@@
 The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance.
-The cluster validation statistics computed using @fpc suggest that $k=4$ or $k=5$ is reasonable, so for simplicity we will work with $k=4$.
+We use the Dunn index @dunn, computed using @fpc, to select the preferred number of clusters, $k=5$.
 
-The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions.
+The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see Fig.
+@fig-weather (a).
 It provides a global picture showing where the outlying points differ from the normal distribution.
+We can see that the contrast between maximum (atmx) and minimum (atmn) temperature is primarily used in the direction where points are most outlying.
+In the orthogonal direction the variables precipitation (prc) and number of days with maximum temperature above 25°C (nsm) can be used to separate the clusters 2, 3 and 5.
+With this solution the two points in cluster 4 cannot be resolved, they fall inside the cluster.
+To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig.
+@fig-weather (b).
+From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature.
+We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected cluster.
 
@@
 
-# extract final view, draw with new axis display that
-# only labels longer arrows, draw without red X
-
-# cluster 4 only has two points, they are more in
-# the center when optimizing for all outliers
-# next step: check optimization for only cluster 4
-# also cluster 3 might be of interest, it appears
-# somewhat spread out in the final view, but it is
-# also the largest cluster
-
 ```
@@
 
-XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE
+To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @filzmoser2018 in @fig-shapley.
+On top we can see the cluster assignments for all points that we have identified as outliers using the $p$-dimensional Mahalanobis distance.
+
+For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @filzmoser2018 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables.
+Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temparature above 25°C, and this contrast is also captured in the projection.
+
+When looking at the overall picture (Fig.
+@fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions.
+The direction of cluster 1 is where the average maximum temperature is unusually higher compared to the minimum temperature, while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature.
+This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum and the maximum temperature.
+
+```{r}
+#| eval: false
+#| echo: false
+library(ShapleyOutlier)
+Sigma_inv <- solve(MCD$cov)
+# following the package vignette to calculate SCD
+weather_SCD <- SCD(x = X, mu, Sigma, Sigma_inv, step_size = 0.1, min_deviation = 0.2)
+weather_SCD_rescaled <- weather_SCD
+weather_SCD_rescaled$x_original <- t(apply(weather_SCD_rescaled$x_original,1, function(x) x*attr(X, "scale") + attr(X, "center")))
+weather_SCD_rescaled$x <- t(apply(weather_SCD_rescaled$x,1, function(x) x*attr(X, "scale") + attr(X, "center")))
+
+# when drawing I want to add clustering information
+labels <- data.frame(
+  x = c(seq(1960, 2020, length.out = 7),
+        as.numeric(names(X_km$cluster))),
+  label = c(seq(1960, 2020, length.out = 7),
+        paste0("\n",X_km$cluster))
+)
+cluster_df <- as_tibble(X_km$cluster, rownames = "year") %>%
+  add_column(variable = "num_precp_01")
+p <- plot(weather_SCD_rescaled, type = "cell", n_digits = 0, continuous_rowname = 10, rotate_x = FALSE) +
+  geom_text(data = cluster_df,
+            mapping = aes(x = as.numeric(year),
+                               y = variable, label = value,
+                          color = as.factor(value)),
+            size = 2, fontface = "bold",
+            inherit.aes = F) +
+  scale_color_manual(values = c("#80146E", "#7262B2", "#31A0C2", "#5ECBBB", "#B9E5B9")) +
+  theme(legend.position = "none",
+        aspect.ratio = 0.5,
+        text = element_text(size = 10)) +
+  scale_x_continuous(breaks = seq(1960, 2020, 5),
+                     labels = c(as.character(seq(60, 95, 5)),
+                                "00", "05",
+                                as.character(seq(10, 20, 5))))
+ggsave("images/weather3.png", p, "png", bg = "white",
+       width = 6, height = 3)
+```
+
+::: {#fig-shapley}
+
+![](images/weather3.png){#fig-weather3 width=600 align="center"}
+
+Reproduction of the results from @filzmoser2018, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm.
+Red cells indicate values higher than expected, blue cells values lower than expected.
+For years that have been identified as outliers in our approach we show the cluster assignment at the top.
+:::
 

### 2024-07-05 paper.tex
Comparison with Shapley-Outlier result added.

@@
 \FunctionTok{library}\NormalTok{(tourr)}
-\FunctionTok{library}\NormalTok{(mulgar)}
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{929}\NormalTok{)}
-\NormalTok{vc\_null }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
-\FunctionTok{diag}\NormalTok{(vc\_null) }\OtherTok{\textless{}{-}} \DecValTok{1}
-\NormalTok{m\_null }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{)}
-\NormalTok{vc\_samp }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
-\FunctionTok{diag}\NormalTok{(vc\_samp) }\OtherTok{\textless{}{-}} \DecValTok{1}
-\NormalTok{vc\_samp[}\DecValTok{4}\NormalTok{,}\DecValTok{5}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.47}
-\NormalTok{vc\_samp[}\DecValTok{5}\NormalTok{,}\DecValTok{4}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.56}
-\NormalTok{vc\_samp }\OtherTok{\textless{}{-}}\NormalTok{ vc\_samp}\SpecialCharTok{*}\FloatTok{0.1}
-\NormalTok{m\_samp }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{1.9}\NormalTok{, }\FloatTok{2.3}\NormalTok{)}
-\NormalTok{samp }\OtherTok{\textless{}{-}} \FunctionTok{as.data.frame}\NormalTok{(}\FunctionTok{rmvn}\NormalTok{(}\DecValTok{6}\NormalTok{, }
-                           \AttributeTok{mn =}\NormalTok{ m\_samp, }
-                           \AttributeTok{vc =}\NormalTok{ vc\_samp))}
 \FunctionTok{animate\_xy}\NormalTok{(samp, }\FunctionTok{guided\_anomaly\_tour}\NormalTok{(}\FunctionTok{anomaly\_index}\NormalTok{(),}
@@
 observations to have length \(1\) and then apply k-means clustering with
-Euclidean distance. The cluster validation statistics computed using
-\citet{fpc} suggest that \(k=4\) or \(k=5\) is reasonable, so for
-simplicity we will work with \(k=4\).
+Euclidean distance. We use the Dunn index \citet{dunn}, computed using
+\citet{fpc}, to select the preferred number of clusters, \(k=5\).
 
@@
 the final projection will be affected by averaging distance of points in
-many directions. It provides a global picture showing where the outlying
-points differ from the normal distribution.
+many directions, see Fig. Figure~\ref{fig-weather} (a). It provides a
+global picture showing where the outlying points differ from the normal
+distribution. We can see that the contrast between maximum (atmx) and
+minimum (atmn) temperature is primarily used in the direction where
+points are most outlying. In the orthogonal direction the variables
+precipitation (prc) and number of days with maximum temperature above
+25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this
+solution the two points in cluster 4 cannot be resolved, they fall
+inside the cluster. To better understand what is different for those two
+years we separately run the anomaly index for this subset, see Fig.
+Figure~\ref{fig-weather} (b). From the result we see that the maximum
+temperature is not relevant for this cluster, but that they have low
+values for the minimum temperature. We also see that there is still some
+averaging, and the preferred solution is not placing either point on the
+outside of the projected cluster.
 
@@
 
-XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE
+To compare our results to what is obtained by outlier detection methods,
+we reproduce one of the results from \citet{filzmoser2018} in
+Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
+all points that we have identified as outliers using the
+\(p\)-dimensional Mahalanobis distance.
+
+For example we see that cluster 4 contains the years 1962 and 1996, and
+for these two years the result from \citet{filzmoser2018} also indicates
+unusually low values for the minimum teperature, while there is
+disagreement between outlyingness in other variables. Note that compared
+to the years 1965 and 1966 they have unusually low minimum temperature
+while having average values for the number of days with maximum
+temparature above 25°C, and this contrast is also captured in the
+projection.
+
+When looking at the overall picture (Fig. Figure~\ref{fig-weather} (a)),
+we can see that the linear projection is showing the difference between
+minimum and maximum temperature, which is where clusters are outlying in
+different directions. The direction of cluster 1 is where the average
+maximum temperature is unusually higher compared to the minimum
+temperature, while clusters 3 corresponds to years where the minimum
+temperature is unusually close to the maximum temperature. This combined
+information cannot be resolved with the cell detection algorithm, but we
+see that for example the most recent years (cluster 3) were tagged for
+having unusually high values for both the minimum and the maximum
+temperature.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=6.25in,height=\textheight]{images/weather3.png}
+
+}
+
+\caption{\label{fig-shapley}Reproduction of the results from
+\citet{filzmoser2018}, showing which cells have been identified as
+outlying by their Shapley Cell Detector algorithm. Red cells indicate
+values higher than expected, blue cells values lower than expected. For
+years that have been identified as outliers in our approach we show the
+cluster assignment at the top.}
+
+\end{figure}%
 

### 2024-07-06 paper.qmd
added orcid id

@@
       department: School of Mathematics
+    email: annalisa.calvi@monash.edu
+    ORCID: 0009-0000-2977-0074
   - name: Ursula Laa

### 2024-07-06 paper.qmd
some very minor fixes, feel free to undo

@@
 The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70.
-The lines connect the records in tome order.
+The lines connect the records in time order.
 The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:
@@
 The female patients tend to have lower ALP and ALT than the normal range.
-As the patient aged, the level of ALP increases and ALT decreases.
+As the patient ages, the level of ALP increases and ALT decreases.
 :::
@@
 In the orthogonal direction the variables precipitation (prc) and number of days with maximum temperature above 25°C (nsm) can be used to separate the clusters 2, 3 and 5.
-With this solution the two points in cluster 4 cannot be resolved, they fall inside the cluster.
+With this solution the two points in cluster 4 cannot be resolved, they fall inside the ellipse.
 To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig.
@@
 From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature.
-We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected cluster.
+We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected ellipse.
 
@@
 
-The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, leng themselves to new potential \bD{} projection pursuit indexes.
+The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, lend themselves to new potential \bD{} projection pursuit indexes.
 These would integrate nicely into a \bD{} projection pursuit guided tour to interactively visualise the potential outliers relative to the rest of the sample.

### 2024-07-07 paper.qmd
acknowledgements

@@
 The equations developed here for drawing the \gD{} projection of a \pD{} ellipse would provide more versatile visualisation for examining the model-based clustering fits.
+
+## Acknowledgements {-}
+
+The work of first author Calvi was supported by the ResearchFirst undergraduate research program at Monash University. German Valencia advised on the project, in relation to physics applications.
+
+The full paper, code and data to reproduce the work, are publicly available at https://github.com/uschiLaa/anomaly_ppi.

### 2024-07-07 paper.tex
acknowledgements

@@
 measured at ages 45, 50, 55, 65 and 70. The lines connect the records in
-tome order. The projection corresponds to the maxima from a
+time order. The projection corresponds to the maxima from a
 projection-pursuit guided tour using the anomaly index:
@@
 tend to have lower ALP and ALT than the normal range. As the patient
-aged, the level of ALP increases and ALT decreases.}
+ages, the level of ALP increases and ALT decreases.}
 
@@
 solution the two points in cluster 4 cannot be resolved, they fall
-inside the cluster. To better understand what is different for those two
+inside the ellipse. To better understand what is different for those two
 years we separately run the anomaly index for this subset, see Fig.
@@
 averaging, and the preferred solution is not placing either point on the
-outside of the projected cluster.
+outside of the projected ellipse.
 
@@
 The methods on directional outlyingness or Stahel-Donoho outlyingness,
-described above, leng themselves to new potential \bD{} projection
+described above, lend themselves to new potential \bD{} projection
 pursuit indexes. These would integrate nicely into a \bD{} projection
@@
 
+\section*{Acknowledgements}\label{acknowledgements}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+The work of first author Calvi was supported by the ResearchFirst
+undergraduate research program at Monash University. German Valencia
+advised on the project, in relation to physics applications.
+
+The full paper, code and data to reproduce the work, are publicly
+available at https://github.com/uschiLaa/anomaly\_ppi.
+
   \bibliography{bibliography.bib}

### 2024-07-16 paper.qmd
Small fixes

@@
 #| fig-height: 2
-#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1.
+#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1 (b).
 A confidence ellipsoid is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipsoid (d)."
@@
 Although this is flexible, this does not make it easy to guide the tour towards the directions (projections) where the samples are most different from the normal.
-What would be desirable is to analytically define the confidence ellipsoid, compute flag observations that are outside, steer the tour to projections that reveal the extent of the difference.
+What would be desirable is to analytically define the confidence ellipsoid, flag observations that are outside, steer the tour to projections that reveal the extent of the difference.
 And also display the projected ellipsoid as a geometric shape rather than a sample of points.
@@
 The lines connect the records in time order.
-The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:
+The projection corresponds to the maximum from a projection-pursuit guided tour using the anomaly index:
 
@@
 
-The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see Fig.
-@fig-weather (a).
+The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see @fig-weather (a).
 It provides a global picture showing where the outlying points differ from the normal distribution.
@@
 With this solution the two points in cluster 4 cannot be resolved, they fall inside the ellipse.
-To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig.
-@fig-weather (b).
+To better understand what is different for those two years we separately run the anomaly index for this subset, see @fig-weather (b).
 From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature.
@@
 For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @filzmoser2018 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables.
-Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temparature above 25°C, and this contrast is also captured in the projection.
+Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.
 
-When looking at the overall picture (Fig.
-@fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions.
+When looking at the overall picture (@fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions.
 The direction of cluster 1 is where the average maximum temperature is unusually higher compared to the minimum temperature, while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature.

### 2024-07-16 paper.tex
Small fixes

@@
 sampling from a multivariate normal (a) and transforming each
-observation to have length equal to 1. A confidence ellipsoid is
+observation to have length equal to 1 (b). A confidence ellipsoid is
 generated by transforming the sphere relative to a specified
@@
 different from the normal. What would be desirable is to analytically
-define the confidence ellipsoid, compute flag observations that are
-outside, steer the tour to projections that reveal the extent of the
-difference. And also display the projected ellipsoid as a geometric
-shape rather than a sample of points. These are the procedures that are
-described in the next section.
+define the confidence ellipsoid, flag observations that are outside,
+steer the tour to projections that reveal the extent of the difference.
+And also display the projected ellipsoid as a geometric shape rather
+than a sample of points. These are the procedures that are described in
+the next section.
 
@@
 measured at ages 45, 50, 55, 65 and 70. The lines connect the records in
-time order. The projection corresponds to the maxima from a
+time order. The projection corresponds to the maximum from a
 projection-pursuit guided tour using the anomaly index:
@@
 the final projection will be affected by averaging distance of points in
-many directions, see Fig. Figure~\ref{fig-weather} (a). It provides a
-global picture showing where the outlying points differ from the normal
+many directions, see Figure~\ref{fig-weather} (a). It provides a global
+picture showing where the outlying points differ from the normal
 distribution. We can see that the contrast between maximum (atmx) and
@@
 inside the ellipse. To better understand what is different for those two
-years we separately run the anomaly index for this subset, see Fig.
+years we separately run the anomaly index for this subset, see
 Figure~\ref{fig-weather} (b). From the result we see that the maximum
@@
 while having average values for the number of days with maximum
-temparature above 25°C, and this contrast is also captured in the
+temperature above 25°C, and this contrast is also captured in the
 projection.
 
-When looking at the overall picture (Fig. Figure~\ref{fig-weather} (a)),
-we can see that the linear projection is showing the difference between
+When looking at the overall picture (Figure~\ref{fig-weather} (a)), we
+can see that the linear projection is showing the difference between
 minimum and maximum temperature, which is where clusters are outlying in

### 2024-07-19 paper.qmd
fix typo

@@
 
-<IMAGES OF PROJECTED ELLIPSE>
-
 ### Index specification
 
-To define a measure of an interesting projection is to maximize the average Mahahlanobis distance [@mahalanobis] in the projection for a subset of points, $W$.
+To define a measure of an interesting projection is to maximize the average Mahalanobis distance [@mahalanobis] in the projection for a subset of points, $W$.
 The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}.

### 2024-07-19 paper.tex
fix typo

@@
 To define a measure of an interesting projection is to maximize the
-average Mahahlanobis distance \citep{mahalanobis} in the projection for
-a subset of points, \(W\). The set of points could be chosen in
-different ways, but the default is those that are outside the specified
-ellipsoid in \pD{}. Alternatives could be to select a set of
-observations with the largest Mahalanobis distance, manually select
-observations or possibly a group of points identified via clustering of
-the extremes.
+average Mahalanobis distance \citep{mahalanobis} in the projection for a
+subset of points, \(W\). The set of points could be chosen in different
+ways, but the default is those that are outside the specified ellipsoid
+in \pD{}. Alternatives could be to select a set of observations with the
+largest Mahalanobis distance, manually select observations or possibly a
+group of points identified via clustering of the extremes.
 

### 2024-07-24 paper.qmd
Formatting changed to tandf template

--- 2024-07-16 paper.qmd
@@
 format:
-  jasa-pdf:
+  tandf-pdf:
     keep-tex: true  
+    linestretch: 2
+    fontsize: 12pt
 abstract: |

### 2024-07-24 paper.tex
Formatting changed to tandf template

@@
 \documentclass[
-  12pt]{article}
+  12pt,
+]{interact}
 
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-1in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.7in}%
-\addtolength{\topmargin}{-1in}%
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
+\usepackage{orcidlink}
 \usepackage{amsmath}
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{agsm}
 \usepackage{bookmark}
@@
 
+\title{Is this normal? A new projection pursuit index to assess a sample
+against a multivariate null distribution}
+\author{Annalisa Calvi$\textsuperscript{1}$, Ursula
+Laa$\textsuperscript{2}$, Dianne Cook$\textsuperscript{3}$}
+
+\thanks{CONTACT: Annalisa
+Calvi. Email: \href{mailto:annalisa.calvi@monash.edu}{\nolinkurl{annalisa.calvi@monash.edu}}. Ursula
+Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}. Dianne
+Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. }
 \begin{document}
-
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{\bf Is this normal? A new projection pursuit index to assess a
-sample against a multivariate null distribution}
-\author{
-Annalisa Calvi\\
-School of Mathematics, Monash University\\
-and\\Ursula Laa\\
-Institute of Statistics, University of Natural Resources and Life
-Sciences, Vienna\\
-and\\Dianne Cook\\
-Department of Econometrics and Business Statistics, Monash University\\
-}
+\captionsetup{labelsep=space}
 \maketitle
-
-\bigskip
-\bigskip
+\textsuperscript{1} School of Mathematics, Monash
+University,  \\ \textsuperscript{2} Institute of Statistics, University
+of Natural Resources and Life Sciences,
+Vienna,  \\ \textsuperscript{3} Department of Econometrics and Business
+Statistics, Monash University,  
 \begin{abstract}
@@
 \end{abstract}
+\begin{keywords}
+\def\sep{;\ }
+exploratory data analysis\sep statistical graphics\sep dimension
+reduction\sep high-dimensional visualisation\sep climate change\sep 
+clinical trials
+\end{keywords}
 
-\noindent%
-{\it Keywords:} exploratory data analysis, statistical
-graphics, dimension reduction, high-dimensional visualisation, climate
-change, clinical trials
-\vfill
-
-\newpage
-\spacingset{1.9} % DON'T change the spacing!
-
+\setstretch{2}
 \section{Introduction}\label{introduction}
@@
 linear projection provides a dimension reduction while maintaining
-interpretability. For example, a biplot \citep{Ga71, GH96} shows the
-structure creating the maximum variance in the data, and also
-visualizing the projection matrix to learn which variables contribute to
-it. We might find clusters of outliers that were hiding in high
-dimensions.
+interpretability. For example, a biplot (Gabriel 1971; Gower and Hand
+1996) shows the structure creating the maximum variance in the data, and
+also visualizing the projection matrix to learn which variables
+contribute to it. We might find clusters of outliers that were hiding in
+high dimensions.
 
-More generally, projection pursuit \citep{FT74, JS87, Hu85} defines a
-quantitative criterion for the interestingness of a projection (a
-projection pursuit index), and searches the space of possible
-projections for the most interesting one to display. We can also define
-sequences of interpolated linear projections to better understand a
-multivariate distribution. Animating a randomly selected interpolated
-sequence of linear projections is called a grand tour
-\citep{As85, BA86b, BCAH05, CLBW06, tours2022}. The combination of these
+More generally, projection pursuit (Friedman and Tukey 1974; Jones and
+Sibson 1987; Huber 1985) defines a quantitative criterion for the
+interestingness of a projection (a projection pursuit index), and
+searches the space of possible projections for the most interesting one
+to display. We can also define sequences of interpolated linear
+projections to better understand a multivariate distribution. Animating
+a randomly selected interpolated sequence of linear projections is
+called a grand tour (Asimov 1985; Buja and Asimov 1986; Buja et al.
+2005; D. Cook et al. 2006; Lee et al. 2022). The combination of these
 two approaches would then use a projection pursuit index to select
 interesting projections, but display them via an interpolated path to
-provide context. This is called a guided tour \citep{cook1995}.
+provide context. This is called a guided tour (Dianne Cook et al. 1995).
 
@@
 To define a measure of an interesting projection is to maximize the
-average Mahalanobis distance \citep{mahalanobis} in the projection for a
+average Mahahlanobis distance (Mahalanobis 1936) in the projection for a
 subset of points, \(W\). The set of points could be chosen in different
@@
 
-This is implemented in the \texttt{tourr} \citep{tourr, tourr-cran}
-package, where the projected ellipsoid can be drawn for each projection.
-The guided tour will take arguments specifying the data, and the null
-variance-covariance matrix.
+This is implemented in the \texttt{tourr} (Hadley Wickham et al. 2011;
+H. Wickham et al. 2024) package, where the projected ellipsoid can be
+drawn for each projection. The guided tour will take arguments
+specifying the data, and the null variance-covariance matrix.
 
@@
 alanine aminotransferase (ALT). There are normal ranges on these
-measurements reported by \citet{lib-med-liver-norms} and listed in
+measurements reported by Lala, Zubair, and Minter (2023) and listed in
 Table~\ref{tbl-liver-norms}.
@@
 sample has slightly lower ALT and ALP, which is consistent with what is
-reported in \citet{lib-med-liver-norms}.
+reported in Lala, Zubair, and Minter (2023).
 
@@
 
-This example is motivated by the weather data example from
-\citet{filzmoser2018}. We illustrate using the anomaly index to compare
+This example is motivated by the weather data example from Filzmoser,
+Hron, and Templ (2018). We illustrate using the anomaly index to compare
 potential outliers with a reference normal distribution that is derived
@@
 June, July and August, reported for 68 years (1955 - 2022) (see
-\citet{filzmoser2018} for more details).
+Filzmoser, Hron, and Templ (2018) for more details).
 
@@
 (MAD), before applying the minimum covariance determinant (MCD)
-estimator \citep{rousseeuw1985multivariate} using the implementation in
-\citet{robustbase}. The MCD estimates for the mean vector and the
+estimator (Peter J. Rousseeuw 1985) using the implementation in Maechler
+et al. (2024). The MCD estimates for the mean vector and the
 variance-covariance matrix are then used to define the reference normal
@@
 out of the \(68\) observations as outlying. Since these are outlying in
-different combinations of variables, as found in \citet{filzmoser2018},
-we will further separate the outlying points into clusters based on
-similarity in direction. The similarity is computed by first normalizing
-observations to have length \(1\) and then apply k-means clustering with
-Euclidean distance. We use the Dunn index \citet{dunn}, computed using
-\citet{fpc}, to select the preferred number of clusters, \(k=5\).
+different combinations of variables, as found in Filzmoser, Hron, and
+Templ (2018), we will further separate the outlying points into clusters
+based on similarity in direction. The similarity is computed by first
+normalizing observations to have length \(1\) and then apply k-means
+clustering with Euclidean distance. We use the Dunn index Halkidi,
+Batistakis, and Vazirgiannis (2002), computed using Hennig (2024), to
+select the preferred number of clusters, \(k=5\).
 
@@
 To compare our results to what is obtained by outlier detection methods,
-we reproduce one of the results from \citet{filzmoser2018} in
-Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
-all points that we have identified as outliers using the
+we reproduce one of the results from Filzmoser, Hron, and Templ (2018)
+in Figure~\ref{fig-shapley}. On top we can see the cluster assignments
+for all points that we have identified as outliers using the
 \(p\)-dimensional Mahalanobis distance.
@@
 For example we see that cluster 4 contains the years 1962 and 1996, and
-for these two years the result from \citet{filzmoser2018} also indicates
-unusually low values for the minimum teperature, while there is
-disagreement between outlyingness in other variables. Note that compared
-to the years 1965 and 1966 they have unusually low minimum temperature
-while having average values for the number of days with maximum
-temperature above 25°C, and this contrast is also captured in the
-projection.
+for these two years the result from Filzmoser, Hron, and Templ (2018)
+also indicates unusually low values for the minimum teperature, while
+there is disagreement between outlyingness in other variables. Note that
+compared to the years 1965 and 1966 they have unusually low minimum
+temperature while having average values for the number of days with
+maximum temperature above 25°C, and this contrast is also captured in
+the projection.
 
@@
 
-\caption{\label{fig-shapley}Reproduction of the results from
-\citet{filzmoser2018}, showing which cells have been identified as
+\caption{\label{fig-shapley}Reproduction of the results from Filzmoser,
+Hron, and Templ (2018), showing which cells have been identified as
 outlying by their Shapley Cell Detector algorithm. Red cells indicate
@@
 
-The work is related to outlier detection methods \citep[see,][for
-example]{filzmoser2018}, particularly those that use robust statistics,
-as illustrated in the weather example. Methods discussed in
-\citet{directional-outlyingness} and \citet{donoho1982} describe
-detecting outlyingness using projections of the data. When these
-projections are collected and used as a large set, they define a
-high-dimensional confidence region of an indeterminate shape. There is
-no analytical definition. However, we could explore the data relative to
-the high-dimensional shapes using the same approach described in
-Section~\ref{sec-background}. You would generate points on the surface
-of the irregular shape, overlay the data on this to visualise with a
-tour.
+The work is related to outlier detection methods (see, Filzmoser, Hron,
+and Templ 2018, for example), particularly those that use robust
+statistics, as illustrated in the weather example. Methods discussed in
+Peter J. Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko
+(1992) describe detecting outlyingness using projections of the data.
+When these projections are collected and used as a large set, they
+define a high-dimensional confidence region of an indeterminate shape.
+There is no analytical definition. However, we could explore the data
+relative to the high-dimensional shapes using the same approach
+described in Section~\ref{sec-background}. You would generate points on
+the surface of the irregular shape, overlay the data on this to
+visualise with a tour.
 
@@
 Another potential direction of this work is with model-based clustering
-using Gaussian mixtures \citep{FR02}, where ellipses form the basis of
-the model. Currently, the \texttt{mclust} software \citep{mclust} has
-the capacity to show \gD{} ellipses for two variables, or axis parallel
-\gD{} ellipses for \pD{}. The equations developed here for drawing the
-\gD{} projection of a \pD{} ellipse would provide more versatile
-visualisation for examining the model-based clustering fits.
+using Gaussian mixtures (Fraley and Raftery 2002), where ellipses form
+the basis of the model. Currently, the \texttt{mclust} software (Scrucca
+et al. 2023) has the capacity to show \gD{} ellipses for two variables,
+or axis parallel \gD{} ellipses for \pD{}. The equations developed here
+for drawing the \gD{} projection of a \pD{} ellipse would provide more
+versatile visualisation for examining the model-based clustering fits.
 
@@
 
-  \bibliography{bibliography.bib}
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-As85}
+Asimov, D. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
+{M}ultidimensional {D}ata.''} \emph{SIAM Journal of Scientific and
+Statistical Computing} 6 (1): 128--43.
+
+\bibitem[\citeproctext]{ref-BA86b}
+Buja, A., and D. Asimov. 1986. {``{G}rand {T}our {M}ethods: {A}n
+{O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.
+
+\bibitem[\citeproctext]{ref-BCAH05}
+Buja, A., D. Cook, D. Asimov, and C. Hurley. 2005. {``{C}omputational
+{M}ethods for {H}igh-{D}imensional {R}otations in {D}ata
+{V}isualization.''} In \emph{Handbook of Statistics: Data Mining and
+Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
+391--414. Amsterdam, The Netherlands: Elsevier/North-Holland.
+
+\bibitem[\citeproctext]{ref-cook1995}
+Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
+{``Grand Tour and Projection Pursuit.''} \emph{Journal of Computational
+and Graphical Statistics} 4 (3): 155--72.
+\url{https://doi.org/10.1080/10618600.1995.10474674}.
+
+\bibitem[\citeproctext]{ref-CLBW06}
+Cook, D., E.-K. Lee, A. Buja, and H. Wickham. 2006. {``{G}rand {T}ours,
+{P}rojection {P}ursuit {G}uided {T}ours and {M}anual {C}ontrols.''} In
+\emph{Handbook of {D}ata {V}isualization}, edited by C.-H. Chen, W.
+Härdle, and A. Unwin. Berlin: Springer.
+
+\bibitem[\citeproctext]{ref-donoho1982}
+Donoho, David L., and Miriam Gasko. 1992. {``{Breakdown Properties of
+Location Estimates Based on Halfspace Depth and Projected
+Outlyingness}.''} \emph{The Annals of Statistics} 20 (4): 1803--27.
+\url{https://doi.org/10.1214/aos/1176348890}.
+
+\bibitem[\citeproctext]{ref-filzmoser2018}
+Filzmoser, P., K. Hron, and M Templ. 2018. \emph{Applied Compositional
+Data Analysis}. Cham, Switzerland: Springer.
+\url{https://doi.org/10.1007/978-3-319-96422-5}.
+
+\bibitem[\citeproctext]{ref-FR02}
+Fraley, C., and A. E. Raftery. 2002. {``Model-Based {C}lustering,
+{D}iscriminant {A}nalysis, {D}ensity {E}stimation.''} \emph{Journal of
+the American Statistical Association} 97: 611--31.
+
+\bibitem[\citeproctext]{ref-FT74}
+Friedman, J. H., and J. W. Tukey. 1974. {``{A} {P}rojection {P}ursuit
+{A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
+Transactions on Computing C} 23: 881--89.
+
+\bibitem[\citeproctext]{ref-Ga71}
+Gabriel, K. R. 1971. {``The {B}iplot {G}raphical {D}isplay of {M}atrices
+with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
+\emph{Biometrika} 58: 453--67.
+
+\bibitem[\citeproctext]{ref-GH96}
+Gower, J. C., and D. J. Hand. 1996. \emph{Biplots}. London: Chapman;
+Hall.
+
+\bibitem[\citeproctext]{ref-dunn}
+Halkidi, Maria, Yannis Batistakis, and Michalis Vazirgiannis. 2002.
+{``Clustering Validity Checking Methods: Part II.''} \emph{ACM SIGMOD
+Record} 31 (September). \url{https://doi.org/10.1145/601858.601862}.
+
+\bibitem[\citeproctext]{ref-fpc}
+Hennig, Christian. 2024. \emph{Fpc: Flexible Procedures for Clustering}.
+\url{https://CRAN.R-project.org/package=fpc}.
+
+\bibitem[\citeproctext]{ref-Hu85}
+Huber, P. J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
+\emph{Annals of Statistics} 13: 435--525.
+
+\bibitem[\citeproctext]{ref-JS87}
+Jones, M. C., and R. Sibson. 1987. {``{W}hat Is {P}rojection {P}ursuit?
+(With Discussion).''} \emph{Journal of the Royal Statistical Society,
+Series A} 150: 1--36.
+
+\bibitem[\citeproctext]{ref-lib-med-liver-norms}
+Lala, V., M. Zubair, and D. A. Minter. 2023. {``Liver Function Tests.''}
+\url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.
+
+\bibitem[\citeproctext]{ref-tours2022}
+Lee, Stuart, Dianne Cook, Natalia da Silva, Ursula Laa, Nicholas
+Spyrison, Earo Wang, and H. Sherry Zhang. 2022. {``The State-of-the-Art
+on Tours for Dynamic Visualization of High-Dimensional Data.''}
+\emph{WIREs Computational Statistics} 14 (4): e1573.
+\url{https://doi.org/10.1002/wics.1573}.
+
+\bibitem[\citeproctext]{ref-robustbase}
+Maechler, Martin, Peter Rousseeuw, Christophe Croux, Valentin Todorov,
+Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel
+Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
+\emph{Robustbase: Basic Robust Statistics}.
+\url{http://robustbase.r-forge.r-project.org/}.
+
+\bibitem[\citeproctext]{ref-mahalanobis}
+Mahalanobis, P. C. 1936. {``On the Generalised Distance in
+Statistics.''} \emph{Sankhya A} 80 (Suppl 1): 1--7.
+\url{https://doi.org/10.1007/s13171-019-00164-5}.
+
+\bibitem[\citeproctext]{ref-rousseeuw1985multivariate}
+Rousseeuw, Peter J. 1985. {``Multivariate Estimation with High Breakdown
+Point.''} \emph{Mathematical Statistics and Applications} 8 (283-297):
+37.
+
+\bibitem[\citeproctext]{ref-directional-outlyingness}
+Rousseeuw, Peter J., Jakob Raymaekers, and Mia Hubert. 2018. {``A
+Measure of Directional Outlyingness with Applications to Image Data and
+Video.''} \emph{Journal of Computational and Graphical Statistics} 27
+(2): 345--59. \url{https://doi.org/10.1080/10618600.2017.1366912}.
+
+\bibitem[\citeproctext]{ref-mclust}
+Scrucca, Luca, Chris Fraley, T. Brendan Murphy, and Adrian E. Raftery.
+2023. \emph{Model-Based Clustering, Classification, and Density
+Estimation Using {mclust} in {R}}. Chapman; Hall/CRC.
+\url{https://doi.org/10.1201/9781003277965}.
+
+\bibitem[\citeproctext]{ref-tourr}
+Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
+{``Tourr: {An} {R} {Package} for {Exploring} {Multivariate} {Data} with
+{Projections}.''} \emph{Journal of Statistical Software} 40 (2).
+\url{https://doi.org/10.18637/jss.v040.i02}.
+
+\bibitem[\citeproctext]{ref-tourr-cran}
+Wickham, H., D. Cook, N. Spyrison, N. Laa, H. S. Zhang, and S. Lee.
+2024. \emph{Tour Methods for Multivariate Data Visualisation}.
+\url{https://CRAN.R-project.org/package=tourr}.
+
+\end{CSLReferences}
 

### 2024-07-24 paper.qmd
References fixed

@@
   Many data problems contain some reference or normal conditions, upon which to compare newly collected data.
-This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms.
+This scenario occurs in data collected as part of clinical trials to detect adverse events, or for measuring climate change against historical norms.
 The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution.
@@
 @sec-background provides more context for the methods and visualisation.
-@sec-anomaly-index provides the details of the new index, and example use is illustrated in @sec-examples.
+@sec-anomaly-index provides the details of the new index, and @sec-implementation describes the implementation.
+Example use is illustrated in @sec-examples.
 
@@
 
-## Implementation
+## Implementation {#sec-implementation}
 
@@
 
-This example is motivated by the weather data example from @filzmoser2018.
+This example is motivated by the weather data example from @MAYRHOFER2023.
 We illustrate using the anomaly index to compare potential outliers with a reference normal distribution that is derived using robust methods on the original data.
-The data contains 16 numerical measurements that are averages across the three summer months June, July and August, reported for 68 years (1955 - 2022) (see @filzmoser2018 for more details).
+The data contains 16 numerical measurements that are averages across the three summer months June, July and August, reported for 68 years (1955 - 2022) (see @MAYRHOFER2023 for more details).
 
@@
 This will identify $20$ out of the $68$ observations as outlying.
-Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters based on  similarity in direction.
+Since these are outlying in different combinations of variables, as found in @MAYRHOFER2023, we will further separate the outlying points into clusters based on  similarity in direction.
 The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance.
@@
 
-To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @filzmoser2018 in @fig-shapley.
+To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @MAYRHOFER2023 in @fig-shapley.
 On top we can see the cluster assignments for all points that we have identified as outliers using the $p$-dimensional Mahalanobis distance.
 
-For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @filzmoser2018 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables.
+For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables.
 Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.
@@
 
-Reproduction of the results from @filzmoser2018, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm.
+Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm.
 Red cells indicate values higher than expected, blue cells values lower than expected.
@@
 
-The work is related to outlier detection methods [see, @filzmoser2018, for example], particularly those that use robust statistics, as illustrated in the weather example.
+The work is related to outlier detection methods [see, @MAYRHOFER2023, for example], particularly those that use robust statistics, as illustrated in the weather example.
 Methods discussed in @directional-outlyingness and @donoho1982 describe detecting outlyingness using projections of the data.
@@
 
+## Supplementary Materials {-}
 The full paper, code and data to reproduce the work, are publicly available at https://github.com/uschiLaa/anomaly_ppi.
+
+## References {-}

### 2024-07-24 paper.tex
References fixed

@@
 which to compare newly collected data. This scenario occurs in data
-collected as part of clinical trials to detect adverse events, and or
-for measuring climate change against historical norms. The data is
-typically multivariate, and often the normal ranges are specified by a
+collected as part of clinical trials to detect adverse events, or for
+measuring climate change against historical norms. The data is typically
+multivariate, and often the normal ranges are specified by a
 multivariate normal distribution. The work presented in this paper
@@
 called a grand tour (Asimov 1985; Buja and Asimov 1986; Buja et al.
-2005; D. Cook et al. 2006; Lee et al. 2022). The combination of these
-two approaches would then use a projection pursuit index to select
+2005; Cook et al. 2006; Lee et al. 2022). The combination of these two
+approaches would then use a projection pursuit index to select
 interesting projections, but display them via an interpolated path to
-provide context. This is called a guided tour (Dianne Cook et al. 1995).
+provide context. This is called a guided tour (Cook et al. 1995).
 
@@
 visualisation. Section~\ref{sec-anomaly-index} provides the details of
-the new index, and example use is illustrated in
+the new index, and Section~\ref{sec-implementation} describes the
+implementation. Example use is illustrated in
 Section~\ref{sec-examples}.
@@
 To define a measure of an interesting projection is to maximize the
-average Mahahlanobis distance (Mahalanobis 1936) in the projection for a
+average Mahalanobis distance (Mahalanobis 1936) in the projection for a
 subset of points, \(W\). The set of points could be chosen in different
@@
 
-\section{Implementation}\label{implementation}
+\section{Implementation}\label{sec-implementation}
 
-This is implemented in the \texttt{tourr} (Hadley Wickham et al. 2011;
-H. Wickham et al. 2024) package, where the projected ellipsoid can be
-drawn for each projection. The guided tour will take arguments
-specifying the data, and the null variance-covariance matrix.
+This is implemented in the \texttt{tourr} (Wickham et al. 2011, 2024)
+package, where the projected ellipsoid can be drawn for each projection.
+The guided tour will take arguments specifying the data, and the null
+variance-covariance matrix.
 
@@
 
-This example is motivated by the weather data example from Filzmoser,
-Hron, and Templ (2018). We illustrate using the anomaly index to compare
+This example is motivated by the weather data example from Mayrhofer and
+Filzmoser (2023). We illustrate using the anomaly index to compare
 potential outliers with a reference normal distribution that is derived
@@
 June, July and August, reported for 68 years (1955 - 2022) (see
-Filzmoser, Hron, and Templ (2018) for more details).
+Mayrhofer and Filzmoser (2023) for more details).
 
@@
 (MAD), before applying the minimum covariance determinant (MCD)
-estimator (Peter J. Rousseeuw 1985) using the implementation in Maechler
-et al. (2024). The MCD estimates for the mean vector and the
+estimator (Rousseeuw 1985) using the implementation in Maechler et al.
+(2024). The MCD estimates for the mean vector and the
 variance-covariance matrix are then used to define the reference normal
@@
 out of the \(68\) observations as outlying. Since these are outlying in
-different combinations of variables, as found in Filzmoser, Hron, and
-Templ (2018), we will further separate the outlying points into clusters
-based on similarity in direction. The similarity is computed by first
+different combinations of variables, as found in Mayrhofer and Filzmoser
+(2023), we will further separate the outlying points into clusters based
+on similarity in direction. The similarity is computed by first
 normalizing observations to have length \(1\) and then apply k-means
@@
 To compare our results to what is obtained by outlier detection methods,
-we reproduce one of the results from Filzmoser, Hron, and Templ (2018)
-in Figure~\ref{fig-shapley}. On top we can see the cluster assignments
-for all points that we have identified as outliers using the
+we reproduce one of the results from Mayrhofer and Filzmoser (2023) in
+Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
+all points that we have identified as outliers using the
 \(p\)-dimensional Mahalanobis distance.
@@
 For example we see that cluster 4 contains the years 1962 and 1996, and
-for these two years the result from Filzmoser, Hron, and Templ (2018)
-also indicates unusually low values for the minimum teperature, while
-there is disagreement between outlyingness in other variables. Note that
+for these two years the result from Mayrhofer and Filzmoser (2023) also
+indicates unusually low values for the minimum teperature, while there
+is disagreement between outlyingness in other variables. Note that
 compared to the years 1965 and 1966 they have unusually low minimum
@@
 
-\caption{\label{fig-shapley}Reproduction of the results from Filzmoser,
-Hron, and Templ (2018), showing which cells have been identified as
+\caption{\label{fig-shapley}Reproduction of the results from Mayrhofer
+and Filzmoser (2023), showing which cells have been identified as
 outlying by their Shapley Cell Detector algorithm. Red cells indicate
@@
 
-The work is related to outlier detection methods (see, Filzmoser, Hron,
-and Templ 2018, for example), particularly those that use robust
+The work is related to outlier detection methods (see, Mayrhofer and
+Filzmoser 2023, for example), particularly those that use robust
 statistics, as illustrated in the weather example. Methods discussed in
-Peter J. Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko
-(1992) describe detecting outlyingness using projections of the data.
-When these projections are collected and used as a large set, they
-define a high-dimensional confidence region of an indeterminate shape.
-There is no analytical definition. However, we could explore the data
-relative to the high-dimensional shapes using the same approach
-described in Section~\ref{sec-background}. You would generate points on
-the surface of the irregular shape, overlay the data on this to
-visualise with a tour.
+Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko (1992)
+describe detecting outlyingness using projections of the data. When
+these projections are collected and used as a large set, they define a
+high-dimensional confidence region of an indeterminate shape. There is
+no analytical definition. However, we could explore the data relative to
+the high-dimensional shapes using the same approach described in
+Section~\ref{sec-background}. You would generate points on the surface
+of the irregular shape, overlay the data on this to visualise with a
+tour.
 
@@
 
+\section*{Supplementary Materials}\label{supplementary-materials}
+\addcontentsline{toc}{section}{Supplementary Materials}
+
 The full paper, code and data to reproduce the work, are publicly
 available at https://github.com/uschiLaa/anomaly\_ppi.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 
@@
 \bibitem[\citeproctext]{ref-As85}
-Asimov, D. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
+Asimov, Daniel. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
 {M}ultidimensional {D}ata.''} \emph{SIAM Journal of Scientific and
@@
 \bibitem[\citeproctext]{ref-BA86b}
-Buja, A., and D. Asimov. 1986. {``{G}rand {T}our {M}ethods: {A}n
-{O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.
+Buja, Andreas, and Daniel Asimov. 1986. {``{G}rand {T}our {M}ethods:
+{A}n {O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.
 
 \bibitem[\citeproctext]{ref-BCAH05}
-Buja, A., D. Cook, D. Asimov, and C. Hurley. 2005. {``{C}omputational
-{M}ethods for {H}igh-{D}imensional {R}otations in {D}ata
-{V}isualization.''} In \emph{Handbook of Statistics: Data Mining and
-Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
+Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
+{``{C}omputational {M}ethods for {H}igh-{D}imensional {R}otations in
+{D}ata {V}isualization.''} In \emph{Handbook of Statistics: Data Mining
+and Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
 391--414. Amsterdam, The Netherlands: Elsevier/North-Holland.
@@
 \bibitem[\citeproctext]{ref-CLBW06}
-Cook, D., E.-K. Lee, A. Buja, and H. Wickham. 2006. {``{G}rand {T}ours,
-{P}rojection {P}ursuit {G}uided {T}ours and {M}anual {C}ontrols.''} In
-\emph{Handbook of {D}ata {V}isualization}, edited by C.-H. Chen, W.
-Härdle, and A. Unwin. Berlin: Springer.
+Cook, Dianne, Eun-Kyung Lee, Andreas Buja, and Hadley Wickham. 2006.
+{``{G}rand {T}ours, {P}rojection {P}ursuit {G}uided {T}ours and {M}anual
+{C}ontrols.''} In \emph{Handbook of {D}ata {V}isualization}, edited by
+C.-H. Chen, W. Härdle, and A. Unwin. Berlin: Springer.
 
@@
 
-\bibitem[\citeproctext]{ref-filzmoser2018}
-Filzmoser, P., K. Hron, and M Templ. 2018. \emph{Applied Compositional
-Data Analysis}. Cham, Switzerland: Springer.
-\url{https://doi.org/10.1007/978-3-319-96422-5}.
-
 \bibitem[\citeproctext]{ref-FR02}
-Fraley, C., and A. E. Raftery. 2002. {``Model-Based {C}lustering,
+Fraley, Chris, and Adrian Raftery. 2002. {``Model-Based {C}lustering,
 {D}iscriminant {A}nalysis, {D}ensity {E}stimation.''} \emph{Journal of
@@
 \bibitem[\citeproctext]{ref-FT74}
-Friedman, J. H., and J. W. Tukey. 1974. {``{A} {P}rojection {P}ursuit
-{A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
+Friedman, Jerome H., and John W. Tukey. 1974. {``{A} {P}rojection
+{P}ursuit {A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
 Transactions on Computing C} 23: 881--89.
@@
 \bibitem[\citeproctext]{ref-Ga71}
-Gabriel, K. R. 1971. {``The {B}iplot {G}raphical {D}isplay of {M}atrices
-with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
+Gabriel, K. Ruben. 1971. {``The {B}iplot {G}raphical {D}isplay of
+{M}atrices with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
 \emph{Biometrika} 58: 453--67.
@@
 \bibitem[\citeproctext]{ref-GH96}
-Gower, J. C., and D. J. Hand. 1996. \emph{Biplots}. London: Chapman;
-Hall.
+Gower, John C., and David J. Hand. 1996. \emph{Biplots}. London:
+Chapman; Hall.
 
@@
 \bibitem[\citeproctext]{ref-Hu85}
-Huber, P. J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
+Huber, Peter J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
 \emph{Annals of Statistics} 13: 435--525.
@@
 \bibitem[\citeproctext]{ref-JS87}
-Jones, M. C., and R. Sibson. 1987. {``{W}hat Is {P}rojection {P}ursuit?
-(With Discussion).''} \emph{Journal of the Royal Statistical Society,
-Series A} 150: 1--36.
+Jones, M. C., and Robin Sibson. 1987. {``{W}hat Is {P}rojection
+{P}ursuit? (With Discussion).''} \emph{Journal of the Royal Statistical
+Society, Series A} 150: 1--36.
 
 \bibitem[\citeproctext]{ref-lib-med-liver-norms}
-Lala, V., M. Zubair, and D. A. Minter. 2023. {``Liver Function Tests.''}
-\url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.
+Lala, Vasimahmed, Muhammad Zubair, and David A. Minter. 2023. {``Liver
+Function Tests.''} \url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.
 
@@
 \bibitem[\citeproctext]{ref-robustbase}
-Maechler, Martin, Peter Rousseeuw, Christophe Croux, Valentin Todorov,
-Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel
-Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
+Maechler, Martin, Peter J. Rousseeuw, Christophe Croux, Valentin
+Todorov, Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke,
+Manuel Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
 \emph{Robustbase: Basic Robust Statistics}.
@@
 \bibitem[\citeproctext]{ref-mahalanobis}
-Mahalanobis, P. C. 1936. {``On the Generalised Distance in
+Mahalanobis, Prasanta Chandra. 1936. {``On the Generalised Distance in
 Statistics.''} \emph{Sankhya A} 80 (Suppl 1): 1--7.
 \url{https://doi.org/10.1007/s13171-019-00164-5}.
+
+\bibitem[\citeproctext]{ref-MAYRHOFER2023}
+Mayrhofer, Marcus, and Peter Filzmoser. 2023. {``Multivariate Outlier
+Explanations Using Shapley Values and Mahalanobis Distances.''}
+\emph{Econometrics and Statistics}.
 
@@
 \bibitem[\citeproctext]{ref-tourr-cran}
-Wickham, H., D. Cook, N. Spyrison, N. Laa, H. S. Zhang, and S. Lee.
-2024. \emph{Tour Methods for Multivariate Data Visualisation}.
-\url{https://CRAN.R-project.org/package=tourr}.
+Wickham, Hadley, Dianne Cook, Nicholas Spyrison, Ursula Laa, H. Sherry
+Zhang, and Stuart Lee. 2024. \emph{Tour Methods for Multivariate Data
+Visualisation}. \url{https://CRAN.R-project.org/package=tourr}.
 

### 2024-12-11 paper.qmd
Start working on the revision

@@
 This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution.
+\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
+However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures.
+Classical methods to detect outliers with projection pursuit are based on skewness (Loperfido) (even the classical indexes were supposed to detect outliers I think?)....
+also dobin package by Sevvandi might be relevant here.
+XXX when/why might it be better to use our method?}
+
 It is organised as follows.
@@
 The norm is represented by points on the surface of a $p$-dimensional ellipsoid, corresponding to a confidence level.
-A sample of points uniformly distributed on a $p$-dimensional sphere is generated by
+A sample of points on a $p$-dimensional ellipsoid is generated by
 
@@
 from $N_p(\mathbfit{\mu}, \Sigma)$. 
-2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. 
-To convert this to points on the surface of a confidence ellipsoid,
-
-3. transform the shape using a specific variance covariance, and shift to center on the mean vector.
+2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. The sample is now uniform on the surface of the hypersphere.
+3. Transform the shape from a sphere to an ellipsoid using the specific variance covariance, and shift to center on the mean vector.
 
@@
 
-A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by
+Let $x$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by
 
-$$
+\begin{equation}
 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
-$$
+\label{eq:pd}
+\end{equation}
+
 where $c$ a constant that depends on a specific confidence level.
@@
   The projection of this \pD{} ellipsoid in \gD{} has the equation
-$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
+\begin{equation}
+(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.
+\end{equation}
 \end{theorem}
@@
 
-$$\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
-$$
+\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+\end{equation}
 
@@
 projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
-Making this substitution in the \pD{} ellipsoid equation yields
+Making this substitution in Equation~(\ref{eq:pd}) yields
 
-$$c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
-$$
+\begin{equation}
+c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+\label{eq:c2}
+\end{equation}
 
@@
 Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
-We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the equation \(c^2 = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T\).
+We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in Equation~(\ref{eq:c2}).
 From this we can compute the analog equation for the projection as
 
-$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
+\begin{equation}
+(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2
+\end{equation}
 
@@
 To define a measure of an interesting projection is to maximize the average Mahalanobis distance [@mahalanobis] in the projection for a subset of points, $W$.
-The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}.
+The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}, $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$.
 Alternatives could be to select a set of observations with the largest Mahalanobis distance, manually select observations or possibly a group of points identified via clustering of the extremes.
 
-The index is written as
+We define our new index as
 
-$$
-\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
-$$
-
-where by default $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$, is the set of observations outside the \pD{} ellipsoid.
+\begin{equation}
+\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T.
+\end{equation}
 
@@
 Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient.
-Red cross indicates observation is outide the 4-D confidence ellipse.
+Red cross indicates observation is outside the 4-D confidence ellipse.
 The female patients tend to have lower ALP and ALT than the normal range.

### 2024-12-11 paper.tex
Final fix references
Start working on the revision

@@
 by projections where a new sample is most distant from the existing
-normal distribution. It is organised as follows.
-Section~\ref{sec-background} provides more context for the methods and
-visualisation. Section~\ref{sec-anomaly-index} provides the details of
-the new index, and Section~\ref{sec-implementation} describes the
-implementation. Example use is illustrated in
-Section~\ref{sec-examples}.
+normal distribution.
+\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
+However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures.
+Classical methods to detect outliers with projection pursuit are based on skewness (Loperfido) (even the classical indexes were supposed to detect outliers I think?)....
+also dobin package by Sevvandi might be relevant here.
+XXX when/why might it be better to use our method?}
+
+It is organised as follows. Section~\ref{sec-background} provides more
+context for the methods and visualisation.
+Section~\ref{sec-anomaly-index} provides the details of the new index,
+and Section~\ref{sec-implementation} describes the implementation.
+Example use is illustrated in Section~\ref{sec-examples}.
 
@@
 surface of a \(p\)-dimensional ellipsoid, corresponding to a confidence
-level. A sample of points uniformly distributed on a \(p\)-dimensional
-sphere is generated by
+level. A sample of points on a \(p\)-dimensional ellipsoid is generated
+by
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
@@
   Transforming each observation to have unit distance from the mean,
-  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\). To convert this
-  to points on the surface of a confidence ellipsoid,
+  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\). The sample is now
+  uniform on the surface of the hypersphere.
 \item
-  transform the shape using a specific variance covariance, and shift to
-  center on the mean vector.
+  Transform the shape from a sphere to an ellipsoid using the specific
+  variance covariance, and shift to center on the mean vector.
 \end{enumerate}
@@
 
-A \pD{} ellipsoid corresponding to a given variance-covariance
-(\(\Sigma\)) is defined by
-
-\[
+Let \(x\) be a \pD{} vector. A \pD{} ellipsoid corresponding to a given
+variance-covariance (\(\Sigma\)) is defined by
+
+\begin{equation}
 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
-\] where \(c\) a constant that depends on a specific confidence level.
+\label{eq:pd}
+\end{equation}
+
+where \(c\) a constant that depends on a specific confidence level.
 
@@
 projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
-Making this substitution in the \pD{} ellipsoid equation yields
-
-$$c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
-$$
+Making this substitution in Equation~(\ref{eq:pd}) yields
+
+\begin{equation}
+c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
+\label{eq:c2}
+\end{equation}
 
@@
 
-\[
+\begin{equation}
 \sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
-\]
+\end{equation}
 
@@
 \emph{Econometrics and Statistics}.
+\url{https://doi.org/10.1016/j.ecosta.2023.04.003}.
 

### 2024-12-11 review/reply.Rmd
Start working on the revision

---
title: "Reply to reviewers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We thank the reviewers for their feedback on the paper. Since the article was submitted as a *short technical note* it is not possible to accomodate all suggestions.

## Reviewer 1

1. Providing a more comprehensive review of existing methods would enhance the context for the proposed index.

\textcolor{purple}{A comprehensive review would be beyond the scope of a short technical note, but we added additional context XXXX do this? how?}

2. A detailed comparison with existing methods, emphasizing how the new index offers improvements, would clarify the study's rationale. Including a comparative analysis with current projection pursuit methods, while highlighting the strengths and potential limitations of the proposed index, would further strengthen the paper.

\textcolor{purple}{To the best of our knowledge, the proposed method is the first that allows the comparison of a reference normal distribution to observations using projection pursuit.}

3. Including additional real-world data examples would enhance the practical relevance and robustness of the findings.

\textcolor{purple}{The second example in the paper is using real-world data, adding more examples would be beyond the scope of a short technical note.}

4. The methodology is well-explained, but providing more details on the computational aspects and potential limitations would improve reproducibility. Clarification on how to obtain and interpret the "index" results is needed. Offering more detailed descriptions of the algorithms and their implementation would aid in understanding. A deeper discussion of the practical implications of these results would add valuable context.

\textcolor{purple}{XXXX need to add something? There is also the implementation and its documentation we could point to...}

## Reviewer 2

The paper's ideas are interesting, but some sections of the article are very brief and summarized, making it difficult to understand the method. Additional references to similar works are needed to better justify the main contribution of the work. It is also necessary to number the main equations and reference them in the text when appropriate. More details in the pdf.

\textcolor{purple}{The description is brief, fitting the style of a short technical note. XXX check where we could be more clear...}

\textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow. XXX to check @Annalisa}

The proposed method addresses a similar question to that of a normality test for a specific data sample, but it can also be used as a method to detect outliers—data that were generated from a distribution other than the reference multivariate normal. In this regard, what differentiates your method from similar approaches, and why do you propose it as an improvement over existing methods? I believe it would be helpful to include additional context and references on other projection pursuit contributions aimed at finding interesting projections in the same direction as your approach.

\textcolor{purple}{XXX (Need to finish, add references) We have added a paragraph in the introduction discussing outlier detection with other projection pursuit index functions, including the most important references.}

In page 2 line 47: “This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution.” At this point, it would be beneficial to emphasize the significance of your findings and highlight the main contribution of your paper. Clarifying how your method advances the field or outperforms existing techniques will strengthen your argument and provide a clearer understanding of the value of your work.

\textcolor{purple}{We have added a short review of traditional outlier detection methods and relevant methods here, and emphasized where they currently do not apply.}

Page 3 line 26: Could you explicitly define the shape transformation mentioned in point 3?

\textcolor{purple}{We have rephrased points 2 and 3 to make this more clear.}

Page 3 line 31: Are the new observations transformed in any way before being plotted for comparison? If so, could you clarify what kind of transformations are applied and how they impact the comparison with the reference distribution?

\textcolor{purple}{XXX I think the answer should be no, there is no comparison? Only rescaling for fitting into the plotting window which would be applied to both the nex points and the ellipsoid?}

Page 3 line 57: “Figure 2 compares a new sample of patient scores against the normal range”, However, there is no prior context provided for this example, making it unclear to the reader. Additionally, on page 4, you mention “normal patients” as if the reader is already familiar with the patient example. It would be helpful to introduce and explain the patient data context earlier in the paper to avoid confusion.

\textcolor{purple}{XXX todo @Di}

Figure 2: The labels in the figures are not clear. In panel a), the label “ci” is confusing, and in panel b), the label “norm” may lead to confusion regarding normal patients. To enhance clarity, consider increasing the point size and applying transparency, similar to the size and transparency used in Figure 1.

\textcolor{purple}{XXX todo @Di}

Section 3: The flow of the content is difficult to follow, and some equations would benefit from being described in a different order. Additionally, numbering some of the equations and referencing them in the text would significantly improve readability.

\textcolor{purple}{We have numbered the main equations and now use referencing in the description. XXX check flow of content @Annalisa}

Page 4 line 57: Should include “Let x a p dimensional vector”

\textcolor{purple}{This has been added.}

Page 5 line 7: Please number the equations and reference them in the text (for example in line 16 and 43). This will enhance clarity and make it easier for readers to follow your arguments

\textcolor{purple}{Equation numbers and references have been added.}

Page 5 line 16: “Theorem. The projection of this p = D ellipsoid in 2-D has the equation” you should change “this” for “a p-D ellipsoid from equation 1” or something similar. Additionally, after the theorem, please explain the variables mu, p, P and y. While you provide some of this information in line 41 and on page 6, it would benefit from better organization for clarity. Furthermore, consider restructuring the proof to enhance its clarity.

\textcolor{purple}{XXX todo @Annalisa}

Subsection 3.2 is unclear and would benefit from additional details:
Page 6 line 34: “To define a measure of an interesting projection is to maximize the average Mahalanobis distance”, you should rephrase “is to maximize”. Additionally, since the Mahalanobis distance serves as a fundamental tool for detecting multivariate outliers, it would be beneficial to mention this and provide a definition of the distance before explaining your method. Moreover, I recommend considering a robust version of the Mahalanobis distance, as this could enhance the robustness of your analysis.

\textcolor{purple}{XXX todo @Ursula}
\textcolor{purple}{XXX I think we actually implicitly use robust version, since we use robust estimation when getting the covariance for the data in the second example}

Page 6 line 55: Please number the equation and explicitly state that this is the new projection pursuit index you are proposing.

\textcolor{purple}{Done.}

Page 6 line 37: You mentioned W but it would be beneficial to define it here. While you provide this information on page 7, including a definition in this context will enhance clarity for the reader.

\textcolor{purple}{We have moved the default definition for W to where it is first mentioned.}

Page 7 subection 3.3 It would be useful to include a simulation example demonstrating how to apply the method in cases where the observations deviate from the norm in different directions. This example should illustrate how to group these observations effectively, providing readers with a clearer understanding of the practical application of your method. If you address this point later in the paper, please mention it here to guide the reader.

\textcolor{purple}{XXX todo @Ursula (this is done in the second example, will mention it, not add simulations)}

Page 8 Figure 3: You can justify that your method is useful for gaining insights into the outliers in the data. Emphasizing this point will help underscore the practical significance of your approach and its role in identifying and understanding anomalies within the dataset.

\textcolor{purple}{XXX todo @Ursula}

Page 9 line 18: “There are normal ranges” maybe acceptable ranges is better

\textcolor{purple}{XXX todo @Di}

You should move Figure 4 to page 10. In the figure caption, please correct the phrase “Red cross indicates observation is outide the 4-D confidence ellipse” to fix the spelling of “outside” and 2-D.

\textcolor{purple}{XXX todo @Ursula (spelling is fixed, check position when other changes are done)}

Page 12: All your comments regarding the clusters are unclear because Figure 5 lacks color labels. Including these labels will help clarify your points and enhance the reader’s understanding of the clustering in the visual representation.

\textcolor{purple}{XXX todo @Ursula}

Page 13: In Figure 5, the selected color palette is not effective, making it difficult to distinguish the five clusters. Additionally, using white for one of the groups against a white background further complicates visibility. I recommend trying the Dark2 palette or another contrasting palette. Please also include color labels to enhance clarity and help the reader identify the different clusters easily.

\textcolor{purple}{XXX todo @Ursula}

Page 14 Figure 6, In Figure 6, please change the cluster number colors to match those in Figure 5. The cluster numbers are difficult to see, so consider increasing the font size or other option. Additionally, you should include a color legend for clarity. When describing the variables in the text, please include their code names in parentheses. Also, ensure that the variable names in Figures 5 and 6 are consistent for better coherence throughout the paper.

\textcolor{purple}{XXX todo @Ursula}

Page 14 line 48: Is Shapley Cell Detector algorithm defined in Mayrofer and Filzmoser(2023)?

\textcolor{purple}{XXX todo @Ursula (it is, check if this needs to be made more clear in the paper)}

Page 15 line 19: In the conclusion, you mention that your work relates to outlier detection methods, particularly those using robust statistics. However, this important point is not clearly explained in your paper until the examples. It is essential to clarify this relationship earlier in the paper, before the examples, to provide context for the reader. Additionally, in the conclusion, you include some references that could be incorporated earlier in the text. This would help justify the differences between your method and existing ones, enhancing the overall clarity and depth of your discussion.

\textcolor{purple}{XXX todo (check with what we change in the introduction)}

### 2024-12-12 paper.tex
fixing up theorem and proof

@@
 
-Let \(x\) be a \pD{} vector. A \pD{} ellipsoid corresponding to a given
-variance-covariance (\(\Sigma\)) is defined by
+Let \(\mathbfit{x}\) be a \(p\) dimensional vector. A \pD{} ellipsoid corresponding to a given
+variance-covariance (\(\Sigma\)) is described by the equation
 
@@
 
-where \(c\) a constant that depends on a specific confidence level.
-
+where \(c\) is a constant that depends on a specific confidence level.
+
+Let \(P\) be a \((p\times 2)\) matrix, whose columns form an orthonormal basis for the \gD{} projection space.
+Let \(\mathbfit{y}\) be a 2 dimensional vector, and let \(\mathbfit{\mu}_p := \mathbfit{\mu} P\) denote the projected mean.
 \begin{theorem}
-  The projection of this \pD{} ellipsoid in \gD{} has the equation
+  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by \(P\) has the equation
 \begin{equation}
@@
 \begin{proof}
-The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
-That is, the curve
-consists of $\mathbfit{x}$ satisfying
-
-\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+The projection of a hyperellipsoid onto \gD{} space is an ellipse.
+The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid.
+For \(\mathbfit{x}\) on the hyperellipsoid, its orthogonal projection \(\mathbfit{x}P\) is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at \(\mathbfit{x}\) is perpendicular to the projection plane described by \(P\).
+As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at \(\mathbfit{x}\) is parallel to the projection plane.
+Note that any vector parallel to the projection plane can be expressed as \(2\mathbfit{s}P^T\) for some \(\mathbfit{s} \in \mathbb{R}^2\).
+
+Therefore, the boundary of the ellipse consists of the points \(\mathbfit{x}P\) for which $\mathbfit{x}$ satisfies
+
+\begin{equation}
+2\mathbfit{s}P^T = \nabla \left[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T\right] = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}
+\label{eq:grad}
 \end{equation}
 
-for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
-projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+for some \(\mathbfit{s} \in \mathbb{R}^2\). Rearranging this equation gives \(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\).
 Making this substitution in Equation~(\ref{eq:pd}) yields
@@
 
-We call points in the projection that are on the curve $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for $\mathbfit{x}$ on the \pD{} curve.
-Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
-We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in Equation~(\ref{eq:c2}).
+We denote points on the boundary of the projected ellipse by \(\mathbfit{y}\), so that \(\mathbfit{y} = \mathbfit{x}P\) for some \(\mathbfit{x}\) satifying Equation~(\ref{eq:grad}).
+Then, as \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) by definition, \(\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P\), again substituting \(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\).
+We then substitute \((\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}\) for \(\mathbfit{s}\) in Equation~(\ref{eq:c2}).
 From this we can compute the analog equation for the projection as
@@
 
-This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
+This means the matrix \((P^T \Sigma P)^{-1}\) defines the ellipse in
 the \gD{} projection. In general \(c\) could be any constant, but

### 2025-01-03 paper.qmd
finished changes requested from me

@@
 It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference.
-A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically.
+A new projection pursuit index is defined for this purpose.
+The tour visualisation also includes drawing of the projected ellipse, which is computed analytically, corresponding to the reference distribution.
 The methods are implemented in the R package, `tourr`.
@@
   \newcommand\bD{$1\text{-}D$}
+  \newcommand\fD{$4\text{-}D$}
   \newtheorem*{theorem}{Theorem}
@@
 The question is whether we can use these techniques to assess new data samples in the context of an established normal, such as a specific multivariate normal distribution.
-In physics, the normal distribution may describe experimental results, or a global fit for a selected model, and we might want to compare to a set of other models.
+In physics, the normal distribution may describe experimental results, or a global fit for a selected model, and we might want to compare to sets of other models.
 In medical applications, the normal distribution might summarize historic data of a healthy population and we compare it to samples from new patients.
@@
 
-This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution.
-\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
+This paper describes a new projection pursuit index to find projections where a new sample is most distant from the existing normal distribution, that is integrated with the guided tour to visualise the ways that the data departs from the reference distribution.
+This work also expands on current outlier detection methods such as @Kandanaarachchi02012021 and @loperfido by providing integrated multivariate visualisation to examine the outlyingness in different projections.
+It also provides the ability to examine potential outliers interactively - when they are outlying in different ways it can interfere with detection numerically, but visually one might group different types of outlyingness, and operate the outlier detection on subsets.
+<!--\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
 However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures.
@@
 also dobin package by Sevvandi might be relevant here.
-XXX when/why might it be better to use our method?}
+XXX when/why might it be better to use our method?}-->
 
-It is organised as follows.
+The paper is organised as follows.
 @sec-background provides more context for the methods and visualisation.
-@sec-anomaly-index provides the details of the new index, and @sec-implementation describes the implementation.
-Example use is illustrated in @sec-examples.
+@sec-anomaly-index provides the details of the new index, and rendering projections of the reference ellipse.
+@sec-implementation describes the implementation.
+Two examples, one medical and one on climate extremes, illustrate usage in @sec-examples.
 
@@
 
-To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points.
-The norm is represented by points on the surface of a $p$-dimensional ellipsoid, corresponding to a confidence level.
-A sample of points on a $p$-dimensional ellipsoid is generated by
+Following the ideas expressed in @CS07, to compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, you would use two samples of points.
+The norm is represented by points on the surface of a $p$-dimensional ellipsoid, with size corresponding to a chosen level curve of the density function of the reference distribution.
+This can be generated using these steps:
 
-1. Simulating a sample of observations ($\mathbfit{x}$, which are \pD{} vectors)
-from $N_p(\mathbfit{\mu}, \Sigma)$. 
-2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. The sample is now uniform on the surface of the hypersphere.
-3. Transform the shape from a sphere to an ellipsoid using the specific variance covariance, and shift to center on the mean vector.
+1. Simulate a sample of observations ($\mathbfit{x}$, which are \pD{} vectors) from $N_p(\mathbfit{\mu}, \Sigma)$. 
+2. Transform each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. The sample is now uniform on the surface of the hypersphere.
+3. Transform the shape from a sphere to an ellipsoid using the specific variance-covariance.
+4.
+Shift the resulting sample to center it on the mean vector, and resize to correspond to the desired density level.
+This could also be treated like a confidence ellipse, where observations outside are more than, say, three standard deviations from the sample mean.
 
-Finally, new observations can be visually compared with this ellipsoid by
-
-4. plotting them together.
-
-@fig-ci illustrates this process for \gD. This is easiest way to view this normal region relative to a new sample for any \pD{} problem.
+New observations can be visually compared with this ellipsoid by plotting them together using a tour. @fig-ci illustrates this process for \gD.
 
@@
 #| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1 (b).
-A confidence ellipsoid is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipsoid (d)."
+A 95% confidence ellipsoid is generated by transforming the sphere relative to a specified variance-covariance matrix, and sizing using a critical value (c), and new observations (triangle, square) can be visually assessed to be inside or outside by plotting with the ellipsoid (d)."
 set.seed(717)
@@
 
-For example, @fig-compare compares a new sample of patient scores against the normal range represented by an ellipse (a) and also against a simulated sample of normal patients (b).
-While these are useful approaches, the ragged edges of the projected ellipse make it difficult to compare the new sample precisely agains the normal ranges.
-
 ```{r}
@@
 #| echo: false
-liver_stats <- tribble(
-  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
-  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
-  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
-  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
-  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
-)
 set.seed(1158)
@@
     norm_vc[j, i] <- norm_vc[i, j]
-colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
-norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
-colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
+colnames(norm_vc) <- c("V1", "V2", "V3", "V4")
+norm_samp <- rmvnorm(5000, sigma=as.matrix(norm_vc))
+colnames(norm_samp) <- c("V1", "V2", "V3", "V4")
 norm_samp <- as.data.frame(norm_samp)
-norm_ci <- gen_vc_ellipse(norm_vc) * 2
-colnames(norm_ci) <- c("GGT", "AST", "ALP", "ALT")
+norm_ci <- gen_vc_ellipse(norm_vc, n=5000) * 2
+colnames(norm_ci) <- c("V1", "V2", "V3", "V4")
 norm_ci <- as.data.frame(norm_ci)
@@
 set.seed(217)
-women_vc <- norm_vc
-women_vc[1,4] <- 0.5
-women_vc[4,1] <- 0.5
-women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
-                 sigma = women_vc)
-colnames(women) <- c("GGT", "AST", "ALP", "ALT")
-women <- as.data.frame(women)
+new_vc <- norm_vc
+new_vc[1,4] <- 0.5
+new_vc[4,1] <- 0.5
+new <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
+                 sigma = new_vc)
+colnames(new) <- c("V1", "V2", "V3", "V4")
+new <- as.data.frame(new)
 
-norm_plus <- bind_rows(norm_samp, women)
+norm_plus <- bind_rows(norm_samp, new)
 norm_plus <- norm_plus |>
-  mutate(type = factor(c(rep("norm", 500), rep("samp", 16))))
+  mutate(type = factor(c(rep("ref", 5000), rep("samp", 16))))
 animate_xy(norm_plus[,1:4], col=norm_plus$type)
+set.seed(114)
+tpath <- save_history(norm_plus[,1:4],
+  max = 500)
+ptcex <- ifelse(norm_plus$type == "samp", 2, 0.5)
+render_gif(norm_plus[,1:4], 
+           planned_tour(tpath),
+           display_xy(col=norm_plus$type,
+                      cex=ptcex,
+                      scale=TRUE, center=FALSE, 
+                      axes="bottomleft",
+                      half_range=5.2),
+           gif_file = "gifs/fig2b.gif",
+           frames = 500,
+           width = 400,
+           height = 400)
 
-norm_ci_plus <- bind_rows(norm_ci, women)
+norm_ci_plus <- bind_rows(norm_ci, new)
 norm_ci_plus <- norm_ci_plus |>
-  mutate(type = factor(c(rep("ci", 500), rep("samp", 16))))
+  mutate(type = factor(c(rep("ref", 5000), rep("samp", 16))))
 animate_xy(norm_ci_plus[,1:4], col=norm_ci_plus$type)
-
+ptcex <- ifelse(norm_plus$type == "samp", 2, 0.5)
+render_gif(norm_ci_plus[,1:4], 
+           planned_tour(tpath),
+           display_xy(col=norm_ci_plus$type,
+                      cex=ptcex,
+                      scale=TRUE, center=FALSE, 
+                      axes="bottomleft",
+                      half_range=5.2),
+           gif_file = "gifs/fig2a.gif",
+           frames = 500,
+           width = 400,
+           height = 400)
 ```
@@
 
-![Relative to confidence ellipse.](images/example1.png){#fig-compare1 width=250 align="center"}
+![Reference ellipse](images/example1.png){#fig-compare1 width=250 align="center"}
 
-![Relative to normal patients.](images/example2.png){#fig-compare width=250 align="center"}
+![Reference sample](images/example2.png){#fig-compare width=250 align="center"}
 
-Illustration of current procedure: (a) compare new sample with points of the surface of a $p\text{-}D$ ellipse, (b) compare new sample with a simulated sample of normal patients.
-Although both approaches are useful, the rough edges of the projected ellipse points makes it difficult to precisely assess the positions of the new sample against the normal bounds.
+Illustration of existing procedures for a $4\text{-}D$ problem: compare new sample with (a) points on the surface of a $p\text{-}D$ ellipse representing the reference distribution, (b) a sample of observations simulated from the reference distribution.
+A single $2\text{-}D$ projection from a tour.
+(The circle and projected axes represent projection coefficients, and animated gifs of these tours are available in the supplementary material.) Both approaches are useful, but need the refinements described in this paper.
 :::
 
-Although this is flexible, this does not make it easy to guide the tour towards the directions (projections) where the samples are most different from the normal.
-What would be desirable is to analytically define the confidence ellipsoid, flag observations that are outside, steer the tour to projections that reveal the extent of the difference.
-And also display the projected ellipsoid as a geometric shape rather than a sample of points.
-These are the procedures that are described in the next section.
+This approach for comparing new observations with a reference distribution can be applied to any \pD{} problem.
+@fig-compare illustrates the procedures for a \fD{} problem.
+The new sample of observations is compared against the reference distribution in two ways (a) against a \fD{} ellipse corresponding to a level curve of density function of the reference distribution and (b) against a sample simulated from the reference distribution.
+Here we show a single \gD{} projection from a tour of the full \pD{}.
+The circle and line segments indicate the coefficients of the $4\times 2$ basis used to project the data in these plots.
+The difference between the new sample of points (dark/red) from the sample is in the bottom left to top right direction, corresponding mostly to the V3 axis direction.
+Roughly orthogonal to this is a combination of V1 and V2, so the difference between new and reference as indicated by this projection is due to V4 contrasted with V1 and V2.
+The ellipse makes it clearer see but the sample can be useful, too, particularly if the reference distribution is not a multivariate distribution.
+(The supplementary material includes animated gifs showing these data in a tour, allowing one to see many different projections of the \fD{} points.)
+
+While this is a flexible and useful approach, there are two problems: (1) the ragged edges of the points marking the projected ellipse make it difficult to compare the new sample precisely against acceptable ranges, and (2) there is no projection pursuit index that can guide the tour towards the directions (projections) where the samples are most different from the reference.
+
+What would be better, and what is developed in this paper, is to analytically define the confidence ellipsoid and display the projected ellipsoid as a geometric shape rather than a sample of points.
+Further, a new projection pursuit index that is based on flagging observations that are outside, and would steer the tour to projections that reveal the extent of the difference.
+These new procedures are described in the next section.
 
@@
 
-This example is motivated by a problem posed during consulting with a pharmaceutical company, but the data shown here is simulated, simply to illustrate the application.
+This example is motivated by a problem posed during consulting with a pharmaceutical company.
+The data shown here is simulated, for privacy purposes, but reflects the patterns seen in the clinical data.
 Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).

### 2025-01-03 paper.tex
finished changes requested from me

--- 2024-12-11 paper.tex
@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
@@
 \newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
 \newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \newcommand\bD{$1\text{-}D$}
+\newcommand\fD{$4\text{-}D$}
 \newtheorem*{theorem}{Theorem}
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
 \usepackage{bookmark}
@@
 steered towards those where the new sample is most different from the
-reference. A new projection pursuit index is defined, and the drawing of
-the projected ellipse is computed analytically. The methods are
-implemented in the R package, \texttt{tourr}.
+reference. A new projection pursuit index is defined for this purpose.
+The tour visualisation also includes drawing of the projected ellipse,
+which is computed analytically, corresponding to the reference
+distribution. The methods are implemented in the R package,
+\texttt{tourr}.
 \end{abstract}
@@
 may describe experimental results, or a global fit for a selected model,
-and we might want to compare to a set of other models. In medical
+and we might want to compare to sets of other models. In medical
 applications, the normal distribution might summarize historic data of a
@@
 
-This paper describes a new projection pursuit index which is optimized
-by projections where a new sample is most distant from the existing
-normal distribution.
-\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
-However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures.
-Classical methods to detect outliers with projection pursuit are based on skewness (Loperfido) (even the classical indexes were supposed to detect outliers I think?)....
-also dobin package by Sevvandi might be relevant here.
-XXX when/why might it be better to use our method?}
-
-It is organised as follows. Section~\ref{sec-background} provides more
-context for the methods and visualisation.
+This paper describes a new projection pursuit index to find projections
+where a new sample is most distant from the existing normal
+distribution, that is integrated with the guided tour to visualise the
+ways that the data departs from the reference distribution. This work
+also expands on current outlier detection methods such as
+Kandanaarachchi and Hyndman (2021) and Loperfido (2018) by providing
+integrated multivariate visualisation to examine the outlyingness in
+different projections. It also provides the ability to examine potential
+outliers interactively - when they are outlying in different ways it can
+interfere with detection numerically, but visually one might group
+different types of outlyingness, and operate the outlier detection on
+subsets.
+
+The paper is organised as follows. Section~\ref{sec-background} provides
+more context for the methods and visualisation.
 Section~\ref{sec-anomaly-index} provides the details of the new index,
-and Section~\ref{sec-implementation} describes the implementation.
-Example use is illustrated in Section~\ref{sec-examples}.
+and rendering projections of the reference ellipse.
+Section~\ref{sec-implementation} describes the implementation. Two
+examples, one medical and one on climate extremes, illustrate usage in
+Section~\ref{sec-examples}.
 
@@
 
-To compare a new sample with an existing norm, like a multivariate
-normal distribution, in higher than two dimensions, we have typically
-used two samples of points. The norm is represented by points on the
-surface of a \(p\)-dimensional ellipsoid, corresponding to a confidence
-level. A sample of points on a \(p\)-dimensional ellipsoid is generated
-by
+Following the ideas expressed in Cook and Swayne (2007), to compare a
+new sample with an existing norm, like a multivariate normal
+distribution, in higher than two dimensions, you would use two samples
+of points. The norm is represented by points on the surface of a
+\(p\)-dimensional ellipsoid, with size corresponding to a chosen level
+curve of the density function of the reference distribution. This can be
+generated using these steps:
 
@@
 \item
-  Simulating a sample of observations (\(\mathbfit{x}\), which are \pD{}
+  Simulate a sample of observations (\(\mathbfit{x}\), which are \pD{}
   vectors) from \(N_p(\mathbfit{\mu}, \Sigma)\).
 \item
-  Transforming each observation to have unit distance from the mean,
+  Transform each observation to have unit distance from the mean,
   \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\). The sample is now
@@
   Transform the shape from a sphere to an ellipsoid using the specific
-  variance covariance, and shift to center on the mean vector.
+  variance-covariance.
+\item
+  Shift the resulting sample to center it on the mean vector, and resize
+  to correspond to the desired density level. This could also be treated
+  like a confidence ellipse, where observations outside are more than,
+  say, three standard deviations from the sample mean.
 \end{enumerate}
 
-Finally, new observations can be visually compared with this ellipsoid
-by
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
-\item
-  plotting them together.
-\end{enumerate}
-
-Figure~\ref{fig-ci} illustrates this process for \gD. This is easiest
-way to view this normal region relative to a new sample for any \pD{}
-problem.
+New observations can be visually compared with this ellipsoid by
+plotting them together using a tour. Figure~\ref{fig-ci} illustrates
+this process for \gD.
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-ci-1.pdf}
 
@@
 sampling from a multivariate normal (a) and transforming each
-observation to have length equal to 1 (b). A confidence ellipsoid is
-generated by transforming the sphere relative to a specified
-variance-covariance matrix (c), and new observations can be visually
-assessed to be inside or outside by plotting with the ellipsoid (d).}
+observation to have length equal to 1 (b). A 95\% confidence ellipsoid
+is generated by transforming the sphere relative to a specified
+variance-covariance matrix, and sizing using a critical value (c), and
+new observations (triangle, square) can be visually assessed to be
+inside or outside by plotting with the ellipsoid (d).}
 
@@
 
-For example, Figure~\ref{fig-compare} compares a new sample of patient
-scores against the normal range represented by an ellipse (a) and also
-against a simulated sample of normal patients (b). While these are
-useful approaches, the ragged edges of the projected ellipse make it
-difficult to compare the new sample precisely agains the normal ranges.
-
 \begin{figure}
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/example1.png}
-
-}
-
-\subcaption{\label{fig-compare1}Relative to confidence ellipse.}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example1.png}
+
+}
+
+\subcaption{\label{fig-compare1}Reference ellipse}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/example2.png}
-
-}
-
-\subcaption{\label{fig-compare}Relative to normal patients.}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example2.png}
+
+}
+
+\subcaption{\label{fig-compare}Reference sample}
 
@@
 
-\caption{\label{fig-compare}Illustration of current procedure: (a)
-compare new sample with points of the surface of a \(p\text{-}D\)
-ellipse, (b) compare new sample with a simulated sample of normal
-patients. Although both approaches are useful, the rough edges of the
-projected ellipse points makes it difficult to precisely assess the
-positions of the new sample against the normal bounds.}
+\caption{\label{fig-compare}Illustration of existing procedures for a
+\(4\text{-}D\) problem: compare new sample with (a) points on the
+surface of a \(p\text{-}D\) ellipse representing the reference
+distribution, (b) a sample of observations simulated from the reference
+distribution. A single \(2\text{-}D\) projection from a tour. (The
+circle and projected axes represent projection coefficients, and
+animated gifs of these tours are available in the supplementary
+material.) Both approaches are useful, but need the refinements
+described in this paper.}
 
@@
 
-Although this is flexible, this does not make it easy to guide the tour
+This approach for comparing new observations with a reference
+distribution can be applied to any \pD{} problem.
+Figure~\ref{fig-compare} illustrates the procedures for a \fD{} problem.
+The new sample of observations is compared against the reference
+distribution in two ways (a) against a \fD{} ellipse corresponding to a
+level curve of density function of the reference distribution and (b)
+against a sample simulated from the reference distribution. Here we show
+a single \gD{} projection from a tour of the full \pD{}. The circle and
+line segments indicate the coefficients of the \(4\times 2\) basis used
+to project the data in these plots. The difference between the new
+sample of points (dark/red) from the sample is in the bottom left to top
+right direction, corresponding mostly to the V3 axis direction. Roughly
+orthogonal to this is a combination of V1 and V2, so the difference
+between new and reference as indicated by this projection is due to V4
+contrasted with V1 and V2. The ellipse makes it clearer see but the
+sample can be useful, too, particularly if the reference distribution is
+not a multivariate distribution. (The supplementary material includes
+animated gifs showing these data in a tour, allowing one to see many
+different projections of the \fD{} points.)
+
+While this is a flexible and useful approach, there are two problems:
+(1) the ragged edges of the points marking the projected ellipse make it
+difficult to compare the new sample precisely against the normal ranges,
+and (2) there is no projection pursuit index that can guide the tour
 towards the directions (projections) where the samples are most
-different from the normal. What would be desirable is to analytically
-define the confidence ellipsoid, flag observations that are outside,
-steer the tour to projections that reveal the extent of the difference.
-And also display the projected ellipsoid as a geometric shape rather
-than a sample of points. These are the procedures that are described in
-the next section.
+different from the reference.
+
+What would be better, and what is developed in this paper, is to
+analytically define the confidence ellipsoid and display the projected
+ellipsoid as a geometric shape rather than a sample of points. Further,
+a new projection pursuit index that is based on flagging observations
+that are outside, and would steer the tour to projections that reveal
+the extent of the difference. These new procedures are described in the
+next section.
 
@@
 ways, but the default is those that are outside the specified ellipsoid
-in \pD{}. Alternatives could be to select a set of observations with the
-largest Mahalanobis distance, manually select observations or possibly a
-group of points identified via clustering of the extremes.
-
-The index is written as
+in \pD{},
+\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\).
+Alternatives could be to select a set of observations with the largest
+Mahalanobis distance, manually select observations or possibly a group
+of points identified via clustering of the extremes.
+
+We define our new index as
 
 \begin{equation}
-\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
+\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T.
 \end{equation}
-
-where by default
-\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\),
-is the set of observations outside the \pD{} ellipsoid.
 
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/anomaly1.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/anomaly2.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly2.png}
 
@@
 This example is motivated by a problem posed during consulting with a
-pharmaceutical company, but the data shown here is simulated, simply to
-illustrate the application. Liver function tests commonly provide
-measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase
-(GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and
-alanine aminotransferase (ALT). There are normal ranges on these
-measurements reported by Lala, Zubair, and Minter (2023) and listed in
-Table~\ref{tbl-liver-norms}.
+pharmaceutical company. The data shown here is simulated, for privacy
+purposes, but reflects the patterns seen in the clinical data. Liver
+function tests commonly provide measurements on albumin, protein,
+bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase
+(AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).
+There are normal ranges on these measurements reported by Lala, Zubair,
+and Minter (2023) and listed in Table~\ref{tbl-liver-norms}.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0421}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2000}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0421}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.2000}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}@{}}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver2.png}
 
@@
 longitudinal test results for a single patient. Red cross indicates
-observation is outide the 4-D confidence ellipse. The female patients
+observation is outside the 4-D confidence ellipse. The female patients
 tend to have lower ALP and ALT than the normal range. As the patient
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/weather1.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight]{images/weather2.png}
+\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather2.png}
 
@@
 
-\includegraphics[width=6.25in,height=\textheight]{images/weather3.png}
+\includegraphics[width=6.25in,height=\textheight,keepaspectratio]{images/weather3.png}
 
@@
 
+\bibitem[\citeproctext]{ref-CS07}
+Cook, Dianne, and Deborah F. Swayne. 2007. \emph{Interactive and Dynamic
+Graphics for Data Analysis: With {R} and {GGobi}}. Use r! New York:
+Springer-Verlag. \url{https://doi.org/10.1007/978-0-387-71762-3}.
+
 \bibitem[\citeproctext]{ref-donoho1982}
@@
 Society, Series A} 150: 1--36.
+
+\bibitem[\citeproctext]{ref-Kandanaarachchi02012021}
+Kandanaarachchi, Sevvandi, and Rob J. Hyndman. 2021. {``Dimension
+Reduction for Outlier Detection Using DOBIN.''} \emph{Journal of
+Computational and Graphical Statistics} 30 (1): 204--19.
+https://doi.org/\url{https://doi.org/10.1080/10618600.2020.1807353}.
 
@@
 
+\bibitem[\citeproctext]{ref-loperfido}
+Loperfido, Nicola. 2018. {``Skewness-Based Projection Pursuit: A
+Computational Approach.''} \emph{Computational Statistics \& Data
+Analysis} 120: 42--57.
+https://doi.org/\url{https://doi.org/10.1016/j.csda.2017.11.001}.
+
 \bibitem[\citeproctext]{ref-robustbase}

### 2025-01-03 review/reply.Rmd
finished changes requested from me

@@
 
-We thank the reviewers for their feedback on the paper. Since the article was submitted as a *short technical note* it is not possible to accomodate all suggestions.
+We thank the Associate editor and anonymous reviewers for the feedback on the paper.
+In discussion with the editor, it is agreed that becuase the article was submitted as a *short technical note* that the revision would focus on these changes only:
+
+1. Make all the small corrections listed by Reviewer 2.
+2. Check the computational details to ensure that it is clear.
+3. Check the literature references, to ensure nothing major is missed. The reviewers didn’t suggest articles that we had missed citing. We can try to make the context clearer.
+4. Ignore the request for a broader comparison or add more examples. The two examples provided are contemporary and differ in nature, illustrating general usage.
 
@@
 
-\textcolor{purple}{A comprehensive review would be beyond the scope of a short technical note, but we added additional context XXXX do this? how?}
+\textcolor{purple}{The introduction was slightly expanded, and two new references added to provide additional context.}
 
@@
 
-\textcolor{purple}{To the best of our knowledge, the proposed method is the first that allows the comparison of a reference normal distribution to observations using projection pursuit.}
+\textcolor{purple}{This is first work that allows the comparison of a reference normal distribution to observations using projection pursuit, and a projection pursuit guided tour.
+New text in the introduction helps clarify this.}
 
@@
 
-\textcolor{purple}{The second example in the paper is using real-world data, adding more examples would be beyond the scope of a short technical note.}
+\textcolor{purple}{The two examples used in the paper illustrate different uses of this work in practical settings.
+The first example uses simulated data for privacy purposes but the data reflects patterns seen in clinical data.
+The text has been modified to make this clear.
+There is no room in a short technical report to add more examples.}
 
@@
 
-\textcolor{purple}{The description is brief, fitting the style of a short technical note. XXX check where we could be more clear...}
+\textcolor{purple}{The introduction has been expanded slightly with two more references added to provide more context. XXX check where we could be more clear...}
 
@@
 
-\textcolor{purple}{XXX (Need to finish, add references) We have added a paragraph in the introduction discussing outlier detection with other projection pursuit index functions, including the most important references.}
+\textcolor{purple}{Done}
 
@@
 
-\textcolor{purple}{XXX I think the answer should be no, there is no comparison? Only rescaling for fitting into the plotting window which would be applied to both the nex points and the ellipsoid?}
+\textcolor{purple}{The new data needs to be pre-processed in the same manner as the reference distribution.
+For example, if the reference distribution is scaled using means and standard deviations of historical data, these same means and standard deviations need to be used to scale the new data.
+XXX Add some sentences to make this clear in the paper}
 
@@
 
-\textcolor{purple}{XXX todo @Di}
+\textcolor{purple}{The text in section 2 has been changed to remove any specialist language. The explanation is also more detailed now.}
 
@@
 
-\textcolor{purple}{XXX todo @Di}
+\textcolor{purple}{This figure has been modified to make it clearer, and to correspond with the changed text in the section.}
 
@@
 
-\textcolor{purple}{XXX todo @Di}
+\textcolor{purple}{Done}
 

### 2025-01-03 paper.qmd
copying edits from earlier

@@
 ### Projecting an ellipsoid
-
-Let $x$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by
+Let \(\mathbfit{x}\) be a \(p\) dimensional vector. A \pD{} ellipsoid corresponding to a given
+variance-covariance (\(\Sigma\)) is described by the equation
 
@@
 
-where $c$ a constant that depends on a specific confidence level.
+where \(c\) is a constant that depends on a specific confidence level.
 
+Let \(P\) be a \((p\times 2)\) matrix, whose columns form an orthonormal basis for the \gD{} projection space.
+Let \(\mathbfit{y}\) be a 2 dimensional vector, and let \(\mathbfit{\mu}_p := \mathbfit{\mu} P\) denote the projected mean.
 \begin{theorem}
-  The projection of this \pD{} ellipsoid in \gD{} has the equation
+  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by \(P\) has the equation
 \begin{equation}
@@
 \begin{proof}
-The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
-That is, the curve
-consists of $\mathbfit{x}$ satisfying
+The projection of a hyperellipsoid onto \gD{} space is an ellipse.
+The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid.
+For \(\mathbfit{x}\) on the hyperellipsoid, its orthogonal projection \(\mathbfit{x}P\) is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at \(\mathbfit{x}\) is perpendicular to the projection plane described by \(P\).
+As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at \(\mathbfit{x}\) is parallel to the projection plane.
+Note that any vector parallel to the projection plane can be expressed as \(2\mathbfit{s}P^T\) for some \(\mathbfit{s} \in \mathbb{R}^2\).
 
-\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+Therefore, the boundary of the ellipse consists of the points \(\mathbfit{x}P\) for which $\mathbfit{x}$ satisfies
+
+\begin{equation}
+2\mathbfit{s}P^T = \nabla \left[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T\right] = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}
+\label{eq:grad}
 \end{equation}
 
-for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
-projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+for some \(\mathbfit{s} \in \mathbb{R}^2\). Rearranging this equation gives \(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\).
 Making this substitution in Equation~(\ref{eq:pd}) yields
@@
 
-This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection.
-In general $c$ could be any constant, but typically we would select it as a quantile of the $\chi^2$ distribution, so that the size of the ellipse corresponds to a selected probability.
+This means the matrix \((P^T \Sigma P)^{-1}\) defines the ellipse in
+the \gD{} projection. In general \(c\) could be any constant, but
+typically we would select it as a quantile of the \(\chi^2\)
+distribution, so that the size of the ellipse corresponds to a selected
+probability.
 

### 2025-01-03 review/reply.Rmd
my responses

@@
 
-\textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow. XXX to check @Annalisa}
+\textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow.
+More detail has been added to the theorem and proof in section 3.}
 
@@
 
-\textcolor{purple}{We have numbered the main equations and now use referencing in the description. XXX check flow of content @Annalisa}
+\textcolor{purple}{We have numbered the main equations and now use referencing in the description.
+Segments of this section have been rewritten to include more explanation and enhance flow, and variables have been defined earlier, before they are used.}
 
@@
 
-\textcolor{purple}{XXX todo @Annalisa}
+\textcolor{purple}{The ellipsoid equation is now specifically referenced in the statement of the theorem.
+The mentioned variables are now defined before the statement of the theorem, helping reorganise the section to improve clarity.
+Although the proof is not fundamentally restructured, it has been reworded with further explanation added.}
 

### 2025-01-03 paper.qmd
checked and smoothed changes made by Annalisa

--- 2025-01-03 paper.qmd
@@
 ### Projecting an ellipsoid
-
-Let $x$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by
+Let $\mathbfit{x}$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given
+variance-covariance ($\Sigma$) is described by the equation
 
@@
 
-where $c$ a constant that depends on a specific confidence level.
+where $c$ is a constant that depends on a specific confidence level.
 
+Let $P$ be a ($p\times 2$) matrix, whose columns form an orthonormal basis for the \gD{} projection space.
+Let $\mathbfit{y}$ be a \gD{} vector, and let $\mathbfit{\mu}_p := \mathbfit{\mu} P$ denote the projected mean.
 \begin{theorem}
-  The projection of this \pD{} ellipsoid in \gD{} has the equation
+  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
 \begin{equation}
@@
 \begin{proof}
-The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
-That is, the curve
-consists of $\mathbfit{x}$ satisfying
+The projection of a hyperellipsoid onto \gD{} space is an ellipse.
+The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid.
+For $\mathbfit{x}$ on the hyperellipsoid, its orthogonal projection $\mathbfit{x}P$ is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at $\mathbfit{x}$ is perpendicular to the projection plane described by $P$.
+As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at $\mathbfit{x}$ is parallel to the projection plane.
+Note that any vector parallel to the projection plane can be expressed as $2\mathbfit{s}P^T$ for some $\mathbfit{s} \in \mathbb{R}^2$.
 
-\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+Therefore, the boundary of the ellipse consists of the points $\mathbfit{x}P$ for which $\mathbfit{x}$ satisfies
+
+\begin{equation}
+2\mathbfit{s}P^T = \nabla \left[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T\right] = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}
+\label{eq:grad}
 \end{equation}
 
-for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
-projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+for some $\mathbfit{s} \in \mathbb{R}^2$. Rearranging this equation gives $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
 Making this substitution in Equation~(\ref{eq:pd}) yields
@@
 
-We call points in the projection that are on the curve $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for $\mathbfit{x}$ on the \pD{} curve.
-Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean.
+We denote points on the boundary of the projected ellipse by $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for some $\mathbfit{x}$ satifying Equation~(\ref{eq:grad}).
+Then, as $\mathbfit{\mu}_p = \mathbfit{\mu} P$ by definition, $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, again substituting $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
 We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in Equation~(\ref{eq:c2}).
@@
 
-This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection.
-In general $c$ could be any constant, but typically we would select it as a quantile of the $\chi^2$ distribution, so that the size of the ellipse corresponds to a selected probability.
+This means the matrix $(P^T \Sigma P)^{-1}$ defines the ellipse in
+the \gD{} projection. In general $c$ could be any constant, but
+typically we would select it as a quantile of the $\chi^2$
+distribution, so that the size of the ellipse corresponds to a selected
+probability.
 
@@
 
-![](images/weather3.png){#fig-weather3 width=600 align="center"}
+![](images/weather3.png){#fig-weather3 width=550 align="center"}
 

### 2025-01-03 paper.tex
checked and smoothed changes made by Annalisa

@@
 (1) the ragged edges of the points marking the projected ellipse make it
-difficult to compare the new sample precisely against the normal ranges,
+difficult to compare the new sample precisely against acceptable ranges,
 and (2) there is no projection pursuit index that can guide the tour
@@
 
-Let \(x\) be a \pD{} vector. A \pD{} ellipsoid corresponding to a given
-variance-covariance (\(\Sigma\)) is defined by
+Let \(\mathbfit{x}\) be a \pD{} vector. A \pD{} ellipsoid corresponding
+to a given variance-covariance (\(\Sigma\)) is described by the equation
 
@@
 
-where \(c\) a constant that depends on a specific confidence level.
+where \(c\) is a constant that depends on a specific confidence level.
+
+Let \(P\) be a (\(p\times 2\)) matrix, whose columns form an orthonormal
+basis for the \gD{} projection space. Let \(\mathbfit{y}\) be a \gD{}
+vector, and let \(\mathbfit{\mu}_p := \mathbfit{\mu} P\) denote the
+projected mean.
 
 \begin{theorem}
-  The projection of this \pD{} ellipsoid in \gD{} has the equation
+  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
 \begin{equation}
@@
 \begin{proof}
-The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane.
-That is, the curve
-consists of $\mathbfit{x}$ satisfying
-
-\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
+The projection of a hyperellipsoid onto \gD{} space is an ellipse.
+The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid.
+For $\mathbfit{x}$ on the hyperellipsoid, its orthogonal projection $\mathbfit{x}P$ is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at $\mathbfit{x}$ is perpendicular to the projection plane described by $P$.
+As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at $\mathbfit{x}$ is parallel to the projection plane.
+Note that any vector parallel to the projection plane can be expressed as $2\mathbfit{s}P^T$ for some $\mathbfit{s} \in \mathbb{R}^2$.
+
+Therefore, the boundary of the ellipse consists of the points $\mathbfit{x}P$ for which $\mathbfit{x}$ satisfies
+
+\begin{equation}
+2\mathbfit{s}P^T = \nabla \left[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T\right] = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}
+\label{eq:grad}
 \end{equation}
 
-for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
-projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
+for some $\mathbfit{s} \in \mathbb{R}^2$. Rearranging this equation gives $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
 Making this substitution in Equation~(\ref{eq:pd}) yields
@@
 
-This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
-the \gD{} projection. In general \(c\) could be any constant, but
-typically we would select it as a quantile of the \(\chi^2\)
-distribution, so that the size of the ellipse corresponds to a selected
-probability.
+This means the matrix \((P^T \Sigma P)^{-1}\) defines the ellipse in the
+\gD{} projection. In general \(c\) could be any constant, but typically
+we would select it as a quantile of the \(\chi^2\) distribution, so that
+the size of the ellipse corresponds to a selected probability.
 
@@
 
-\includegraphics[width=6.25in,height=\textheight,keepaspectratio]{images/weather3.png}
+\includegraphics[width=5.72917in,height=\textheight,keepaspectratio]{images/weather3.png}
 

### 2025-01-31 paper.qmd
Working on Fig 5 and 6 + other small changes

@@
 Roughly orthogonal to this is a combination of V1 and V2, so the difference between new and reference as indicated by this projection is due to V4 contrasted with V1 and V2.
-The ellipse makes it clearer see but the sample can be useful, too, particularly if the reference distribution is not a multivariate distribution.
+The ellipse makes it clearer see but the sample can be useful, too, particularly if the reference distribution is not a multivariate normal distribution.
 (The supplementary material includes animated gifs showing these data in a tour, allowing one to see many different projections of the \fD{} points.)
@@
 Let $\mathbfit{x}$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given
-variance-covariance ($\Sigma$) is described by the equation
+variance-covariance ($\Sigma$) and mean vector ($\mathbfit{\mu}$) is described by the equation
 
@@
 The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance.
-We use the Dunn index @dunn, computed using @fpc, to select the preferred number of clusters, $k=5$.
+We use the Dunn index [@dunn], computed using @fpc, to select the preferred number of clusters, $k=5$.
 
-The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see @fig-weather (a).
+The new anomaly index is first applied to the full dataset, such that the final projection will be found by averaging distance of points in many directions, see @fig-weather (a).
 It provides a global picture showing where the outlying points differ from the normal distribution.
@@
            col = as.factor(cluster_labels),
-           palette = "Purple-Yellow",
+           palette = "Batlow",
            center = FALSE, rescale = FALSE,
@@
 # color codes
-# "#80146E" "#7262B2" "#31A0C2" "#5ECBBB" "#B9E5B9"
-# "#F5F2D8"
+#"#201158" "#005260" "#007C46" "#999600" "#F9A391" "#FFCEF4"
+# "#005260"
 # wait for getting random starting planes for
@@
            axes="bottomleft",
-           col = "#5ECBBB", half_range = 7, cex = 2,
+           col = "#999600", half_range = 7, cex = 2,
            center = FALSE, rescale = FALSE,
@@
 
-![All outlying points clustered](images/weather1.png){#fig-weather1 width=300 align="center"}
+![All outlying points clustered](images/weather1.png){#fig-weather1 width=275 align="center"}
 
-![Only cluster 4](images/weather2.png){#fig-weather2 width=300 align="center"}
+![Only cluster 4](images/weather2.png){#fig-weather2 width=275 align="center"}
 
@@
 # following the package vignette to calculate SCD
+# abbreviate(str_remove_all(colnames(X), "_"), 3, named = F)
 weather_SCD <- SCD(x = X, mu, Sigma, Sigma_inv, step_size = 0.1, min_deviation = 0.2)
@@
                           color = as.factor(value)),
-            size = 2, fontface = "bold",
+            size = 3, fontface = "bold",
             inherit.aes = F) +
-  scale_color_manual(values = c("#80146E", "#7262B2", "#31A0C2", "#5ECBBB", "#B9E5B9")) +
-  theme(legend.position = "none",
+  scale_color_manual(values = c("#201158", "#005260", "#007C46", "#999600", "#F9A391"),
+                     guide = "none") +
+  theme(
+    #legend.position = "none",
         aspect.ratio = 0.5,

### 2025-01-31 paper.tex
Working on Fig 5 and 6 + other small changes

@@
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
 
@@
 \newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
@@
 \newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
 \newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
@@
 \makeatother
-
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage{bookmark}
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-ci-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example1.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/example1.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example2.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/example2.png}
 
@@
 sample can be useful, too, particularly if the reference distribution is
-not a multivariate distribution. (The supplementary material includes
-animated gifs showing these data in a tour, allowing one to see many
-different projections of the \fD{} points.)
+not a multivariate normal distribution. (The supplementary material
+includes animated gifs showing these data in a tour, allowing one to see
+many different projections of the \fD{} points.)
 
@@
 Let \(\mathbfit{x}\) be a \pD{} vector. A \pD{} ellipsoid corresponding
-to a given variance-covariance (\(\Sigma\)) is described by the equation
+to a given variance-covariance (\(\Sigma\)) and mean vector
+(\(\mathbfit{\mu}\)) is described by the equation
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly1.png}
+\includegraphics[width=3.125in,height=\textheight]{images/anomaly1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly2.png}
+\includegraphics[width=3.125in,height=\textheight]{images/anomaly2.png}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0421}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.2000}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0421}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2000}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}@{}}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver1.png}
+\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver2.png}
+\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}
 
@@
 normalizing observations to have length \(1\) and then apply k-means
-clustering with Euclidean distance. We use the Dunn index Halkidi,
-Batistakis, and Vazirgiannis (2002), computed using Hennig (2024), to
+clustering with Euclidean distance. We use the Dunn index (Halkidi,
+Batistakis, and Vazirgiannis 2002), computed using Hennig (2024), to
 select the preferred number of clusters, \(k=5\).
@@
 The new anomaly index is first applied to the full dataset, such that
-the final projection will be affected by averaging distance of points in
+the final projection will be found by averaging distance of points in
 many directions, see Figure~\ref{fig-weather} (a). It provides a global
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather1.png}
+\includegraphics[width=2.86458in,height=\textheight]{images/weather1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather2.png}
+\includegraphics[width=2.86458in,height=\textheight]{images/weather2.png}
 
@@
 
-\includegraphics[width=5.72917in,height=\textheight,keepaspectratio]{images/weather3.png}
+\includegraphics[width=5.72917in,height=\textheight]{images/weather3.png}
 

### 2025-01-31 review/reply.Rmd
Working on Fig 5 and 6 + other small changes

@@
 
-\textcolor{purple}{Done}
+\textcolor{purple}{Done XXX mention what was changed}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{A color legend has been added in Figure 5a.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{We have tried a few palettes and changed the color palette to "Bartlow".
+Note here that "cluster" 0 are points that are not tagged as outliers, this is why we use a lighter color to de-emphasize those points in the image.
+The color legend has been added to the Figure.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{The colors are matched between Figure 5 and 6.
+We have increased the size of the cluster labels in Figure 6 to make them more readable, and have added a color legend.
+XXX RE variable names}
 

### 2025-01-31 paper.qmd
fig 5 re-done

@@
                          levels = c(1:5, 0))
+ptshape <- factor(ifelse(cluster_labels_all$cluster>0, "out", "in"))
 
@@
            col = as.factor(cluster_labels),
-           palette = "Batlow",
+           pch = ptshape,
+           palette = c("#040404", "#55125D", "#A01975", "#E45358", "#F7AA2D", "#559999"),
+           shapeset = c(1, 16),
            center = FALSE, rescale = FALSE,
@@
            axes="bottomleft",
-           col = "#999600", half_range = 7, cex = 2,
+           col = "#E45358", pch=16, half_range = 7, cex = 2,
            center = FALSE, rescale = FALSE,
@@
 
-![Only cluster 4](images/weather2.png){#fig-weather2 width=275 align="center"}
+![Focus on cluster 4](images/weather2.png){#fig-weather2 width=275 align="center"}
 
-Examining weather anomalies using projections identified by the anomaly index: (a) computed on all outlying points, (b) computed only for cluster 4.
-In (a) the outlying points are somewhat orthogonal to the normal ellipse, which is primarily using the temperature min (`atmn`) and max (`atmx`) variables.
-In (b) the two observations are primarily extreme in the minimum temperature (`atmn`) variable, and have lower than normal minima.
+Examining weather anomalies using the anomaly index guided tour to assess what combinations of variables are creating the extremes.
+Anomalies were grouped based on similarity in direction, and colour indicates cluster membership.
+In (a) the anomaly index was computed on all outlying points, and most clusters (except for cluster 4) are outlying points are somewhat orthogonal to the normal ellipse.
+This extremeness is primarily due to temperature min (`atmn`) and max (`atmx`) variables as read from the axes.
+In (b) only the two points in cluster 4 are used for computing the index.
+These two observations are primarily extreme in the minimum temperature (`atmn`) variable, and have lower than normal minima.
 :::

### 2025-01-31 paper.tex
fig 5 re-done

--- 2025-01-03 paper.tex
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather1.png}
+\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather2.png}
-
-}
-
-\subcaption{\label{fig-weather2}Only cluster 4}
+\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather2.png}
+
+}
+
+\subcaption{\label{fig-weather2}Focus on cluster 4}
 
@@
 
-\caption{\label{fig-weather}Examining weather anomalies using
-projections identified by the anomaly index: (a) computed on all
-outlying points, (b) computed only for cluster 4. In (a) the outlying
-points are somewhat orthogonal to the normal ellipse, which is primarily
-using the temperature min (\texttt{atmn}) and max (\texttt{atmx})
-variables. In (b) the two observations are primarily extreme in the
-minimum temperature (\texttt{atmn}) variable, and have lower than normal
+\caption{\label{fig-weather}Examining weather anomalies using the
+anomaly index guided tour to assess what combinations of variables are
+creating the extremes. Anomalies were grouped based on similarity in
+direction, and colour indicates cluster membership. In (a) the anomaly
+index was computed on all outlying points, and most clusters (except for
+cluster 4) are outlying points are somewhat orthogonal to the normal
+ellipse. This extremeness is primarily due to temperature min
+(\texttt{atmn}) and max (\texttt{atmx}) variables as read from the axes.
+In (b) only the two points in cluster 4 are used for computing the
+index. These two observations are primarily extreme in the minimum
+temperature (\texttt{atmn}) variable, and have lower than normal
 minima.}

### 2025-01-31 paper.qmd
fig 6 fixed

@@
   add_column(variable = "num_precp_01")
+colnames(weather_SCD_rescaled$phi) <-
+  c("atmax", "atmin", 
+         "nsumm",
+         "nheat", "p",
+         "sunh",
+         "nclear",
+         "ncloud",
+         "rh",
+         "rhmax" ,
+         "rhmin",
+         "vwind",
+         "vw60",
+         "vwmax",
+         "prec",
+         "nprec")
+cluster_df$variable <- "nprec"
 p <- plot(weather_SCD_rescaled, type = "cell", n_digits = 0, continuous_rowname = 10, rotate_x = FALSE) +
@@
             inherit.aes = F) +
-  scale_color_manual(values = c("#201158", "#005260", "#007C46", "#999600", "#F9A391"),
+  scale_color_manual(values = c("#040404", "#55125D", "#A01975", "#E45358", "#F7AA2D"),
+                       #c("#201158", "#005260", "#007C46", "#999600", "#F9A391"),
                      guide = "none") +

### 2025-02-01 paper.qmd
Fixed labels and cluster label position in Fig 6

--- 2025-01-31 paper.qmd
@@
 cluster_df <- as_tibble(X_km$cluster, rownames = "year") %>%
-  add_column(variable = "num_precp_01")
+  add_column(variable = "clust")
+weather_SCD_rescaled$phi <- cbind((rep(0, nrow(weather_SCD_rescaled$phi))), weather_SCD_rescaled$phi)
+weather_SCD_rescaled$x_original <- cbind((rep(0, nrow(weather_SCD_rescaled$x_original))), weather_SCD_rescaled$x_original)
+colnames(weather_SCD_rescaled$phi) <-
+  c(".", "atmx", "atmn", 
+         "nsm",
+         "nheat", "p",
+         "sunh",
+         "nclear",
+         "ncloud",
+         "rh",
+         "rhmx" ,
+         "rhmn",
+         "vwind",
+         "vw60",
+         "vwmx",
+         "prc",
+         "nprec")
+
+cluster_df$variable <- "."
 p <- plot(weather_SCD_rescaled, type = "cell", n_digits = 0, continuous_rowname = 10, rotate_x = FALSE) +
@@
 Red cells indicate values higher than expected, blue cells values lower than expected.
-For years that have been identified as outliers in our approach we show the cluster assignment at the top.
+For years that have been identified as outliers in our approach we show the cluster assignment in the last row.
 :::

### 2025-02-01 paper.tex
Fixed labels and cluster label position in Fig 6

--- 2025-01-31 paper.tex
@@
 
-\subcaption{\label{fig-weather2}Only cluster 4}
+\subcaption{\label{fig-weather2}Focus on cluster 4}
 
@@
 years that have been identified as outliers in our approach we show the
-cluster assignment at the top.}
+cluster assignment in the last row.}
 

### 2025-02-01 paper.qmd
Completed revision.

@@
 
-New observations can be visually compared with this ellipsoid by plotting them together using a tour. @fig-ci illustrates this process for \gD.
+New observations can be visually compared with this ellipsoid by plotting them together using a tour, making sure that any transformations used on the reference distribution are also applied during pre-processing.
+@fig-ci illustrates this process for \gD.
 
@@
 
-To define a measure of an interesting projection is to maximize the average Mahalanobis distance [@mahalanobis] in the projection for a subset of points, $W$.
+Mahalanobis distance [@mahalanobis] measures the distance from the center $\mathbfit{\mu}$ in units of standard deviations, and is computed using the variance-covariance matrix $\Sigma$.
+
+\begin{equation}
+d_M (\mathbfit{x}) = \sqrt{(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T}
+\end{equation}
+
+To define a measure of an interesting projection we propose to use the average Mahalanobis distance in the projection, for a subset of points, $W$.
 The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}, $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$.
@@
 One could consider clustering these observations using angular distance to find groups of observations that have similar directions of departure.
+This is illustrated in the second example, see Section @sec-weather.
 
@@
 This is implemented in the `tourr` [@tourr;@tourr-cran] package, where the projected ellipsoid can be drawn for each projection.
-The guided tour will take arguments specifying the data, and the null variance-covariance matrix.
+The guided tour will take arguments specifying the data, and the null variance-covariance matrix, as shown in the example code.
 
@@
 
+A random projection as well as the final view after optimization are shown in Figure @fig-anomaly.
+The optimal projection (b) can be used to understand what combination of variables is driving the difference from the reference distribution.
+
 ::: {#fig-anomaly layout-ncol=2}
@@
 :::
+
+To optimize the index any optimization algorithm implemented in the `tourr` package can be used.
+The focus here is on the final projection identified, and direct interpretation of index values is not considered.
+It may however be interesting to compare different optimization strategies and compare the index values along the optimization paths, as available in @ferrn.
 
@@
 
+::: {#fig-liver layout-ncol=2}
+
+![Sample of female patients](images/liver1.png){#fig-liver1 width=300 align="center"}
+
+![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=300 align="center"}
+
+Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient.
+Red cross indicates observation is outside the 4-D confidence ellipse.
+The female patients tend to have lower ALP and ALT than the normal range.
+As the patient ages, the level of ALP increases and ALT decreases.
+:::
+
 The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70.
@@
 
-::: {#fig-liver layout-ncol=2}
-
-![Sample of female patients](images/liver1.png){#fig-liver1 width=300 align="center"}
-
-![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=300 align="center"}
-
-Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient.
-Red cross indicates observation is outside the 4-D confidence ellipse.
-The female patients tend to have lower ALP and ALT than the normal range.
-As the patient ages, the level of ALP increases and ALT decreases.
-:::
-
-### Robust statistics: weather extremes
+### Robust statistics: weather extremes {#sec-weather}
 
@@
 
-For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables.
-Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.
+For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature (atmn), while there is disagreement between outlyingness in other variables.
+Note that compared to the years 1965 and 1966 they have unusually low minimum temperature (atmn) while having average values for the number of days with maximum temperature above 25°C (nsm), and this contrast is also captured in the projection.
 
-When looking at the overall picture (@fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions.
-The direction of cluster 1 is where the average maximum temperature is unusually higher compared to the minimum temperature, while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature.
-This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum and the maximum temperature.
+When looking at the overall picture (@fig-weather (a)), we can see that the linear projection is showing the difference between minimum (atmn) and maximum temperature (atmx), which is where clusters are outlying in different directions.
+The direction of cluster 1 is where the average maximum temperature (atmx) is unusually higher compared to the minimum temperature (atmn), while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature.
+This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum (atmn) and the maximum temperature (atmx).
 
@@
 
-Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm.
+Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm [@MAYRHOFER2023].
 Red cells indicate values higher than expected, blue cells values lower than expected.

### 2025-02-01 paper.tex
Completed revision.

@@
 New observations can be visually compared with this ellipsoid by
-plotting them together using a tour. Figure~\ref{fig-ci} illustrates
-this process for \gD.
+plotting them together using a tour, making sure that any
+transformations used on the reference distribution are also applied
+during pre-processing. Figure~\ref{fig-ci} illustrates this process for
+\gD.
 
@@
 
-To define a measure of an interesting projection is to maximize the
-average Mahalanobis distance (Mahalanobis 1936) in the projection for a
-subset of points, \(W\). The set of points could be chosen in different
-ways, but the default is those that are outside the specified ellipsoid
-in \pD{},
+Mahalanobis distance (Mahalanobis 1936) measures the distance from the
+center \(\mathbfit{\mu}\) in units of standard deviations, and is
+computed using the variance-covariance matrix \(\Sigma\).
+
+\begin{equation}
+d_M (\mathbfit{x}) = \sqrt{(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T}
+\end{equation}
+
+To define a measure of an interesting projection we propose to use the
+average Mahalanobis distance in the projection, for a subset of points,
+\(W\). The set of points could be chosen in different ways, but the
+default is those that are outside the specified ellipsoid in \pD{},
 \(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\).
@@
 consider clustering these observations using angular distance to find
-groups of observations that have similar directions of departure.
+groups of observations that have similar directions of departure. This
+is illustrated in the second example, see Section
+Section~\ref{sec-weather}.
 
@@
 The guided tour will take arguments specifying the data, and the null
-variance-covariance matrix.
+variance-covariance matrix, as shown in the example code.
 
@@
 
+A random projection as well as the final view after optimization are
+shown in Figure Figure~\ref{fig-anomaly}. The optimal projection (b) can
+be used to understand what combination of variables is driving the
+difference from the reference distribution.
+
 \begin{figure}
@@
 \end{figure}%
+
+To optimize the index any optimization algorithm implemented in the
+\texttt{tourr} package can be used. The focus here is on the final
+projection identified, and direct interpretation of index values is not
+considered. It may however be interesting to compare different
+optimization strategies and compare the index values along the
+optimization paths, as available in Zhang et al. (2021).
 
@@
 
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}
+
+}
+
+\subcaption{\label{fig-liver1}Sample of female patients}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}
+
+}
+
+\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient}
+
+\end{minipage}%
+
+\caption{\label{fig-liver}Two projections of simulated example data
+corresponding to common liver tests: (a) sample of female patients, (b)
+longitudinal test results for a single patient. Red cross indicates
+observation is outside the 4-D confidence ellipse. The female patients
+tend to have lower ALP and ALT than the normal range. As the patient
+ages, the level of ALP increases and ALT decreases.}
+
+\end{figure}%
+
 The second example shows a longitudinal record of a single patient,
@@
 
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}
-
-}
-
-\subcaption{\label{fig-liver1}Sample of female patients}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}
-
-}
-
-\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient}
-
-\end{minipage}%
-
-\caption{\label{fig-liver}Two projections of simulated example data
-corresponding to common liver tests: (a) sample of female patients, (b)
-longitudinal test results for a single patient. Red cross indicates
-observation is outside the 4-D confidence ellipse. The female patients
-tend to have lower ALP and ALT than the normal range. As the patient
-ages, the level of ALP increases and ALT decreases.}
-
-\end{figure}%
-
-\subsection{Robust statistics: weather
-extremes}\label{robust-statistics-weather-extremes}
+\subsection{Robust statistics: weather extremes}\label{sec-weather}
 
@@
 for these two years the result from Mayrhofer and Filzmoser (2023) also
-indicates unusually low values for the minimum teperature, while there
-is disagreement between outlyingness in other variables. Note that
+indicates unusually low values for the minimum teperature (atmn), while
+there is disagreement between outlyingness in other variables. Note that
 compared to the years 1965 and 1966 they have unusually low minimum
-temperature while having average values for the number of days with
-maximum temperature above 25°C, and this contrast is also captured in
-the projection.
+temperature (atmn) while having average values for the number of days
+with maximum temperature above 25°C (nsm), and this contrast is also
+captured in the projection.
 
@@
 can see that the linear projection is showing the difference between
-minimum and maximum temperature, which is where clusters are outlying in
-different directions. The direction of cluster 1 is where the average
-maximum temperature is unusually higher compared to the minimum
-temperature, while clusters 3 corresponds to years where the minimum
-temperature is unusually close to the maximum temperature. This combined
-information cannot be resolved with the cell detection algorithm, but we
-see that for example the most recent years (cluster 3) were tagged for
-having unusually high values for both the minimum and the maximum
-temperature.
+minimum (atmn) and maximum temperature (atmx), which is where clusters
+are outlying in different directions. The direction of cluster 1 is
+where the average maximum temperature (atmx) is unusually higher
+compared to the minimum temperature (atmn), while clusters 3 corresponds
+to years where the minimum temperature is unusually close to the maximum
+temperature. This combined information cannot be resolved with the cell
+detection algorithm, but we see that for example the most recent years
+(cluster 3) were tagged for having unusually high values for both the
+minimum (atmn) and the maximum temperature (atmx).
 
@@
 and Filzmoser (2023), showing which cells have been identified as
-outlying by their Shapley Cell Detector algorithm. Red cells indicate
-values higher than expected, blue cells values lower than expected. For
-years that have been identified as outliers in our approach we show the
-cluster assignment in the last row.}
+outlying by their Shapley Cell Detector algorithm (Mayrhofer and
+Filzmoser 2023). Red cells indicate values higher than expected, blue
+cells values lower than expected. For years that have been identified as
+outliers in our approach we show the cluster assignment in the last
+row.}
 
@@
 
+\bibitem[\citeproctext]{ref-ferrn}
+Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
+Patricia Menéndez. 2021. {``Visual Diagnostics for Constrained
+Optimisation with Application to Guided Tours.''} \emph{The R Journal}
+13 (2): 624--41. \url{https://doi.org/10.32614/RJ-2021-105}.
+
 \end{CSLReferences}

### 2025-02-01 review/reply.Rmd
Completed revision.

--- 2025-01-03 review/reply.Rmd
@@
 
-\textcolor{purple}{XXXX need to add something? There is also the implementation and its documentation we could point to...}
+\textcolor{purple}{We have extended the implementation section to mention the options for optimization of the index, and have included a reference for the ferrn package that can be used to investigate index values and to compare different optimization strategies.
+Including such an investigation would be beyond the scope of a short technical note.
+All implementations are available in CRAN packages and the code to reproduce the examples are available on GitHub for full reproducibility.}
 
@@
 
-\textcolor{purple}{The introduction has been expanded slightly with two more references added to provide more context. XXX check where we could be more clear...}
+\textcolor{purple}{The introduction has been expanded slightly with two more references added to provide more context.}
 
 \textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow.
-More detail has been added to the theorem and proof in section 3.}
+More detail has been added to the theorem and proof in Section 3.}
 
@@
 
-\textcolor{purple}{Done}
+\textcolor{purple}{We have additional context and references in the introduction.}
 
@@
 
-\textcolor{purple}{The new data needs to be pre-processed in the same manner as the reference distribution.
-For example, if the reference distribution is scaled using means and standard deviations of historical data, these same means and standard deviations need to be used to scale the new data.
-XXX Add some sentences to make this clear in the paper}
+\textcolor{purple}{The new data needs to be pre-processed in the same manner as the reference distribution, we have added this information in the description.
+For example, if the reference distribution is scaled using means and standard deviations of historical data, these same means and standard deviations need to be used to scale the new data.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
-\textcolor{purple}{XXX I think we actually implicitly use robust version, since we use robust estimation when getting the covariance for the data in the second example}
+\textcolor{purple}{We have rephrased the sentence and included a definition of Mahalanobis distance.}
+\textcolor{purple}{Robust methods are used when estimating a reference distribution from a data sample containing outliers, thus implicitly we are using a robust version of the distance.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula (this is done in the second example, will mention it, not add simulations)}
+\textcolor{purple}{Including a detailed simulation study is beyond the scope of a short technical note.
+However, the second example shows how clustering can be used in an application.
+This is now mentioned in Section 3.3.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{We have emphasized that point in the text referring to Figure 3.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula (spelling is fixed, check position when other changes are done)}
+\textcolor{purple}{We have moved Figure 4 up (now page 12), and fixed the spelling.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{We have updated the color palette and now use shape to de-emphasize points that are not outlying (cluster 0). The color legend has been added to the Figure.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula}
+\textcolor{purple}{The colors are matched between Figure 5 and 6.
+We have increased the size of the cluster labels in Figure 6 to make them more readable, and have added a color legend.
+Variable nams in Figure 6 have been matched to those in Figure 5, and are now always added in parentheses when mentioned in the text.}
 
@@
 
-\textcolor{purple}{XXX todo @Ursula (it is, check if this needs to be made more clear in the paper)}
+\textcolor{purple}{It is, we have added the reference also to where the algorithm is mentioned to make this more clear.}
 
@@
 
-\textcolor{purple}{XXX todo (check with what we change in the introduction)}
+\textcolor{purple}{We have extended the introduction somewhat to include references to outlier detection in the context of linear dimension reduction.
+The additional references in the conclusions show potential generalizations of the work connecting to other directionsin outlier detection.}

### 2025-02-01 paper.qmd
spell checked

@@
 It also provides the ability to examine potential outliers interactively - when they are outlying in different ways it can interfere with detection numerically, but visually one might group different types of outlyingness, and operate the outlier detection on subsets.
-<!--\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution.
-However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures.
-Classical methods to detect outliers with projection pursuit are based on skewness (Loperfido) (even the classical indexes were supposed to detect outliers I think?)....
-also dobin package by Sevvandi might be relevant here.
-XXX when/why might it be better to use our method?}-->
 
@@
 \begin{theorem}
-  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
+  The projection of the \pD{} ellipsoid described in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
 \begin{equation}
@@
 
-We denote points on the boundary of the projected ellipse by $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for some $\mathbfit{x}$ satifying Equation~(\ref{eq:grad}).
+We denote points on the boundary of the projected ellipse by $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for some $\mathbfit{x}$ satisfying Equation~(\ref{eq:grad}).
 Then, as $\mathbfit{\mu}_p = \mathbfit{\mu} P$ by definition, $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, again substituting $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
@@
 The data shown here is simulated, for privacy purposes, but reflects the patterns seen in the clinical data.
-Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).
+Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl-transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).
 There are normal ranges on these measurements reported by @lib-med-liver-norms and listed in @tbl-liver-norms.
@@
 
-For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature (atmn), while there is disagreement between outlyingness in other variables.
+For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum temperature (atmn), while there is disagreement between outlyingness in other variables.
 Note that compared to the years 1965 and 1966 they have unusually low minimum temperature (atmn) while having average values for the number of days with maximum temperature above 25°C (nsm), and this contrast is also captured in the projection.
@@
 Red cells indicate values higher than expected, blue cells values lower than expected.
-For years that have been identified as outliers in our approach we show the cluster assignment in the last row.
+For years that have been identified as outliers in our approach we show the cluster assignment in the bottom row.
 :::
@@
 This paper has provided a new projection pursuit index for comparing a sample of data against a multivariate normal reference distribution.
-It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corrsponding to the normal reference distribution.
+It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corresponding to the normal reference distribution.
 These combined provide new ways to visualise multivariate data, for this particular scenario.

### 2025-02-01 paper.tex
spell checked

@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/example1.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example1.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight]{images/example2.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/example2.png}
 
@@
 function tests commonly provide measurements on albumin, protein,
-bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase
+bilirubin, gamma-glutamyl-transferase (GGT), aspartate aminotransferase
 (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT).
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0421}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2000}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0421}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1474}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.2000}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1158}}@{}}
 
@@
 
-\includegraphics[width=2.86458in,height=\textheight]{images/weather1.png}
+\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather1.png}
 
@@
 
-\includegraphics[width=2.86458in,height=\textheight]{images/weather2.png}
+\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather2.png}
 
@@
 for these two years the result from Mayrhofer and Filzmoser (2023) also
-indicates unusually low values for the minimum teperature (atmn), while
+indicates unusually low values for the minimum temperature (atmn), while
 there is disagreement between outlyingness in other variables. Note that
@@
 
-\includegraphics[width=5.72917in,height=\textheight]{images/weather3.png}
+\includegraphics[width=5.72917in,height=\textheight,keepaspectratio]{images/weather3.png}
 
@@
 cells values lower than expected. For years that have been identified as
-outliers in our approach we show the cluster assignment in the last
+outliers in our approach we show the cluster assignment in the bottom
 row.}
@@
 has also provided an analytical result for drawing \gD{} projections of
-the \pD{} ellipse corrsponding to the normal reference distribution.
+the \pD{} ellipse corresponding to the normal reference distribution.
 These combined provide new ways to visualise multivariate data, for this

### 2025-02-01 review/reply.Rmd
spell checked

@@
 We thank the Associate editor and anonymous reviewers for the feedback on the paper.
-In discussion with the editor, it is agreed that becuase the article was submitted as a *short technical note* that the revision would focus on these changes only:
+In discussion with the editor, it is agreed that because the article was submitted as a *short technical note* that the revision would focus on these changes only:
 
@@
 \textcolor{purple}{We have extended the implementation section to mention the options for optimization of the index, and have included a reference for the ferrn package that can be used to investigate index values and to compare different optimization strategies.
-Including such an investigation would be beyond the scope of a short technical note.
 All implementations are available in CRAN packages and the code to reproduce the examples are available on GitHub for full reproducibility.}
@@
 
-Page 7 subection 3.3 It would be useful to include a simulation example demonstrating how to apply the method in cases where the observations deviate from the norm in different directions.
+Page 7 subsection 3.3 It would be useful to include a simulation example demonstrating how to apply the method in cases where the observations deviate from the norm in different directions.
 This example should illustrate how to group these observations effectively, providing readers with a clearer understanding of the practical application of your method.
@@
 
-You should move Figure 4 to page 10. In the figure caption, please correct the phrase “Red cross indicates observation is outide the 4-D confidence ellipse” to fix the spelling of “outside” and 2-D.
+You should move Figure 4 to page 10. In the figure caption, please correct the phrase “Red cross indicates observation is outside the 4-D confidence ellipse” to fix the spelling of “outside” and 2-D.
 
@@
 We have increased the size of the cluster labels in Figure 6 to make them more readable, and have added a color legend.
-Variable nams in Figure 6 have been matched to those in Figure 5, and are now always added in parentheses when mentioned in the text.}
+Variable names in Figure 6 have been matched to those in Figure 5, and are now always added in parentheses when mentioned in the text.}
 

### 2025-02-01 paper.qmd
cleaned up code

@@
 #| echo: false
+# This code generates Fig 2: ellipse vs sample
 set.seed(1158)
@@
 #| echo: false
+# Code to generate data for Fig 3
 set.seed(929)
@@
                            vc = vc_samp))
+set.seed(913) 
+render_gif(samp, 
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=vc_null), 
+           display_xy(ellipse=vc_null, 
+                      axes = "bottomleft", 
+                      half_range=5, center=FALSE),
+           gif_file = "gifs/anomaly.gif",
+           frames = 500,
+           width = 400,
+           height = 400)
 ```
@@
 
-![Random projection](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}
+![Random projection](images/anomaly1.png){#fig-anomaly1 width=250 align="center"}
 
-![Optimal projection](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}
+![Optimal projection](images/anomaly2.png){#fig-anomaly2 width=250 align="center"}
 
@@
 These measurements are also likely correlated, based on guidance like the *ratio of AST to ALT of 2:1 indicates possible alcohol abuse*.
-Although this was provided by the pharmaceutical company, correlation between these measurements for normal patients is not readily available.
-When a correlation matrix for normal patients is provided this would allow construction of the null ellipse upon which to examine new samples.
+When a correlation matrix for normal patients is provided the null ellipse can be computed to be used to examine new samples.
 
@@
 #| echo: false
+set.seed(1158)
+norm_vc <- diag(1, 4, 4)
+norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
+for (i in 1:3)
+  for (j in (i+1):4)
+    norm_vc[j, i] <- norm_vc[i, j]
+colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
+norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
+colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
+
+set.seed(217)
+women_vc <- norm_vc
+women_vc[1,4] <- 0.5
+women_vc[4,1] <- 0.5
+women <- rmvnorm(16, mean=c(0.3, -0.3, -1.5, -1.7),
+                 sigma = women_vc)
+colnames(women) <- c("GGT", "AST", "ALP", "ALT")
+women <- as.data.frame(women)
+
 animate_xy(women,            
@@
            center=FALSE)
+set.seed(929) 
+render_gif(women,  
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=norm_vc), 
+           display_xy(ellipse=norm_vc, 
+                      axes = "bottomleft", 
+                      half_range=5, center=FALSE),
+           gif_file = "gifs/women.gif",
+           frames = 500,
+           width = 400,
+           height = 400)
 
@@
            obs_labels = c("45", "50", "55", "65", "70"))
+set.seed(951) #945
+render_gif(pt1,  
+           guided_anomaly_tour(anomaly_index(),
+                               ellipse=norm_vc), 
+           display_xy(ellipse=norm_vc, 
+                      axes = "bottomleft", 
+                      half_range=5, center=FALSE,
+                      edges = matrix(c(1:4, 2:5), 
+                                     byrow=FALSE, ncol=2),
+                      obs_labels = c("45", "50", "55", "65", "70")),
+           gif_file = "gifs/w_long.gif",
+           frames = 500,
+           width = 400,
+           height = 400)
 ```
@@
 Here only four tests are used, GGT, AST, ALP, ALT.
-Plot (a) shows a projection of this sample relative to the normal range.
-Three of the patients are outside the confidence ellipse but all of the patients are located away from the mean.
+Plot (a) shows a projection of this sample relative to the normal range, using the anomaly index guided tour.
+Some of the patients are outside the 4D confidence ellipse but all of the patients are off center, located away from the mean.
 What has been typical in the past is to compute normal values based on tests of healthy young males.
-The samples provided for the project were all recorded on women.
-We see that this sample has slightly lower ALT and ALP, which is consistent with what is reported in @lib-med-liver-norms.
+Although this example is simulated, it matches what was learned during the project, that the samples examined were all recorded on women, and they were systematically different from the normal.
+<!--We see that this sample has slightly lower ALT and ALP, which is consistent with what is reported in @lib-med-liver-norms.-->
 
@@
 
-![Sample of female patients](images/liver1.png){#fig-liver1 width=300 align="center"}
+![Sample of female patients](images/liver1.png){#fig-liver1 width=250 align="center"}
 
-![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=300 align="center"}
+![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=250 align="center"}
 
@@
 Red cross indicates observation is outside the 4-D confidence ellipse.
-The female patients tend to have lower ALP and ALT than the normal range.
-As the patient ages, the level of ALP increases and ALT decreases.
+The female patients, the whole group (a), are systematically different from the normal range.
+As the patient ages (b), the level of ALP increases and ALT decreases.
 :::
@@
 The lines connect the records in time order.
-The projection corresponds to the maximum from a projection-pursuit guided tour using the anomaly index:
+The projection corresponds to the maximum from a projection-pursuit guided tour using the anomaly index.
 
+<!--
 $$
@@
 $$
+-->
 
@@
 This is consistent with aging, where ALP increases and ALT decreases.
+Although this is simulated data, it matches what was observed in the clinical data that ALP and ALT test scores systematically change with age.
+
+These two examples illustrate how visualizing the data using the anomaly guided tour can help to understand differences between groups of correlated test scores.
 

### 2025-02-01 paper.tex
cleaned up code

@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly1.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/anomaly1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/anomaly2.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/anomaly2.png}
 
@@
 the \emph{ratio of AST to ALT of 2:1 indicates possible alcohol abuse}.
-Although this was provided by the pharmaceutical company, correlation
-between these measurements for normal patients is not readily available.
-When a correlation matrix for normal patients is provided this would
-allow construction of the null ellipse upon which to examine new
-samples.
+When a correlation matrix for normal patients is provided the null
+ellipse can be computed to be used to examine new samples.
 
@@
 range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows
-a projection of this sample relative to the normal range. Three of the
-patients are outside the confidence ellipse but all of the patients are
-located away from the mean. What has been typical in the past is to
-compute normal values based on tests of healthy young males. The samples
-provided for the project were all recorded on women. We see that this
-sample has slightly lower ALT and ALP, which is consistent with what is
-reported in Lala, Zubair, and Minter (2023).
+a projection of this sample relative to the normal range, using the
+anomaly index guided tour. Some of the patients are outside the 4D
+confidence ellipse but all of the patients are off center, located away
+from the mean. What has been typical in the past is to compute normal
+values based on tests of healthy young males. Although this example is
+simulated, it matches what was learned during the project, that the
+samples examined were all recorded on women, and they were
+systematically different from the normal.
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver1.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/liver1.png}
 
@@
 
-\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/liver2.png}
+\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/liver2.png}
 
@@
 longitudinal test results for a single patient. Red cross indicates
-observation is outside the 4-D confidence ellipse. The female patients
-tend to have lower ALP and ALT than the normal range. As the patient
-ages, the level of ALP increases and ALT decreases.}
+observation is outside the 4-D confidence ellipse. The female patients,
+the whole group (a), are systematically different from the normal range.
+As the patient ages (b), the level of ALP increases and ALT decreases.}
 
@@
 time order. The projection corresponds to the maximum from a
-projection-pursuit guided tour using the anomaly index:
-
-\[
-P = \left[ \begin{array}{rr}
-            0.371 & -0.128 \\
-            -0.388 & 0.732 \\
-            0.140 & 0.599 \\
-            -0.832 & -0.297 
-            \end{array} \right]
-\]
+projection-pursuit guided tour using the anomaly index.
 
@@
 primarily contrasting ALT (fourth row) and ALP (third row). This is
-consistent with aging, where ALP increases and ALT decreases.
+consistent with aging, where ALP increases and ALT decreases. Although
+this is simulated data, it matches what was observed in the clinical
+data that ALP and ALT test scores systematically change with age.
+
+These two examples illustrate how visualizing the data using the anomaly
+guided tour can help to understand differences between groups of
+correlated test scores.
 

### 2025-02-01 redundaant_code/jenny.R

@@ -1,0 +1,47 @@
+# Jenny Wadsworth's river data
+library(tidyverse)
+river <- read_csv("data/riverFlow1.csv") |>
+  rename(V1 = `1`, V2 = `2`, V3 = `3`, V4 = `4`) |>
+  mutate(type = "river")
+
+surface <- read_csv("data/SurfacePoints.csv") |>
+  select(-`...1`) |>
+  mutate(type = "surface")
+
+river_all <- bind_rows(river, surface) |>
+  mutate(type = factor(type))
+
+animate_xy(river_all[,1:4], col=river_all$type, scale=TRUE)
+# Data needs to be scaled to 0-1 to match the surface range
+
+river_m <- apply(river[,1:4], 2, mean)
+river_s <- apply(river[,1:4], 2, sd)
+
+river_std <- river |>
+  mutate_at(vars(V1:V4), function(x) (x-min(x))/(max(x)-min(x)))
+
+river_all <- bind_rows(river_std, surface) |>
+  mutate(type = factor(type, levels = c("surface", "river")))
+
+animate_xy(river_all[,1:4], 
+           col=river_all$type, 
+           cex = c(rep(1, nrow(river_std)), 
+                   rep(0.2, nrow(surface))),
+           scale=TRUE, center=FALSE, 
+           axes="bottomleft")
+
+ptsize <- c(rep(1, nrow(river_std)), 
+            rep(0.2, nrow(surface)))
+render_gif(river_all[,1:4], 
+           grand_tour(),
+           display_xy(col=river_all$type, 
+                      cex = ptsize, 
+                      scale=TRUE, center=FALSE, 
+                      axes="bottomleft",
+                      half_range=1.5),
+           gif_file = "river.gif",
+           frames = 500,
+           width = 500,
+           height = 500
+)
+

### 2025-02-04 paper.qmd
Final changes before resubmitting

@@
 Roughly orthogonal to this is a combination of V1 and V2, so the difference between new and reference as indicated by this projection is due to V4 contrasted with V1 and V2.
-The ellipse makes it clearer see but the sample can be useful, too, particularly if the reference distribution is not a multivariate normal distribution.
+The ellipse makes it clearer but the sample can be useful, too, particularly if the reference distribution is not a multivariate normal distribution.
 (The supplementary material includes animated gifs showing these data in a tour, allowing one to see many different projections of the \fD{} points.)
@@
 One could consider clustering these observations using angular distance to find groups of observations that have similar directions of departure.
-This is illustrated in the second example, see Section @sec-weather.
+This is illustrated in the second example, see @sec-weather.
 

### 2025-02-04 paper.tex
Final changes before resubmitting

@@
 between new and reference as indicated by this projection is due to V4
-contrasted with V1 and V2. The ellipse makes it clearer see but the
-sample can be useful, too, particularly if the reference distribution is
-not a multivariate normal distribution. (The supplementary material
-includes animated gifs showing these data in a tour, allowing one to see
-many different projections of the \fD{} points.)
+contrasted with V1 and V2. The ellipse makes it clearer but the sample
+can be useful, too, particularly if the reference distribution is not a
+multivariate normal distribution. (The supplementary material includes
+animated gifs showing these data in a tour, allowing one to see many
+different projections of the \fD{} points.)
 
@@
 groups of observations that have similar directions of departure. This
-is illustrated in the second example, see Section
-Section~\ref{sec-weather}.
+is illustrated in the second example, see Section~\ref{sec-weather}.
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/anomaly1.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/anomaly1.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/anomaly2.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/anomaly2.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/liver1.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/liver1.png}
 
@@
 
-\includegraphics[width=2.60417in,height=\textheight,keepaspectratio]{images/liver2.png}
+\includegraphics[width=2.60417in,height=\textheight]{images/liver2.png}
 
@@
 
-\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather1.png}
+\includegraphics[width=2.86458in,height=\textheight]{images/weather1.png}
 
@@
 
-\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather2.png}
+\includegraphics[width=2.86458in,height=\textheight]{images/weather2.png}
 

### 2025-02-04 review/reply.Rmd
Final changes before resubmitting

@@
 \textcolor{purple}{We have extended the implementation section to mention the options for optimization of the index, and have included a reference for the ferrn package that can be used to investigate index values and to compare different optimization strategies.
-All implementations are available in CRAN packages and the code to reproduce the examples are available on GitHub for full reproducibility.}
+All implementations are available in CRAN packages and the code to reproduce the examples is available on GitHub for full reproducibility.}
 
@@
 \textcolor{purple}{We have extended the introduction somewhat to include references to outlier detection in the context of linear dimension reduction.
-The additional references in the conclusions show potential generalizations of the work connecting to other directionsin outlier detection.}
+The additional references in the conclusions show potential generalizations of the work connecting to other directions in outlier detection.}
